### 2024-12-12 manuscript/paper.tex
Rename manuscript file

%%
%% Copyright 2022 OXFORD UNIVERSITY PRESS
%%
%% This file is part of the 'oup-authoring-template Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'oup-authoring-template Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
%% with bibliographic references
%%

%%%CONTEMPORARY%%%
%\documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
\documentclass[unnumsec,webpdf,contemporary,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}

%%%MODERN%%%
%\documentclass[unnumsec,webpdf,modern,large]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}

%%%TRADITIONAL%%%
%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}

%\onecolumn % for one column layouts

%\usepackage{showframe}

\graphicspath{{Fig/}}

% line numbers
%\usepackage[mathlines, switch]{lineno}
%\usepackage[right]{lineno}
\usepackage{hyperref}
\usepackage{lmodern}
%\usepackage{graphicx}
%\usepackage{xr}

\begin{document}

\journaltitle{Bioinformatics}
\journaltitle{Preprint}
\DOI{DOI HERE}
\copyrightyear{2024}
% \pubyear{2019}
% \access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

\firstpage{1}

\title[tsbrowse]{Tsbrowse: a genome browser for Ancestral Recombination Graphs}

\author[1$\ast$]{Savita Karthikeyan}
\author[1]{Ben Jeffery}
\author[1]{Duncan Mbuli-Robertson}
\author[1]{Jerome Kelleher \ORCID{0000-0000-0000-0000}}

\authormark{Karthikeyan et al.}

\address[1]{\orgdiv{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery}, \orgname{University of Oxford}, \orgaddress{\street{Old Road Campus}, \postcode{Oxford OX3 7LF}, \country{United Kingdom}}}

\corresp[$\ast$]{Corresponding author. \href{email:savita.karthikeyan@st-hughs.ox.ac.uk}{savita.karthikeyan@st-hughs.ox.ac.uk}}

\received{Date}{0}{Year}
\revised{Date}{0}{Year}
\accepted{Date}{0}{Year}

\abstract{
%\textbf{Motivation:}
Ancestral Recombination Graphs (ARGs) represent the interwoven paths of genetic ancestry for a set of recombining sequences. The ability to capture the evolutionary history of samples makes ARGs valuable in a wide range of applications in population and statistical genetics. ARG-based approaches are increasingly becoming a part of genetic data analysis pipelines due to breakthroughs enabling ARG inference at biobank-scale. However, there is a lack of visualisation tools, which are crucial for validating inferences and generating hypotheses. 
%\textbf{Results:} 
We present \texttt{tsbrowse}, an open-source Python web-app for the interactive visualisation of the fundamental building-blocks of ARGs, i.e.,  nodes, edges and mutations. We demonstrate the application of \texttt{tsbrowse} to various data sources and scenarios, and highlight its key features of browsability along the genome, user interactivity, and scalability to very large sample sizes.\\
\textbf{Availability:}\\ Python package: \texttt{\url{https://pypi.org/project/tsbrowse/}}, \\ Development version: \texttt{\url{https://github.com/tskit.dev/tsbrowse}}, \\
Documentation: \texttt{\url{https://tskit.dev/tsbrowse/docs/}}\\
%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
%\textbf{Supplementary information:} Available at \textit{\textcolor{red}{\href{https://www.google.com}{Bioinformatics} online}.}}
}

\keywords{Ancestral Recombination Graphs, Holoviz, biobank-scale ARG visualisation, genome browser}
\maketitle
\section{Introduction}
Visualisation is an essential component of data analysis pipelines. In the past few decades, genome browsers have become synonymous with sequence visualisation, by providing user-friendly, interactive interfaces to navigate genomic sequences and annotations. From early standalone tools to manage \textit{C. elegans} data \citep{Eeckman1995}, to modern web-apps for vertebrate genomes \citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve with the changing landscape of genomics. Modern genome browsers can load datasets from a URL or local file system \citep{IGV-web,Lee2013,Robinson2023}. The data security and scalability offered by this architecture ensures that genome browsers continue to be relevant in the biobank era.

Biobank-driven research has accelerated the development of several technologies, including the ability to infer Ancestral Recombination Graphs (ARGs) from large-scale genetic data \citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}. ARGs encode the interwoven paths of genetic inheritance resulting from recombination, and can be represented as genetic genealogies along the sequence \citep{Wong2024,Hudson1983,Griffiths1997}. They capture rich detail about ancestral processes influencing the samples, and have garnered great interest in several applications in population and statistical genetics. These include the detection of ultra-rare genetic variation \citep{Zhang2023}, the mapping of quantitative trait loci \citep{Link2023}, and the sparse and efficient modeling of linkage disequilibrium \citep{Nowbandegani2023}. 
It is evident that ARG inference is increasingly becoming a part of genetic data analysis pipelines. Yet, ARG visualisation, which is crucial for downstream analyses remains challenging due to the complexity of inferences from empirical data. Existing visualisation tools (see \hyperref[sec:Discussion]{Discussion}) focus on the semantic representation of ARGs, but such representations have limited utility in informing downstream analyses.

To address this gap, we have developed \texttt{tsbrowse}, an open-source app intended as a genome browser for ARGs. Its  focus is on visualising the information structure in ARGs, thereby improving  inference and analysis at biobank-scale. 
\texttt{Tsbrowse} provides interactive visual summaries of  the fundamental properties of ARGs, i.e. the mutations, nodes, edges and trees they encode. \texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences \citep{Wong2024}. This is the native format of \texttt{tskit} \citep{tskit}, a widely-used library for interchanging ARGs inferred using a variety of methods \citep{Speidel2019,Zhang2023,Haller2019,Baumdicker2021,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}. The interface is run on a web browser. It works on biobank-scale datasets without requiring \textcolor{red}{raw data uploads to a server}. This is ideal for performing quality control and exploratory analyses on ARGs inferred from biobank datasets.
\section{Results}
\subsection{Architecture}
    \texttt{Tsbrowse} is run in a step-wise approach (Figure \ref{fig:Figure_1}), which makes it convenient to handle large datasets. The initial step (\texttt{tsbrowse preprocess}) prepares the data for visualisation. It can be performed separately on a large machine, and the output stored for later use. The second step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates and renders visualisations. This step follows a client-server architecture. On the server side (the machine on which \texttt{tsbrowse serve} is run), the Holoviz suite of libraries \citep{Holoviz} is leveraged to create visualisations based on the pre-processed data. This choice of architecture is prompted by the need to represent properties of large-scale ARGs in a meaningful and performant way. For example, a visual representation of millions of mutations in SARS-Cov-2 ARGs \citep{Zhan2023.06.08.544212} could quickly lead to performance bottlenecks and overplotting. Datashader \citep{Datashader}, a component Holoviz package, overcomes this challenge by rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface} for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the server to render and interact with these visualisations. Heavy data processing is limited to the server side, keeping the client-side interface lightweight. For large datasets with complex visualisations, this achieves a smoother user experience.

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{manuscript/figures/MainFig1.png}
    \caption{\textbf{Overview of tsbrowse architecture. } 
    In tskit, ARGs are defined by a collection of tables which broadly map to ARG properties. Exemplar table names are denoted in dark blue and columns that describe the property in light blue. In the pre-process step, tables are augmented with additional information computed for each property (yellow). The output from this pre-processing step is stored as a \texttt{.tsbrowse} file. Next the serve step renders the visualisation in a web browser by leveraging tools in the Holoviz ecosystem. Optional annotations are provided as an input file, allowing users to overlay contextual information about genes or other sequence features. 
    }
    \label{fig:Figure_1}
\end{figure}

\subsection{Data model}
\label{subsec:Data_Model}
    The foundation of \texttt{tsbrowse} is the \texttt{tskit} data model. As \texttt{tsbrowse} is based on this generalised ARG representation \citep{Wong2024} it can be applied to ARGs from several widely-used inference methods including \texttt{Relate} \citep{Speidel2019}, \texttt{ARG-Needle} \citep{Zhang2023} and \texttt{SINGER} \citep{Deng2024.03.16.585351}. In \texttt{tskit}, ARGs are encoded as a collection of tables, each of which store information about a specific ARG property. For example, the Node table describes sampled or ancestral sequences and records their time of birth, and the individual and population they belong to. The Edge table describes the relationship between pairs of nodes, and records the node IDs and the genomic locations over which the edge spans. The Mutation table records mutation events in the history of the sample, including the change of allelic state, time of event and node in which the event occurred. The Node and Edge tables describe the genealogical relationships between the sampled and ancestral sequences, while the Mutation table describes sequence variation contained in the samples \citep{Ralph2020}. This compact representation enables the efficient storage and querying of biobank-scale ARGs. \texttt{Tsbrowse} augments these tables with additional columns to create a compressed \texttt{.tsbrowse} file, containing all necessary semantic metadata required for automatic visualisation (see \hyperref[subsec:User_Interface]{User Interface}).

\subsection{User Interface}
\label{subsec:User_Interface}
\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{manuscript/figures/MainFig2.png}
    \caption{\textbf{A view of ARG edges in \texttt{tsbrowse}.}
    A screenshot of the Edges tab for a 5 Mb sequence simulated with selection coefficient, \textit{s = 0.25}, and beneficial allele situated at \textit{x = 2.5 Mb}. The plot on the left represents each edge as a horizontal line along the sequence (X-axis). The time of parent node is shown on the Y-axis. The persistence of edges across various dimensions is shown through histograms on the right: top, distribution of genomic spans; middle, distribution of time spans; bottom, edge area, which is the product of genomic and time spans.} 
    \label{fig:Figure_2}
\end{figure}

\texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem \citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh}, \texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data visualisation pipelines even for large datasets. A key challenge with large-scale data visualisation is to maintain a responsive user interface. \texttt{Datashader} aggregates data and renders only the necessary data points at a given zoom level, thereby minimising the computational load on the browser. A \texttt{Bokeh} plotting back-end provides useful tools and widgets. The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity with individual data points, allowing users to gain deeper insights from the data. The \texttt{serve} step accepts an optional input file containing sequence annotations. This allows users to overlay contextual information about genes or other sequence features.

The user interface is presented as a \texttt{Panel} dashboard, with pages to describe mutations, nodes, edges and trees in the input ARG. Figure 2 demonstrates the user interface using the Edges page as an example. The plot on the left is an overview of all 33,929 edges (parent-child relationships between node-pairs; see \hyperref[subsec:Data_Model]{Data Model}) in a simulated ARG \hyperref[subsec:sweep_simulation]{described in Supplementary Methods}; each edge is depicted as a horizontal line connecting the genomic coordinates of the parent and child nodes on the x-axis. The y-axis denotes the age of either node. A useful feature of the app is the ability to interact with the visualisations. The tools on the right of the plot allow the user to examine edges within a sequence/time slice, and query individual edges. The histograms to the right provide summaries of edge spans over sequence and time.
%The ability to explore ARG properties across the sequence and interact with the data makes the analysis of local genealogical patterns more intuitive.

 Tsbrowse can be applied to versatile scenarios. Figure \ref{fig:Figure_2} highlights a use-case in visually characterising ARGs. The input ARG is a selective sweep dataset simulated using \texttt{msprime} \citep{Baumdicker2021}. The distinct pattern of edges observed in the middle of the sequence between \textit{y=0} to \textit{y=2.5} is attributable to a burst in recent coalescent events, showing that tsbrowse captures this characteristic of a selective sweep. Supplementary Figure \ref{fig:Supplementary_Figure_1} is another example of characterising ARGs with tsbrowse. The view of edges from four different inferences of the same dataset (inferred with \texttt{tsinfer} + \texttt{tsdate} (Panel A), \texttt{Relate} (Panel B), \texttt{ARG-Needle} (Panel C) and \texttt{SINGER} (Panel D)), reflects qualitative differences between inference methods.

We demonstrate a further application of tsbrowse in validating inferences (Supplementary Figures \ref{fig:Supplementary_Figure_2} and \ref{fig:Supplementary_Figure_3}). Supplementary Figure \ref{fig:Supplementary_Figure_2} shows how mutation patterns in a 2D plot of genomic position and mutation age are helpful in spotting poorly inferred regions. Excluding regions with low site density from the variant data improves the quality of inference (Panel B). Supplementary Figure \ref{fig:Supplementary_Figure_3} shows the genomic spans of ancestral nodes in an ARG inferred with tsinfer. Population genetics theory predicts that IBD (Identical By Descent) segments break down over time, resulting in shorter ancestral segments further back in the history of the samples. At time, \textit{y \textgreater{0.8}}, the plot reveals a pattern of node spans contrary to this expectation. It confirms a known algorithmic artefact due to which disproportionately long old ancestors are generated.

Supplementary Fig. \ref{fig:Supplementary_Figure_4}, a screenshot of the Mutations page for SARS-CoV-2 ARGs \citep{Zhan2023.06.08.544212} demonstrates tsbrowse's ability to handle diverse data sources. This plot represents a total of \textcolor{red}{X million} mutations from an inference of \textcolor{red}{1.27 million} sequences. After executing the \texttt{preprocess} step, the page loads in \textcolor{red}{X} seconds on a system with \textcolor{red}{Y GB RAM}, showing that it is possible to visualise large-scale datasets in a performant way. Panel B demonstrates a useful interactive feature; additional information about a mutation event (for example, the mutational frequency in predefined populations, or the number of sample nodes carrying the mutation) is displayed on mouse-over of individual data points.

\section{Discussion}
\label{sec:Discussion}
Visualisation techniques are essential for data-driven insights. Specialised tools such as \textit{Treenome browser} \citep{Kramer2023} tackle the challenge of semantic visualisation of large phylogenies. Yet capturing the data model underlying ARGs is crucial for the systematic exploration of local genealogies. \texttt{Tsbrowse} addresses this open problem by condensing ARG properties into a set of browsable, interactive visual summaries, thus aiding in ARG-based data exploration and hypothesis generation.

In current practice, simulated ARGs are used as ground-truth to validate inferences from empirical data \citep{Kelleher2019,Zhang2023,Speidel2019,Gunnarsson2024.08.31.610248}. Sophisticated techniques \citep{Haller2019,Baumdicker2021,Adrion2020,Lauterbur2023} to simulate ARGs under complex, realistic demographic models \cite{Anderson-Trocmé2023} now exist. However, models of simulation do not capture \textit{all} the nuances of population datasets. This results in a clear gap between ARG inference and application. \texttt{Tsbrowse} captures ARG properties in a model-free way, providing the means to qualitatively assess inferences. 
With notable features of browsability along the genome, scalability and applicability to diverse data sources, inference and simulation methods, \texttt{tsbrowse} has the potential to be integrated into biobank-scale data pipelines to improve ARG inference and analysis.

\section*{Acknowledgments}
\section*{Author contributions}
\section*{Supplementary Data}
Supplementary data are available at \textcolor{red}{Bioinformatics online}.
\section*{Conflict of interest}
No competing interest is declared.
\section*{Funding}
This work was supported by funding from the Biotechnology and Biological Sciences Research Council (UKRI-BBSRC) [grant number BB/T008784/1], \textcolor{red}{add Novo funding and funding details for Ben, Duncan, Jerome}.
\bibliographystyle{natbib}
\bibliography{manuscript/reference}

\clearpage
\onecolumn
\setcounter{figure}{0}  % Reset the figure counter
\renewcommand{\thefigure}{S\arabic{figure}}  % Change figure numbering to S1, S2, etc.
\section{Supplementary Material}
\subsection{Simulation of truth dataset (Fig. \ref{fig:Figure_2})}
\label{subsec:sweep_simulation}
    Ancestral histories of 300 samples were simulated with \texttt{SweepGenicSelection} function in \texttt{msprime (version 1.3.3)}. A combination of models was used: in the recent past, a selective sweep was simulated with a beneficial allele situated in the middle of a 5 Mb sequence. The frequency of the allele in the population was set at 0.0001 at the beginning of the sweep. The allele fixed in the population at a frequency of 0.9999. The strength of selection was set using the selection coefficient, \textit{s = 0.25}. A time increment, \textit{dt = 1e-6} was used to step through the sweep. Mutations were added to the ARG at a rate of 1e-6. For simulating history before occurrence of the sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was used until coalescence was achieved.

\subsection{Inference of selective sweep dataset (Suppl. Fig. \ref{fig:Supplementary_Figure_1})}
    The following software were used to infer ARGs from the truth dataset described above: \texttt{tsinfer version 0.3.3} \citep{Kelleher2019}, \texttt{tsdate version 0.2.1}, \texttt{Relate version 1.2.2} \cite{Speidel2019}, \texttt{ARG-needle version 1.0.3} \citep{Zhang2023}, \texttt{SINGER version 0.1.8-beta} \citep{Deng2024.03.16.585351}. For all methods, the following parameters of inference were used: \textit{recombination rate = 1e-8}, \textit{mutation rate = 1e-8}, \textit{effective population size = 10,000}. Default values were used for all other inference parameters.

\subsection{Inference of 1000 Genomes dataset (Suppl. Fig. \ref{fig:Supplementary_Figure_2}, Suppl. Fig. \ref{fig:Supplementary_Figure_3})}
    The 1000 Genomes dataset was downloaded from \textcolor{red}{[insert link]}. Ancestral fasta sequence for chromosome 20 (GRCh38) was downloaded from the Ensembl database. Inference was performed with a Snakemake pipeline \textcolor{red}{(cite Ben's repo?)} using \texttt{tsinfer version 0.3.3} for the long arm of chromosome 20 after filtering out duplicate variant positions, variants with missing or low quality ancestral allele, singletons, n-1-tons and n-2-tons. Only bi-allelic SNPs were included for inference. For the inference in Suppl. Fig. 2A, all genomic regions were included. For the inference in Suppl. Figs 2B and 3, genomic regions with a site density lesser than 5 per kb were discarded. For Suppl. Fig. 2, \texttt{tsdate version 0.2.1} was used to estimate the age of ancestral nodes with \textit{mutation rate = 1.29e-8}, setting all other parameters to default values.

\subsection{Inference of SARS-Cov-2 ARGs}

The ARG shown in \ref{fig:Supplementary_Figure_4} was inferred with \texttt{sc2ts}~\citep{Zhan2023.06.08.544212}
using the Viridian dataset~\citep{hunt2024addressing}. It consists 
of 516,443 samples, 612,209 nodes, 612,655 edges and 988,303 mutations.

\subsection{}
Code to recreate datasets and infer ARGs are available in an accompanying repository: \texttt{\url{https://github.com/savitakartik/tsbrowse-paper}}.

\clearpage
\section{Supplementary Figures}
\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{manuscript/figures/SuppFig1.png}
    \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate (A), ARG-Needle (B), Relate (C) and SINGER (4) inferences of the truth dataset simulated under a selective sweep model (in Figure 2 of the main text). For SINGER, an examplar ARG from the set of output posterior ARG samples is shown. The X coordinate represents genomic position, each horizontal segment on the plot shows the genomic coordinates that the edge spans, and Y coordinate shows time of either the parent or child node in the edge.}
    \label{fig:Supplementary_Figure_1}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{manuscript/figures/SuppFig2.png}
    \caption{\textbf{A view of mutations in the 1000G WGS dataset.}
    A screenshot of \texttt{tsbrowse}'s Mutations view for a X Mb region on chromosome 17 inferred from from 3,202 participants from the 1000 Genomes Whole Genome Sequencing dataset (1000G) \citep{1000G2015, 1000GWGS}. The genomic position of the mutations is shown on the X axis, and the age of mutation is represented on the Y axis. Panel B represents an inference after filtering out low site-density regions in the samples.}
    \label{fig:Supplementary_Figure_2}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{manuscript/figures/SuppFig3.png}
    \caption{\textbf{A view of the Nodes page for a 1000 Genomes inference.}
    A screenshot of \texttt{tsbrowse}'s Nodes view for an inference of the long arm of chromosome 20 from the 1000 Genomes whole-genome sequencing dataset. At the top is a plot of node spans over time. The length of sequence that the nodes span is shown on the X axis, and the time of nodes is shown on the Y axis. The histogram at the bottom shows the distribution of node spans. }
    \label{fig:Supplementary_Figure_3}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{manuscript/figures/SuppFig4.png}
    \caption{\textbf{\texttt{tsbrowse} applied to SARS-CoV2 ARGs.}
    A screenshot of \texttt{tsbrowse}'s depiction of 988,303 mutations in an 
    ARG inferred by \texttt{sc2ts}; see  text for details. Also shown are the 
    gene annotations along the X-axis.}
    \label{fig:Supplementary_Figure_4}
\end{figure}
\end{document}

### 2024-12-12 manuscript/paper.tex
Wrap lines

@@
 %\textbf{Motivation:}
-Ancestral Recombination Graphs (ARGs) represent the interwoven paths of genetic ancestry for a set of recombining sequences.
-The ability to capture the evolutionary history of samples makes ARGs valuable in a wide range of applications in population and statistical genetics.
-ARG-based approaches are increasingly becoming a part of genetic data analysis pipelines due to breakthroughs enabling ARG inference at biobank-scale.
-However, there is a lack of visualisation tools, which are crucial for validating inferences and generating hypotheses.
+Ancestral Recombination Graphs (ARGs) represent the interwoven paths of genetic
+ancestry for a set of recombining sequences. The ability to capture the
+evolutionary history of samples makes ARGs valuable in a wide range of
+applications in population and statistical genetics. ARG-based approaches are
+increasingly becoming a part of genetic data analysis pipelines due to
+breakthroughs enabling ARG inference at biobank-scale. However, there is a lack
+of visualisation tools, which are crucial for validating inferences and
+generating hypotheses. 
 %\textbf{Results:} 
-We present \texttt{tsbrowse}, an open-source Python web-app for the interactive visualisation of the fundamental building-blocks of ARGs, i.e.,  nodes, edges and mutations.
-We demonstrate the application of \texttt{tsbrowse} to various data sources and scenarios, and highlight its key features of browsability along the genome, user interactivity, and scalability to very large sample sizes.\\
-\textbf{Availability:}\\ Python package: \texttt{\url{https://pypi.org/project/tsbrowse/}}, \\ Development version: \texttt{\url{https://github.com/tskit.dev/tsbrowse}}, \\
-Documentation: \texttt{\url{https://tskit.dev/tsbrowse/docs/}}\\
+We present \texttt{tsbrowse}, an open-source Python web-app for the interactive
+visualisation of the fundamental building-blocks of ARGs, i.e.,  nodes, edges
+and mutations. We demonstrate the application of \texttt{tsbrowse} to various
+data sources and scenarios, and highlight its key features of browsability
+along the genome, user interactivity, and scalability to very large sample
+sizes.\\ \textbf{Availability:}\\ Python package:
+\texttt{\url{https://pypi.org/project/tsbrowse/}}, \\ Development version:
+\texttt{\url{https://github.com/tskit.dev/tsbrowse}}, \\ Documentation:
+\texttt{\url{https://tskit.dev/tsbrowse/docs/}}\\
 %\textbf{Contact:} \href{name@email.com}{name@email.com}\\
-%\textbf{Supplementary information:} Available at \textit{\textcolor{red}{\href{https://www.google.com}{Bioinformatics} online}.}}
+%\textbf{Supplementary information:} Available at
+%\textit{\textcolor{red}{\href{https://www.google.com}{Bioinformatics}
+%online}.}}
 }
 
-\keywords{Ancestral Recombination Graphs, Holoviz, biobank-scale ARG visualisation, genome browser}
-\maketitle
-\section{Introduction}
-Visualisation is an essential component of data analysis pipelines.
-In the past few decades, genome browsers have become synonymous with sequence visualisation, by providing user-friendly, interactive interfaces to navigate genomic sequences and annotations.
-From early standalone tools to manage \textit{C.
-elegans} data \citep{Eeckman1995}, to modern web-apps for vertebrate genomes \citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve with the changing landscape of genomics.
-Modern genome browsers can load datasets from a URL or local file system \citep{IGV-web,Lee2013,Robinson2023}.
-The data security and scalability offered by this architecture ensures that genome browsers continue to be relevant in the biobank era.
+\keywords{Ancestral Recombination Graphs, Holoviz, biobank-scale ARG
+visualisation, genome browser} \maketitle \section{Introduction} Visualisation
+is an essential component of data analysis pipelines. In the past few decades,
+genome browsers have become synonymous with sequence visualisation, by
+providing user-friendly, interactive interfaces to navigate genomic sequences
+and annotations. From early standalone tools to manage \textit{C. elegans} data
+\citep{Eeckman1995}, to modern web-apps for vertebrate genomes
+\citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
+with the changing landscape of genomics. Modern genome browsers can load
+datasets from a URL or local file system \citep{IGV-web,Lee2013,Robinson2023}.
+The data security and scalability offered by this architecture ensures that
+genome browsers continue to be relevant in the biobank era.
 
-Biobank-driven research has accelerated the development of several technologies, including the ability to infer Ancestral Recombination Graphs (ARGs) from large-scale genetic data \citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
-ARGs encode the interwoven paths of genetic inheritance resulting from recombination, and can be represented as genetic genealogies along the sequence \citep{Wong2024,Hudson1983,Griffiths1997}.
-They capture rich detail about ancestral processes influencing the samples, and have garnered great interest in several applications in population and statistical genetics.
-These include the detection of ultra-rare genetic variation \citep{Zhang2023}, the mapping of quantitative trait loci \citep{Link2023}, and the sparse and efficient modeling of linkage disequilibrium \citep{Nowbandegani2023}.
-It is evident that ARG inference is increasingly becoming a part of genetic data analysis pipelines.
-Yet, ARG visualisation, which is crucial for downstream analyses remains challenging due to the complexity of inferences from empirical data.
-Existing visualisation tools (see \hyperref[sec:Discussion]{Discussion}) focus on the semantic representation of ARGs, but such representations have limited utility in informing downstream analyses.
+Biobank-driven research has accelerated the development of several
+technologies, including the ability to infer Ancestral Recombination Graphs
+(ARGs) from large-scale genetic data
+\citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
+ARGs encode the interwoven paths of genetic inheritance resulting from
+recombination, and can be represented as genetic genealogies along the sequence
+\citep{Wong2024,Hudson1983,Griffiths1997}. They capture rich detail about
+ancestral processes influencing the samples, and have garnered great interest
+in several applications in population and statistical genetics. These include
+the detection of ultra-rare genetic variation \citep{Zhang2023}, the mapping of
+quantitative trait loci \citep{Link2023}, and the sparse and efficient modeling
+of linkage disequilibrium \citep{Nowbandegani2023}. It is evident that ARG
+inference is increasingly becoming a part of genetic data analysis pipelines.
+Yet, ARG visualisation, which is crucial for downstream analyses remains
+challenging due to the complexity of inferences from empirical data. Existing
+visualisation tools (see \hyperref[sec:Discussion]{Discussion}) focus on the
+semantic representation of ARGs, but such representations have limited utility
+in informing downstream analyses.
 
-To address this gap, we have developed \texttt{tsbrowse}, an open-source app intended as a genome browser for ARGs.
-Its  focus is on visualising the information structure in ARGs, thereby improving  inference and analysis at biobank-scale.
-\texttt{Tsbrowse} provides interactive visual summaries of  the fundamental properties of ARGs, i.e.
-the mutations, nodes, edges and trees they encode.
-\texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences \citep{Wong2024}.
-This is the native format of \texttt{tskit} \citep{tskit}, a widely-used library for interchanging ARGs inferred using a variety of methods \citep{Speidel2019,Zhang2023,Haller2019,Baumdicker2021,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
-The interface is run on a web browser.
-It works on biobank-scale datasets without requiring \textcolor{red}{raw data uploads to a server}.
-This is ideal for performing quality control and exploratory analyses on ARGs inferred from biobank datasets.
-\section{Results}
-\subsection{Architecture}
-    \texttt{Tsbrowse} is run in a step-wise approach (Figure \ref{fig:Figure_1}), which makes it convenient to handle large datasets.
-The initial step (\texttt{tsbrowse preprocess}) prepares the data for visualisation.
-It can be performed separately on a large machine, and the output stored for later use.
-The second step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates and renders visualisations.
-This step follows a client-server architecture.
-On the server side (the machine on which \texttt{tsbrowse serve} is run), the Holoviz suite of libraries \citep{Holoviz} is leveraged to create visualisations based on the pre-processed data.
-This choice of architecture is prompted by the need to represent properties of large-scale ARGs in a meaningful and performant way.
-For example, a visual representation of millions of mutations in SARS-Cov-2 ARGs \citep{Zhan2023.06.08.544212} could quickly lead to performance bottlenecks and overplotting.
-Datashader \citep{Datashader}, a component Holoviz package, overcomes this challenge by rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface} for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a tsbrowse view of SARS-Cov-2 mutations).
-A web browser client connects to the server to render and interact with these visualisations.
-Heavy data processing is limited to the server side, keeping the client-side interface lightweight.
-For large datasets with complex visualisations, this achieves a smoother user experience.
+To address this gap, we have developed \texttt{tsbrowse}, an open-source app
+intended as a genome browser for ARGs. Its  focus is on visualising the
+information structure in ARGs, thereby improving  inference and analysis at
+biobank-scale. \texttt{Tsbrowse} provides interactive visual summaries of  the
+fundamental properties of ARGs, i.e. the mutations, nodes, edges and trees they
+encode. \texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences
+\citep{Wong2024}. This is the native format of \texttt{tskit} \citep{tskit}, a
+widely-used library for interchanging ARGs inferred using a variety of methods
+\citep{Speidel2019,Zhang2023,Haller2019,Baumdicker2021,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
+The interface is run on a web browser. It works on biobank-scale datasets
+without requiring \textcolor{red}{raw data uploads to a server}. This is ideal
+for performing quality control and exploratory analyses on ARGs inferred from
+biobank datasets. \section{Results} \subsection{Architecture} \texttt{Tsbrowse}
+is run in a step-wise approach (Figure \ref{fig:Figure_1}), which makes it
+convenient to handle large datasets. The initial step (\texttt{tsbrowse
+preprocess}) prepares the data for visualisation. It can be performed
+separately on a large machine, and the output stored for later use. The second
+step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
+and renders visualisations. This step follows a client-server architecture. On
+the server side (the machine on which \texttt{tsbrowse serve} is run), the
+Holoviz suite of libraries \citep{Holoviz} is leveraged to create
+visualisations based on the pre-processed data. This choice of architecture is
+prompted by the need to represent properties of large-scale ARGs in a
+meaningful and performant way. For example, a visual representation of millions
+of mutations in SARS-Cov-2 ARGs \citep{Zhan2023.06.08.544212} could quickly
+lead to performance bottlenecks and overplotting. Datashader
+\citep{Datashader}, a component Holoviz package, overcomes this challenge by
+rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
+for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
+tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
+server to render and interact with these visualisations. Heavy data processing
+is limited to the server side, keeping the client-side interface lightweight.
+For large datasets with complex visualisations, this achieves a smoother user
+experience.
+
+\begin{figure} \centering
+\includegraphics[width=0.95\linewidth]{figures/MainFig1.png}
+\caption{\textbf{Overview of tsbrowse architecture. } In tskit, ARGs are
+defined by a collection of tables which broadly map to ARG properties. Exemplar
+table names are denoted in dark blue and columns that describe the property in
+light blue. In the pre-process step, tables are augmented with additional
+information computed for each property (yellow). The output from this
+pre-processing step is stored as a \texttt{.tsbrowse} file. Next the serve step
+renders the visualisation in a web browser by leveraging tools in the Holoviz
+ecosystem. Optional annotations are provided as an input file, allowing users
+to overlay contextual information about genes or other sequence features. }
+\label{fig:Figure_1} \end{figure}
+
+\subsection{Data model} \label{subsec:Data_Model} The foundation of
+\texttt{tsbrowse} is the \texttt{tskit} data model. As \texttt{tsbrowse} is
+based on this generalised ARG representation \citep{Wong2024} it can be applied
+to ARGs from several widely-used inference methods including \texttt{Relate}
+\citep{Speidel2019}, \texttt{ARG-Needle} \citep{Zhang2023} and \texttt{SINGER}
+\citep{Deng2024.03.16.585351}. In \texttt{tskit}, ARGs are encoded as a
+collection of tables, each of which store information about a specific ARG
+property. For example, the Node table describes sampled or ancestral sequences
+and records their time of birth, and the individual and population they belong
+to. The Edge table describes the relationship between pairs of nodes, and
+records the node IDs and the genomic locations over which the edge spans. The
+Mutation table records mutation events in the history of the sample, including
+the change of allelic state, time of event and node in which the event
+occurred. The Node and Edge tables describe the genealogical relationships
+between the sampled and ancestral sequences, while the Mutation table describes
+sequence variation contained in the samples \citep{Ralph2020}. This compact
+representation enables the efficient storage and querying of biobank-scale
+ARGs. \texttt{Tsbrowse} augments these tables with additional columns to create
+a compressed \texttt{.tsbrowse} file, containing all necessary semantic
+metadata required for automatic visualisation (see
+\hyperref[subsec:User_Interface]{User Interface}).
+
+\subsection{User Interface}
+\label{subsec:User_Interface}
 
@@
     \centering
-    \includegraphics[width=0.95\linewidth]{manuscript/figures/MainFig1.png}
-    \caption{\textbf{Overview of tsbrowse architecture. } 
-    In tskit, ARGs are defined by a collection of tables which broadly map to ARG properties.
-Exemplar table names are denoted in dark blue and columns that describe the property in light blue.
-In the pre-process step, tables are augmented with additional information computed for each property (yellow).
-The output from this pre-processing step is stored as a \texttt{.tsbrowse} file.
-Next the serve step renders the visualisation in a web browser by leveraging tools in the Holoviz ecosystem.
-Optional annotations are provided as an input file, allowing users to overlay contextual information about genes or other sequence features.
-    }
-    \label{fig:Figure_1}
-\end{figure}
-
-\subsection{Data model}
-\label{subsec:Data_Model}
-    The foundation of \texttt{tsbrowse} is the \texttt{tskit} data model.
-As \texttt{tsbrowse} is based on this generalised ARG representation \citep{Wong2024} it can be applied to ARGs from several widely-used inference methods including \texttt{Relate} \citep{Speidel2019}, \texttt{ARG-Needle} \citep{Zhang2023} and \texttt{SINGER} \citep{Deng2024.03.16.585351}.
-In \texttt{tskit}, ARGs are encoded as a collection of tables, each of which store information about a specific ARG property.
-For example, the Node table describes sampled or ancestral sequences and records their time of birth, and the individual and population they belong to.
-The Edge table describes the relationship between pairs of nodes, and records the node IDs and the genomic locations over which the edge spans.
-The Mutation table records mutation events in the history of the sample, including the change of allelic state, time of event and node in which the event occurred.
-The Node and Edge tables describe the genealogical relationships between the sampled and ancestral sequences, while the Mutation table describes sequence variation contained in the samples \citep{Ralph2020}.
-This compact representation enables the efficient storage and querying of biobank-scale ARGs.
-\texttt{Tsbrowse} augments these tables with additional columns to create a compressed \texttt{.tsbrowse} file, containing all necessary semantic metadata required for automatic visualisation (see \hyperref[subsec:User_Interface]{User Interface}).
-
-\subsection{User Interface}
-\label{subsec:User_Interface}
-\begin{figure}
-    \centering
-    \includegraphics[width=0.95\linewidth]{manuscript/figures/MainFig2.png}
+    \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
     \caption{\textbf{A view of ARG edges in \texttt{tsbrowse}.}
@@
 The persistence of edges across various dimensions is shown through histograms on the right: top, distribution of genomic spans; middle, distribution of time spans; bottom, edge area, which is the product of genomic and time spans.}
+
     \label{fig:Figure_2}
@@
 
-\texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem \citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh}, \texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data visualisation pipelines even for large datasets.
-A key challenge with large-scale data visualisation is to maintain a responsive user interface.
-\texttt{Datashader} aggregates data and renders only the necessary data points at a given zoom level, thereby minimising the computational load on the browser.
-A \texttt{Bokeh} plotting back-end provides useful tools and widgets.
-The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity with individual data points, allowing users to gain deeper insights from the data.
-The \texttt{serve} step accepts an optional input file containing sequence annotations.
-This allows users to overlay contextual information about genes or other sequence features.
+\texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem
+\citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh},
+\texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data
+visualisation pipelines even for large datasets. A key challenge with
+large-scale data visualisation is to maintain a responsive user interface.
+\texttt{Datashader} aggregates data and renders only the necessary data points
+at a given zoom level, thereby minimising the computational load on the
+browser. A \texttt{Bokeh} plotting back-end provides useful tools and widgets.
+The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity
+with individual data points, allowing users to gain deeper insights from the
+data. The \texttt{serve} step accepts an optional input file containing
+sequence annotations. This allows users to overlay contextual information about
+genes or other sequence features.
 
-The user interface is presented as a \texttt{Panel} dashboard, with pages to describe mutations, nodes, edges and trees in the input ARG.
-Figure 2 demonstrates the user interface using the Edges page as an example.
-The plot on the left is an overview of all 33,929 edges (parent-child relationships between node-pairs; see \hyperref[subsec:Data_Model]{Data Model}) in a simulated ARG \hyperref[subsec:sweep_simulation]{described in Supplementary Methods}; each edge is depicted as a horizontal line connecting the genomic coordinates of the parent and child nodes on the x-axis.
-The y-axis denotes the age of either node.
-A useful feature of the app is the ability to interact with the visualisations.
-The tools on the right of the plot allow the user to examine edges within a sequence/time slice, and query individual edges.
-The histograms to the right provide summaries of edge spans over sequence and time.
+The user interface is presented as a \texttt{Panel} dashboard, with pages to
+describe mutations, nodes, edges and trees in the input ARG. Figure 2
+demonstrates the user interface using the Edges page as an example. The plot on
+the left is an overview of all 33,929 edges (parent-child relationships between
+node-pairs; see \hyperref[subsec:Data_Model]{Data Model}) in a simulated ARG
+\hyperref[subsec:sweep_simulation]{described in Supplementary Methods}; each
+edge is depicted as a horizontal line connecting the genomic coordinates of the
+parent and child nodes on the x-axis. The y-axis denotes the age of either
+node. A useful feature of the app is the ability to interact with the
+visualisations. The tools on the right of the plot allow the user to examine
+edges within a sequence/time slice, and query individual edges. The histograms
+to the right provide summaries of edge spans over sequence and time.
+
 %The ability to explore ARG properties across the sequence and interact with the data makes the analysis of local genealogical patterns more intuitive.
 
- Tsbrowse can be applied to versatile scenarios.
-Figure \ref{fig:Figure_2} highlights a use-case in visually characterising ARGs.
-The input ARG is a selective sweep dataset simulated using \texttt{msprime} \citep{Baumdicker2021}.
-The distinct pattern of edges observed in the middle of the sequence between \textit{y=0} to \textit{y=2.5} is attributable to a burst in recent coalescent events, showing that tsbrowse captures this characteristic of a selective sweep.
-Supplementary Figure \ref{fig:Supplementary_Figure_1} is another example of characterising ARGs with tsbrowse.
-The view of edges from four different inferences of the same dataset (inferred with \texttt{tsinfer} + \texttt{tsdate} (Panel A), \texttt{Relate} (Panel B), \texttt{ARG-Needle} (Panel C) and \texttt{SINGER} (Panel D)), reflects qualitative differences between inference methods.
+ Tsbrowse can be applied to versatile scenarios. Figure \ref{fig:Figure_2}
+highlights a use-case in visually characterising ARGs. The input ARG is a
+selective sweep dataset simulated using \texttt{msprime}
+\citep{Baumdicker2021}. The distinct pattern of edges observed in the middle of
+the sequence between \textit{y=0} to \textit{y=2.5} is attributable to a burst
+in recent coalescent events, showing that tsbrowse captures this characteristic
+of a selective sweep. Supplementary Figure \ref{fig:Supplementary_Figure_1} is
+another example of characterising ARGs with tsbrowse. The view of edges from
+four different inferences of the same dataset (inferred with \texttt{tsinfer} +
+\texttt{tsdate} (Panel A), \texttt{Relate} (Panel B), \texttt{ARG-Needle}
+(Panel C) and \texttt{SINGER} (Panel D)), reflects qualitative differences
+between inference methods.
 
-We demonstrate a further application of tsbrowse in validating inferences (Supplementary Figures \ref{fig:Supplementary_Figure_2} and \ref{fig:Supplementary_Figure_3}).
-Supplementary Figure \ref{fig:Supplementary_Figure_2} shows how mutation patterns in a 2D plot of genomic position and mutation age are helpful in spotting poorly inferred regions.
-Excluding regions with low site density from the variant data improves the quality of inference (Panel B).
-Supplementary Figure \ref{fig:Supplementary_Figure_3} shows the genomic spans of ancestral nodes in an ARG inferred with tsinfer.
-Population genetics theory predicts that IBD (Identical By Descent) segments break down over time, resulting in shorter ancestral segments further back in the history of the samples.
-At time, \textit{y \textgreater{0.8}}, the plot reveals a pattern of node spans contrary to this expectation.
-It confirms a known algorithmic artefact due to which disproportionately long old ancestors are generated.
+We demonstrate a further application of tsbrowse in validating inferences
+(Supplementary Figures \ref{fig:Supplementary_Figure_2} and
+\ref{fig:Supplementary_Figure_3}). Supplementary Figure
+\ref{fig:Supplementary_Figure_2} shows how mutation patterns in a 2D plot of
+genomic position and mutation age are helpful in spotting poorly inferred
+regions. Excluding regions with low site density from the variant data improves
+the quality of inference (Panel B). Supplementary Figure
+\ref{fig:Supplementary_Figure_3} shows the genomic spans of ancestral nodes in
+an ARG inferred with tsinfer. Population genetics theory predicts that IBD
+(Identical By Descent) segments break down over time, resulting in shorter
+ancestral segments further back in the history of the samples. At time,
+\textit{y \textgreater{0.8}}, the plot reveals a pattern of node spans contrary
+to this expectation. It confirms a known algorithmic artefact due to which
+disproportionately long old ancestors are generated.
 
-Supplementary Fig.
-\ref{fig:Supplementary_Figure_4}, a screenshot of the Mutations page for SARS-CoV-2 ARGs \citep{Zhan2023.06.08.544212} demonstrates tsbrowse's ability to handle diverse data sources.
-This plot represents a total of \textcolor{red}{X million} mutations from an inference of \textcolor{red}{1.27 million} sequences.
-After executing the \texttt{preprocess} step, the page loads in \textcolor{red}{X} seconds on a system with \textcolor{red}{Y GB RAM}, showing that it is possible to visualise large-scale datasets in a performant way.
-Panel B demonstrates a useful interactive feature; additional information about a mutation event (for example, the mutational frequency in predefined populations, or the number of sample nodes carrying the mutation) is displayed on mouse-over of individual data points.
+Supplementary Fig. \ref{fig:Supplementary_Figure_4}, a screenshot of the
+Mutations page for SARS-CoV-2 ARGs \citep{Zhan2023.06.08.544212} demonstrates
+tsbrowse's ability to handle diverse data sources. This plot represents a total
+of \textcolor{red}{X million} mutations from an inference of
+\textcolor{red}{1.27 million} sequences. After executing the
+\texttt{preprocess} step, the page loads in \textcolor{red}{X} seconds on a
+system with \textcolor{red}{Y GB RAM}, showing that it is possible to visualise
+large-scale datasets in a performant way. Panel B demonstrates a useful
+interactive feature; additional information about a mutation event (for
+example, the mutational frequency in predefined populations, or the number of
+sample nodes carrying the mutation) is displayed on mouse-over of individual
+data points.
 
-\section{Discussion}
-\label{sec:Discussion}
-Visualisation techniques are essential for data-driven insights.
-Specialised tools such as \textit{Treenome browser} \citep{Kramer2023} tackle the challenge of semantic visualisation of large phylogenies.
-Yet capturing the data model underlying ARGs is crucial for the systematic exploration of local genealogies.
-\texttt{Tsbrowse} addresses this open problem by condensing ARG properties into a set of browsable, interactive visual summaries, thus aiding in ARG-based data exploration and hypothesis generation.
+\section{Discussion} \label{sec:Discussion} Visualisation techniques are
+essential for data-driven insights. Specialised tools such as \textit{Treenome
+browser} \citep{Kramer2023} tackle the challenge of semantic visualisation of
+large phylogenies. Yet capturing the data model underlying ARGs is crucial for
+the systematic exploration of local genealogies. \texttt{Tsbrowse} addresses
+this open problem by condensing ARG properties into a set of browsable,
+interactive visual summaries, thus aiding in ARG-based data exploration and
+hypothesis generation.
 
-In current practice, simulated ARGs are used as ground-truth to validate inferences from empirical data \citep{Kelleher2019,Zhang2023,Speidel2019,Gunnarsson2024.08.31.610248}.
-Sophisticated techniques \citep{Haller2019,Baumdicker2021,Adrion2020,Lauterbur2023} to simulate ARGs under complex, realistic demographic models \cite{Anderson-Trocmé2023} now exist.
-However, models of simulation do not capture \textit{all} the nuances of population datasets.
-This results in a clear gap between ARG inference and application.
-\texttt{Tsbrowse} captures ARG properties in a model-free way, providing the means to qualitatively assess inferences.
-With notable features of browsability along the genome, scalability and applicability to diverse data sources, inference and simulation methods, \texttt{tsbrowse} has the potential to be integrated into biobank-scale data pipelines to improve ARG inference and analysis.
+In current practice, simulated ARGs are used as ground-truth to validate
+inferences from empirical data
+\citep{Kelleher2019,Zhang2023,Speidel2019,Gunnarsson2024.08.31.610248}.
+Sophisticated techniques
+\citep{Haller2019,Baumdicker2021,Adrion2020,Lauterbur2023} to simulate ARGs
+under complex, realistic demographic models \cite{Anderson-Trocmé2023} now
+exist. However, models of simulation do not capture \textit{all} the nuances of
+population datasets. This results in a clear gap between ARG inference and
+application. \texttt{Tsbrowse} captures ARG properties in a model-free way,
+providing the means to qualitatively assess inferences. With notable features
+of browsability along the genome, scalability and applicability to diverse data
+sources, inference and simulation methods, \texttt{tsbrowse} has the potential
+to be integrated into biobank-scale data pipelines to improve ARG inference and
+analysis.
 
@@
 \section*{Funding}
-This work was supported by funding from the Biotechnology and Biological Sciences Research Council (UKRI-BBSRC) [grant number BB/T008784/1], \textcolor{red}{add Novo funding and funding details for Ben, Duncan, Jerome}.
+This work was supported by funding from the Biotechnology and Biological
+Sciences Research Council (UKRI-BBSRC) [grant number BB/T008784/1],
+\textcolor{red}{add Novo funding and funding details for Ben, Duncan, Jerome}.
 \bibliographystyle{natbib}
-\bibliography{manuscript/reference}
+\bibliography{reference}
 
-\clearpage
-\onecolumn
-\setcounter{figure}{0}  % Reset the figure counter
-\renewcommand{\thefigure}{S\arabic{figure}}  % Change figure numbering to S1, S2, etc.
-\section{Supplementary Material}
-\subsection{Simulation of truth dataset (Fig. \ref{fig:Figure_2})}
-\label{subsec:sweep_simulation}
-    Ancestral histories of 300 samples were simulated with \texttt{SweepGenicSelection} function in \texttt{msprime (version 1.3.3)}.
-A combination of models was used: in the recent past, a selective sweep was simulated with a beneficial allele situated in the middle of a 5 Mb sequence.
-The frequency of the allele in the population was set at 0.0001 at the beginning of the sweep.
-The allele fixed in the population at a frequency of 0.9999.
-The strength of selection was set using the selection coefficient, \textit{s = 0.25}.
-A time increment, \textit{dt = 1e-6} was used to step through the sweep.
-Mutations were added to the ARG at a rate of 1e-6.
-For simulating history before occurrence of the sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was used until coalescence was achieved.
+\clearpage \onecolumn \setcounter{figure}{0}  % Reset the figure counter
+\renewcommand{\thefigure}{S\arabic{figure}}  % Change figure numbering to S1,
+S2, etc. \section{Supplementary Material} \subsection{Simulation of truth
+dataset (Fig. \ref{fig:Figure_2})} \label{subsec:sweep_simulation} Ancestral
+histories of 300 samples were simulated with \texttt{SweepGenicSelection}
+function in \texttt{msprime (version 1.3.3)}. A combination of models was used:
+in the recent past, a selective sweep was simulated with a beneficial allele
+situated in the middle of a 5 Mb sequence. The frequency of the allele in the
+population was set at 0.0001 at the beginning of the sweep. The allele fixed in
+the population at a frequency of 0.9999. The strength of selection was set
+using the selection coefficient, \textit{s = 0.25}. A time increment,
+\textit{dt = 1e-6} was used to step through the sweep. Mutations were added to
+the ARG at a rate of 1e-6. For simulating history before occurrence of the
+sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was
+used until coalescence was achieved.
 
-\subsection{Inference of selective sweep dataset (Suppl. Fig. \ref{fig:Supplementary_Figure_1})}
-    The following software were used to infer ARGs from the truth dataset described above: \texttt{tsinfer version 0.3.3} \citep{Kelleher2019}, \texttt{tsdate version 0.2.1}, \texttt{Relate version 1.2.2} \cite{Speidel2019}, \texttt{ARG-needle version 1.0.3} \citep{Zhang2023}, \texttt{SINGER version 0.1.8-beta} \citep{Deng2024.03.16.585351}.
-For all methods, the following parameters of inference were used: \textit{recombination rate = 1e-8}, \textit{mutation rate = 1e-8}, \textit{effective population size = 10,000}.
-Default values were used for all other inference parameters.
+\subsection{Inference of selective sweep dataset (Suppl. Fig.
+\ref{fig:Supplementary_Figure_1})} The following software were used to infer
+ARGs from the truth dataset described above: \texttt{tsinfer version 0.3.3}
+\citep{Kelleher2019}, \texttt{tsdate version 0.2.1}, \texttt{Relate version
+1.2.2} \cite{Speidel2019}, \texttt{ARG-needle version 1.0.3} \citep{Zhang2023},
+\texttt{SINGER version 0.1.8-beta} \citep{Deng2024.03.16.585351}. For all
+methods, the following parameters of inference were used: \textit{recombination
+rate = 1e-8}, \textit{mutation rate = 1e-8}, \textit{effective population size
+= 10,000}. Default values were used for all other inference parameters.
 
-\subsection{Inference of 1000 Genomes dataset (Suppl. Fig. \ref{fig:Supplementary_Figure_2}, Suppl. Fig. \ref{fig:Supplementary_Figure_3})}
-    The 1000 Genomes dataset was downloaded from \textcolor{red}{[insert link]}.
-Ancestral fasta sequence for chromosome 20 (GRCh38) was downloaded from the Ensembl database.
-Inference was performed with a Snakemake pipeline \textcolor{red}{(cite Ben's repo?)} using \texttt{tsinfer version 0.3.3} for the long arm of chromosome 20 after filtering out duplicate variant positions, variants with missing or low quality ancestral allele, singletons, n-1-tons and n-2-tons.
-Only bi-allelic SNPs were included for inference.
-For the inference in Suppl.
-Fig.
-2A, all genomic regions were included.
-For the inference in Suppl.
-Figs 2B and 3, genomic regions with a site density lesser than 5 per kb were discarded.
-For Suppl.
-Fig.
-2, \texttt{tsdate version 0.2.1} was used to estimate the age of ancestral nodes with \textit{mutation rate = 1.29e-8}, setting all other parameters to default values.
+\subsection{Inference of 1000 Genomes dataset (Suppl. Fig.
+\ref{fig:Supplementary_Figure_2}, Suppl. Fig.
+\ref{fig:Supplementary_Figure_3})} The 1000 Genomes dataset was downloaded from
+\textcolor{red}{[insert link]}. Ancestral fasta sequence for chromosome 20
+(GRCh38) was downloaded from the Ensembl database. Inference was performed with
+a Snakemake pipeline \textcolor{red}{(cite Ben's repo?)} using \texttt{tsinfer
+version 0.3.3} for the long arm of chromosome 20 after filtering out duplicate
+variant positions, variants with missing or low quality ancestral allele,
+singletons, n-1-tons and n-2-tons. Only bi-allelic SNPs were included for
+inference. For the inference in Suppl. Fig. 2A, all genomic regions were
+included. For the inference in Suppl. Figs 2B and 3, genomic regions with a
+site density lesser than 5 per kb were discarded. For Suppl. Fig. 2,
+\texttt{tsdate version 0.2.1} was used to estimate the age of ancestral nodes
+with \textit{mutation rate = 1.29e-8}, setting all other parameters to default
+values.
 
@@
 
-The ARG shown in \ref{fig:Supplementary_Figure_4} was inferred with \texttt{sc2ts}~\citep{Zhan2023.06.08.544212}
-using the Viridian dataset~\citep{hunt2024addressing}. It consists 
-of 516,443 samples, 612,209 nodes, 612,655 edges and 988,303 mutations.
+The ARG shown in \ref{fig:Supplementary_Figure_4} was inferred with
+\texttt{sc2ts}~\citep{Zhan2023.06.08.544212} using the Viridian
+dataset~\citep{hunt2024addressing}. It consists of 516,443 samples, 612,209
+nodes, 612,655 edges and 988,303 mutations.
 
-\subsection{}
-Code to recreate datasets and infer ARGs are available in an accompanying repository: \texttt{\url{https://github.com/savitakartik/tsbrowse-paper}}.
+\subsection{} Code to recreate datasets and infer ARGs are available in an
+accompanying repository:
+\texttt{\url{https://github.com/savitakartik/tsbrowse-paper}}.
 
@@
     \centering
-    \includegraphics[width=0.95\linewidth]{manuscript/figures/SuppFig1.png}
+    \includegraphics[width=0.95\linewidth]{figures/SuppFig1.png}
     \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
-    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate (A), ARG-Needle (B), Relate (C) and SINGER (4) inferences of the truth dataset simulated under a selective sweep model (in Figure 2 of the main text).
-For SINGER, an examplar ARG from the set of output posterior ARG samples is shown.
-The X coordinate represents genomic position, each horizontal segment on the plot shows the genomic coordinates that the edge spans, and Y coordinate shows time of either the parent or child node in the edge.}
+    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate (A),
+ARG-Needle (B), Relate (C) and SINGER (4) inferences of the truth dataset
+simulated under a selective sweep model (in Figure 2 of the main text). For
+SINGER, an examplar ARG from the set of output posterior ARG samples is shown.
+The X coordinate represents genomic position, each horizontal segment on the
+plot shows the genomic coordinates that the edge spans, and Y coordinate shows
+time of either the parent or child node in the edge.}
     \label{fig:Supplementary_Figure_1}
@@
     \centering
-    \includegraphics[width=0.95\linewidth]{manuscript/figures/SuppFig2.png}
-    \caption{\textbf{A view of mutations in the 1000G WGS dataset.}
-    A screenshot of \texttt{tsbrowse}'s Mutations view for a X Mb region on chromosome 17 inferred from from 3,202 participants from the 1000 Genomes Whole Genome Sequencing dataset (1000G) \citep{1000G2015, 1000GWGS}.
-The genomic position of the mutations is shown on the X axis, and the age of mutation is represented on the Y axis.
-Panel B represents an inference after filtering out low site-density regions in the samples.}
+    \includegraphics[width=0.95\linewidth]{figures/SuppFig2.png}
+    \caption{\textbf{A view of mutations in the 1000G WGS dataset.} A
+screenshot of \texttt{tsbrowse}'s Mutations view for a X Mb region on
+chromosome 17 inferred from from 3,202 participants from the 1000 Genomes Whole
+Genome Sequencing dataset (1000G) \citep{1000G2015, 1000GWGS}. The genomic
+position of the mutations is shown on the X axis, and the age of mutation is
+represented on the Y axis. Panel B represents an inference after filtering out
+low site-density regions in the samples.}
     \label{fig:Supplementary_Figure_2}
@@
     \centering
-    \includegraphics[width=0.95\linewidth]{manuscript/figures/SuppFig3.png}
-    \caption{\textbf{A view of the Nodes page for a 1000 Genomes inference.}
-    A screenshot of \texttt{tsbrowse}'s Nodes view for an inference of the long arm of chromosome 20 from the 1000 Genomes whole-genome sequencing dataset.
-At the top is a plot of node spans over time.
-The length of sequence that the nodes span is shown on the X axis, and the time of nodes is shown on the Y axis.
-The histogram at the bottom shows the distribution of node spans.
-}
+    \includegraphics[width=0.95\linewidth]{figures/SuppFig3.png}
+    \caption{\textbf{A view of the Nodes page for a 1000 Genomes inference.} A
+screenshot of \texttt{tsbrowse}'s Nodes view for an inference of the long arm
+of chromosome 20 from the 1000 Genomes whole-genome sequencing dataset. At the
+top is a plot of node spans over time. The length of sequence that the nodes
+span is shown on the X axis, and the time of nodes is shown on the Y axis. The
+histogram at the bottom shows the distribution of node spans. }
     \label{fig:Supplementary_Figure_3}
@@
     \centering
-    \includegraphics[width=0.95\linewidth]{manuscript/figures/SuppFig4.png}
+    \includegraphics[width=0.95\linewidth]{figures/SuppFig4.png}
     \caption{\textbf{\texttt{tsbrowse} applied to SARS-CoV2 ARGs.}

### 2024-12-12 manuscript/paper.tex
Comment out missing fig

@@
 \citep{Haller2019,Baumdicker2021,Adrion2020,Lauterbur2023} to simulate ARGs
-under complex, realistic demographic models \cite{Anderson-Trocmé2023} now
+under complex, realistic demographic models \cite{Anderson-Trocme2023} now
 exist. However, models of simulation do not capture \textit{all} the nuances of
@@
     \centering
-    \includegraphics[width=0.95\linewidth]{figures/SuppFig2.png}
+    % \includegraphics[width=0.95\linewidth]{figures/SuppFig2.png}
     \caption{\textbf{A view of mutations in the 1000G WGS dataset.} A

### 2024-12-12 manuscript/paper.tex
fixup citations a bit

@@
 
+Ancestral recombination graphs (ARGs) describe how a set of sampled sequences 
+are related to each other at each position of the genome in a recombining
+species~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
+wong2024general}, and there has been significant recent progress 
+on inference through a range of approaches~\citep{rasmussen2014genome,
+speidel2019method,kelleher2019inferring,zhang2023biobank,gunnarsson2024scalable,
+deng2024robust}.
+
 Biobank-driven research has accelerated the development of several
@@
 (ARGs) from large-scale genetic data
-\citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
+% \citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
 ARGs encode the interwoven paths of genetic inheritance resulting from
 recombination, and can be represented as genetic genealogies along the sequence
-\citep{Wong2024,Hudson1983,Griffiths1997}. They capture rich detail about
+\citep{wong2024general,Hudson1983,Griffiths1997}. They capture rich detail about
 ancestral processes influencing the samples, and have garnered great interest
 in several applications in population and statistical genetics. These include
-the detection of ultra-rare genetic variation \citep{Zhang2023}, the mapping of
+the detection of ultra-rare genetic variation \citep{zhang2023biobank}, the mapping of
 quantitative trait loci \citep{Link2023}, and the sparse and efficient modeling
@@
 encode. \texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences
-\citep{Wong2024}. This is the native format of \texttt{tskit} \citep{tskit}, a
+\citep{wong2024general}. This is the native format of \texttt{tskit} \citep{tskit}, a
 widely-used library for interchanging ARGs inferred using a variety of methods
-\citep{Speidel2019,Zhang2023,Haller2019,Baumdicker2021,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
+\citep{speidel2019method,zhang2023biobank,Haller2019,Baumdicker2021,gunnarsson2024scalable,
+deng2024robust}.
 The interface is run on a web browser. It works on biobank-scale datasets
@@
 \texttt{tsbrowse} is the \texttt{tskit} data model. As \texttt{tsbrowse} is
-based on this generalised ARG representation \citep{Wong2024} it can be applied
+based on this generalised ARG representation \citep{wong2024general} it can be applied
 to ARGs from several widely-used inference methods including \texttt{Relate}
-\citep{Speidel2019}, \texttt{ARG-Needle} \citep{Zhang2023} and \texttt{SINGER}
-\citep{Deng2024.03.16.585351}. In \texttt{tskit}, ARGs are encoded as a
+\citep{speidel2019method}, \texttt{ARG-Needle} \citep{zhang2023biobank} 
+and \texttt{SINGER} \citep{deng2024robust}. In \texttt{tskit}, ARGs are encoded as a
 collection of tables, each of which store information about a specific ARG
@@
 inferences from empirical data
-\citep{Kelleher2019,Zhang2023,Speidel2019,Gunnarsson2024.08.31.610248}.
+\citep{kelleher2019inferring,zhang2023biobank,speidel2019method,
+gunnarsson2024scalable}.
 Sophisticated techniques
@@
 ARGs from the truth dataset described above: \texttt{tsinfer version 0.3.3}
-\citep{Kelleher2019}, \texttt{tsdate version 0.2.1}, \texttt{Relate version
-1.2.2} \cite{Speidel2019}, \texttt{ARG-needle version 1.0.3} \citep{Zhang2023},
-\texttt{SINGER version 0.1.8-beta} \citep{Deng2024.03.16.585351}. For all
+\citep{kelleher2019inferring}, \texttt{tsdate version 0.2.1}, 
+\texttt{Relate version 1.2.2} \cite{speidel2019method}, 
+\texttt{ARG-needle version 1.0.3} \citep{zhang2023biobank},
+\texttt{SINGER version 0.1.8-beta} \citep{deng2024robust}. 
+For all
 methods, the following parameters of inference were used: \textit{recombination

### 2024-12-12 manuscript/paper.tex

@@
 \keywords{Ancestral Recombination Graphs, Holoviz, biobank-scale ARG
-visualisation, genome browser} \maketitle \section{Introduction} Visualisation
+visualisation, genome browser} \maketitle
+
+\section{Introduction} 
+Ancestral recombination graphs (ARGs) describe how a set of sample
+sequences relate to each other at each position along the genome in a recombining
+species, and are currently the subject of intense research 
+interest~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
+wong2024general}. ARGs are a fundamental object in population genetics,
+and although they have been of theoretical interest for
+decades~\citep{Hudson1983,Griffiths1997} it is only with recent breakthroughs
+in inference methods~\citep{rasmussen2014genome,
+speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
+gunnarsson2024scalable,deng2024robust} that practical application has 
+become possible. Applications to varied questions such as the spatial 
+location of genetic
+ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2024geographic},
+
+as well as the use of ARGs as training data for machine learning 
+methods~\citep{whitehouse2024tree} depend critically on the \emph{accuracy}
+of the inferred ARGs. Yet, despite this explosion of downstream methods using
+inferred ARGs, there is a critical lack of practical tools to help
+researchers make informed assesments about the quality of their inferred ARGs.
+
+% Varied applications to
+% simulation~\citep{kelleher2016efficient,baumdicker2021efficient,Anderson-Trocme2023}
+
+Visualisation
 is an essential component of data analysis pipelines. In the past few decades,
@@
 
-Ancestral recombination graphs (ARGs) describe how a set of sampled sequences 
-are related to each other at each position of the genome in a recombining
-species~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
-wong2024general}, and there has been significant recent progress 
-on inference through a range of approaches~\citep{rasmussen2014genome,
-speidel2019method,kelleher2019inferring,zhang2023biobank,gunnarsson2024scalable,
-deng2024robust}.
-
-Biobank-driven research has accelerated the development of several
-technologies, including the ability to infer Ancestral Recombination Graphs
-(ARGs) from large-scale genetic data
-% \citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
-ARGs encode the interwoven paths of genetic inheritance resulting from
-recombination, and can be represented as genetic genealogies along the sequence
-\citep{wong2024general,Hudson1983,Griffiths1997}. They capture rich detail about
-ancestral processes influencing the samples, and have garnered great interest
-in several applications in population and statistical genetics. These include
-the detection of ultra-rare genetic variation \citep{zhang2023biobank}, the mapping of
-quantitative trait loci \citep{Link2023}, and the sparse and efficient modeling
-of linkage disequilibrium \citep{Nowbandegani2023}. It is evident that ARG
-inference is increasingly becoming a part of genetic data analysis pipelines.
-Yet, ARG visualisation, which is crucial for downstream analyses remains
-challenging due to the complexity of inferences from empirical data. Existing
-visualisation tools (see \hyperref[sec:Discussion]{Discussion}) focus on the
-semantic representation of ARGs, but such representations have limited utility
-in informing downstream analyses.
+% Biobank-driven research has accelerated the development of several
+% technologies, including the ability to infer Ancestral Recombination Graphs
+% (ARGs) from large-scale genetic data
+% % \citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
+% ARGs encode the interwoven paths of genetic inheritance resulting from
+% recombination, and can be represented as genetic genealogies along the sequence
+% \citep{wong2024general,Hudson1983,Griffiths1997}. They capture rich detail about
+% ancestral processes influencing the samples, and have garnered great interest
+% in several applications in population and statistical genetics. These include
+% the detection of ultra-rare genetic variation \citep{zhang2023biobank}, the mapping of
+% quantitative trait loci \citep{Link2023}, and the sparse and efficient modeling
+% of linkage disequilibrium \citep{Nowbandegani2023}. It is evident that ARG
+% inference is increasingly becoming a part of genetic data analysis pipelines.
+% Yet, ARG visualisation, which is crucial for downstream analyses remains
+% challenging due to the complexity of inferences from empirical data. Existing
+% visualisation tools (see \hyperref[sec:Discussion]{Discussion}) focus on the
+% semantic representation of ARGs, but such representations have limited utility
+% in informing downstream analyses.
 

### 2025-03-06 manuscript/paper.tex

@@
 with the changing landscape of genomics. Modern genome browsers can load
-datasets from a URL or local file system \citep{IGV-web,Lee2013,Robinson2023}.
+datasets from a URL or local file system \citep{Lee2013,Robinson2023}.
 The data security and scalability offered by this architecture ensures that
@@
 encode. \texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences
-\citep{wong2024general}. This is the native format of \texttt{tskit} \citep{tskit}, a
-widely-used library for interchanging ARGs inferred using a variety of methods
-\citep{speidel2019method,zhang2023biobank,Haller2019,Baumdicker2021,gunnarsson2024scalable,
-deng2024robust}.
+\citep{wong2024general}. 
+This the \texttt{tskit} library~\citep{ralph2020efficiently}
+and supported by 
+simulation~\citep{kelleher2016efficient,kelleher2018efficient,
+haller2019tree,baumdicker2022efficient,
+adrion2020community,lauterbur2023expanding,tsambos2023link,tagami2024tstrait}
+inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
+mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general}
+and downstream processing methods~\citep{link2023tree,nowbandegani2023extremely}
+
+% \citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
+
+% widely-used library for interchanging ARGs inferred using a variety of methods
+% \citep{Speidel2019,Zhang2023,Haller2019,Baumdicker2021,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
+
 The interface is run on a web browser. It works on biobank-scale datasets
@@
 meaningful and performant way. For example, a visual representation of millions
-of mutations in SARS-Cov-2 ARGs \citep{Zhan2023.06.08.544212} could quickly
+of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
 lead to performance bottlenecks and overplotting. Datashader
@@
 to ARGs from several widely-used inference methods including \texttt{Relate}
-\citep{speidel2019method}, \texttt{ARG-Needle} \citep{zhang2023biobank} 
-and \texttt{SINGER} \citep{deng2024robust}. In \texttt{tskit}, ARGs are encoded as a
+\citep{speidel2019method}, \texttt{ARG-Needle} \citep{zhang2023biobank} and \texttt{SINGER}
+\citep{deng2024robust}. In \texttt{tskit}, ARGs are encoded as a
 collection of tables, each of which store information about a specific ARG
@@
 between the sampled and ancestral sequences, while the Mutation table describes
-sequence variation contained in the samples \citep{Ralph2020}. This compact
+sequence variation contained in the samples. This compact
 representation enables the efficient storage and querying of biobank-scale
@@
 
+\subsection{Applications}
+
  Tsbrowse can be applied to versatile scenarios. Figure \ref{fig:Figure_2}
@@
 selective sweep dataset simulated using \texttt{msprime}
-\citep{Baumdicker2021}. The distinct pattern of edges observed in the middle of
+\citep{baumdicker2022efficient}. The distinct pattern of edges observed in the middle of
 the sequence between \textit{y=0} to \textit{y=2.5} is attributable to a burst
@@
 Supplementary Fig. \ref{fig:Supplementary_Figure_4}, a screenshot of the
-Mutations page for SARS-CoV-2 ARGs \citep{Zhan2023.06.08.544212} demonstrates
+Mutations page for SARS-CoV-2 ARGs \citep{zhan2023towards} demonstrates
 tsbrowse's ability to handle diverse data sources. This plot represents a total
@@
 
-In current practice, simulated ARGs are used as ground-truth to validate
-inferences from empirical data
-\citep{kelleher2019inferring,zhang2023biobank,speidel2019method,
-gunnarsson2024scalable}.
-Sophisticated techniques
-\citep{Haller2019,Baumdicker2021,Adrion2020,Lauterbur2023} to simulate ARGs
-under complex, realistic demographic models \cite{Anderson-Trocme2023} now
-exist. However, models of simulation do not capture \textit{all} the nuances of
-population datasets. This results in a clear gap between ARG inference and
-application. \texttt{Tsbrowse} captures ARG properties in a model-free way,
-providing the means to qualitatively assess inferences. With notable features
-of browsability along the genome, scalability and applicability to diverse data
-sources, inference and simulation methods, \texttt{tsbrowse} has the potential
-to be integrated into biobank-scale data pipelines to improve ARG inference and
-analysis.
+% In current practice, simulated ARGs are used as ground-truth to validate
+% inferences from empirical data
+% \citep{Kelleher2019,Zhang2023,Speidel2019,Gunnarsson2024.08.31.610248}.
+% Sophisticated techniques
+% \citep{haller2019tree,baumdicker2022efficient,adrion2020community,lauterbur2023expanding} to simulate ARGs
+% under complex, realistic demographic models \cite{Anderson-Trocme2023} now
+% exist. However, models of simulation do not capture \textit{all} the nuances of
+% population datasets. This results in a clear gap between ARG inference and
+% application. \texttt{Tsbrowse} captures ARG properties in a model-free way,
+% providing the means to qualitatively assess inferences. With notable features
+% of browsability along the genome, scalability and applicability to diverse data
+% sources, inference and simulation methods, \texttt{tsbrowse} has the potential
+% to be integrated into biobank-scale data pipelines to improve ARG inference and
+% analysis.
 
@@
 The ARG shown in \ref{fig:Supplementary_Figure_4} was inferred with
-\texttt{sc2ts}~\citep{Zhan2023.06.08.544212} using the Viridian
+\texttt{sc2ts}~\citep{zhang2023biobank} using the Viridian
 dataset~\citep{hunt2024addressing}. It consists of 516,443 samples, 612,209

### 2025-03-07 manuscript/paper.tex

@@
 gunnarsson2024scalable,deng2024robust} that practical application has 
-become possible. Applications to varied questions such as the spatial 
+become possible. Varied applications
+such as inferring selection~\citep{stern2019approximate,hejase2022deep} and 
+the spatial 
 location of genetic
 ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2024geographic},
+more powerful approaches to quantifying genetic 
+relatedness~\citep{fan2022genealogical,zhang2023biobank,gunnarsson2024scalable,lehmann2025on}
+and other methodological improvements for genome wide association 
+studies~\citep{nowbandegani2023extremely,link2023tree},
+and the development of machine learning methods using inferred
+ARGs as input~\citep{hejase2022deep,pearson2023local,korfmann2024simultaneous,
+whitehouse2024tree} all depend critically on the \emph{accuracy}
+of the inferred ARGs. 
+Although studies benchmarking the various inference methods on 
+simulated data have emerged 
+\citep{brandt2022evaluation,deng2024robust,peng2024evaluating},
+the practicalities of applying ARG inference to real data is 
+understudied, and there is a critial lack of software infrastructure
+to support evaluating the quality of estimated ARGs.
 
-as well as the use of ARGs as training data for machine learning 
-methods~\citep{whitehouse2024tree} depend critically on the \emph{accuracy}
-of the inferred ARGs. Yet, despite this explosion of downstream methods using
-inferred ARGs, there is a critical lack of practical tools to help
-researchers make informed assesments about the quality of their inferred ARGs.
+Visualisation is fundamentally important to data analysis.
+Many specialised tools exists to aid the visual analysis and 
+quality control
+of assembly~\citep{wick2015bandage,challis2020blobtoolkit},
+read mapping~\citep{robinson2011integrative},
+and variant
+calling~\citep{robinson2017variant,tollefson2019viva,konig2023divbrowse},
+for example.
+At every stage of a bioinformatics pipeline, it is important to visualise 
+results to avoid artefacts and aid understanding of the data.
+Genome browsers such as IGV~\citep{robinson2011integrative} and the 
+UCSC Genome Browser~\citep{nassar2023ucsc} integrate many different
+data modalities, and are vital infrastructure for the field.
+% . In the past few decades,
+% genome browsers have become synonymous with sequence visualisation, by
+% providing user-friendly, interactive interfaces to navigate genomic sequences
+% and annotations. From early standalone tools to manage \textit{C. elegans} data
+% \citep{Eeckman1995}, to modern web-apps for vertebrate genomes
+% \citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
+% with the changing landscape of genomics. Modern genome browsers can load
+% datasets from a URL or local file system \citep{Lee2013,Robinson2023}.
+% The data security and scalability offered by this architecture ensures that
+% genome browsers continue to be relevant in the biobank era.
 
-% Varied applications to
-% simulation~\citep{kelleher2016efficient,baumdicker2021efficient,Anderson-Trocme2023}
+There is currently no straightforward means of visually summarising
+ARGs, presenting a significant stumbling block for the nascent 
+field of practical ARG inference. Inferred ARGs are essentially opaque,
+with only the most basic numerical summaries (such as numbers 
+of nodes, mutations etc) or high-level
+statistics~\citep{ralph2020efficiently} available. While tools 
+for visualising the local tree topologies exist, they are difficult
+to interpret and do not scale well to large sample sizes. 
+To address this gap, we present \texttt{tsbrowse}, a
+% I worry that if we say "genome browser" we'll get stung for not having 
+% deep integration with all sorts of stuff
+client-server application providing genome browser-like 
+functionality for ARGs. 
+It provides interactive visualisations of the information structure 
+of ARGs, smoothly scrolling from chromosome-level views down to 
+individual nodes, edges and mutations. Supporting very large ARGs is a 
+particular focus for \texttt{tsbrowse}, as millions of genome
+sequences have been sampled for several
+species~\citep{cesarani2022multibreed,stark2024call,hunt2024addressing}
+and ARGs of this scale being actively
+inferred~\citep{kelleher2019inferring,zhang2023biobank,
+zhan2023towards,gunnarsson2024scalable}.
 
-Visualisation
-is an essential component of data analysis pipelines. In the past few decades,
-genome browsers have become synonymous with sequence visualisation, by
-providing user-friendly, interactive interfaces to navigate genomic sequences
-and annotations. From early standalone tools to manage \textit{C. elegans} data
-\citep{Eeckman1995}, to modern web-apps for vertebrate genomes
-\citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
-with the changing landscape of genomics. Modern genome browsers can load
-datasets from a URL or local file system \citep{Lee2013,Robinson2023}.
-The data security and scalability offered by this architecture ensures that
-genome browsers continue to be relevant in the biobank era.
+% Its focus is on interactively visualising the
+% information structure in ARGs at all scales.
 
-% Biobank-driven research has accelerated the development of several
-% technologies, including the ability to infer Ancestral Recombination Graphs
-% (ARGs) from large-scale genetic data
-% % \citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
-% ARGs encode the interwoven paths of genetic inheritance resulting from
-% recombination, and can be represented as genetic genealogies along the sequence
-% \citep{wong2024general,Hudson1983,Griffiths1997}. They capture rich detail about
-% ancestral processes influencing the samples, and have garnered great interest
-% in several applications in population and statistical genetics. These include
-% the detection of ultra-rare genetic variation \citep{zhang2023biobank}, the mapping of
-% quantitative trait loci \citep{Link2023}, and the sparse and efficient modeling
-% of linkage disequilibrium \citep{Nowbandegani2023}. It is evident that ARG
-% inference is increasingly becoming a part of genetic data analysis pipelines.
-% Yet, ARG visualisation, which is crucial for downstream analyses remains
-% challenging due to the complexity of inferences from empirical data. Existing
-% visualisation tools (see \hyperref[sec:Discussion]{Discussion}) focus on the
-% semantic representation of ARGs, but such representations have limited utility
-% in informing downstream analyses.
+% \texttt{Tsbrowse} provides interactive visual summaries of  the
+% fundamental properties of ARGs, i.e. the mutations, nodes, edges and trees they
+% encode.
 
-To address this gap, we have developed \texttt{tsbrowse}, an open-source app
-intended as a genome browser for ARGs. Its  focus is on visualising the
-information structure in ARGs, thereby improving  inference and analysis at
-biobank-scale. \texttt{Tsbrowse} provides interactive visual summaries of  the
-fundamental properties of ARGs, i.e. the mutations, nodes, edges and trees they
-encode. \texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences
-\citep{wong2024general}. 
-This the \texttt{tskit} library~\citep{ralph2020efficiently}
-and supported by 
-simulation~\citep{kelleher2016efficient,kelleher2018efficient,
-haller2019tree,baumdicker2022efficient,
-adrion2020community,lauterbur2023expanding,tsambos2023link,tagami2024tstrait}
-inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
-mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general}
-and downstream processing methods~\citep{link2023tree,nowbandegani2023extremely}
+% \texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences
+% \citep{wong2024general}. 
+% This the \texttt{tskit} library~\citep{ralph2020efficiently}
+% and supported by 
+% simulation~\citep{kelleher2016efficient,kelleher2018efficient,
+% haller2019tree,baumdicker2022efficient,
+% adrion2020community,lauterbur2023expanding,korfmann2023weak,tsambos2023link,tagami2024tstrait}
+% inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
+% mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general}
+% and downstream processing methods~\citep{fan2022genealogical,nowbandegani2023extremely}
 
@@
 
-The interface is run on a web browser. It works on biobank-scale datasets
-without requiring \textcolor{red}{raw data uploads to a server}. This is ideal
-for performing quality control and exploratory analyses on ARGs inferred from
-biobank datasets. \section{Results} \subsection{Architecture} \texttt{Tsbrowse}
+% The interface is run on a web browser. It works on biobank-scale datasets
+% without requiring \textcolor{red}{raw data uploads to a server}. This is ideal
+% for performing quality control and exploratory analyses on ARGs inferred from
+% biobank datasets.
+
+\section{Results} \subsection{Architecture} \texttt{Tsbrowse}
 is run in a step-wise approach (Figure \ref{fig:Figure_1}), which makes it

### 2025-03-07 manuscript/paper.tex

@@
 
-% Its focus is on interactively visualising the
-% information structure in ARGs at all scales.
-
-% \texttt{Tsbrowse} provides interactive visual summaries of  the
-% fundamental properties of ARGs, i.e. the mutations, nodes, edges and trees they
-% encode.
-
-% \texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences
-% \citep{wong2024general}. 
-% This the \texttt{tskit} library~\citep{ralph2020efficiently}
-% and supported by 
-% simulation~\citep{kelleher2016efficient,kelleher2018efficient,
-% haller2019tree,baumdicker2022efficient,
-% adrion2020community,lauterbur2023expanding,korfmann2023weak,tsambos2023link,tagami2024tstrait}
-% inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
-% mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general}
-% and downstream processing methods~\citep{fan2022genealogical,nowbandegani2023extremely}
-
 % \citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
@@
 
-\section{Results} \subsection{Architecture} \texttt{Tsbrowse}
-is run in a step-wise approach (Figure \ref{fig:Figure_1}), which makes it
-convenient to handle large datasets. The initial step (\texttt{tsbrowse
+\section{Results}
+
+\begin{figure} \centering
+\includegraphics[width=0.95\linewidth]{figures/MainFig1.png}
+\caption{\textbf{Overview of tsbrowse architecture. } In tskit, ARGs are
+defined by a collection of tables. Exemplar
+table names are denoted in dark blue and columns that describe the property in
+light blue. In the pre-process step, tables are augmented with additional
+information computed for each property (yellow). The output from this
+pre-processing step is stored as a \texttt{.tsbrowse} file. Next the serve step
+renders the visualisation in a web browser by leveraging tools in the Holoviz
+ecosystem. Optional annotations are provided as an input file, allowing users
+to overlay contextual information about genes or other sequence features. }
+\label{fig:Figure_1} \end{figure}
+
+\subsection{Data model} \label{subsec:Data_Model} 
+\texttt{Tsbrowse} uses the ``succinct tree sequence'' encoding of 
+ARGS~\citep{wong2024general}. This efficient ARG encoding 
+is implemented by the 
+\texttt{tskit} library~\citep{ralph2020efficiently}
+and supported by most modern ARG
+simulation~\citep{kelleher2016efficient,kelleher2018efficient,
+haller2019tree,baumdicker2022efficient,
+adrion2020community,lauterbur2023expanding,korfmann2023weak,
+tsambos2023link,tagami2024tstrait},
+inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
+mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general},
+and processing methods~\citep{fan2022genealogical,nowbandegani2023extremely}.
+In \texttt{tskit}, ARGs are encoded as a
+collection of tables, storing information about a specific ARG
+property (Fig.~\ref{fig:Figure_1}).
+% For example, the Node table describes sampled or ancestral sequences
+% and records their time of birth, and the individual and population they belong
+% to. The Edge table describes the relationship between pairs of nodes, and
+% records the node IDs and the genomic locations over which the edge spans. The
+% Mutation table records mutation events in the history of the sample, including
+% the change of allelic state, time of event and node in which the event
+% occurred. 
+The Node and Edge tables describe the genealogical relationships
+between the sampled and ancestral sequences, 
+while the Mutation table describes
+sequence variation contained in the samples. 
+\texttt{Tsbrowse} augments these tables with additional columns to create
+a compressed \texttt{.tsbrowse} file, containing all necessary semantic
+metadata required for automatic visualisation.
+% (see \hyperref[subsec:User_Interface]{User Interface}).
+
+\subsection{Architecture} 
+\texttt{Tsbrowse} is a client-server application (Fig.~\ref{fig:Figure_1}).
+The server is a command-line interface program written in Python, which
+is run on the computer where the ARG we wish to visualise is available.
+The client is a web-browser, which may optionally be running 
+on a different machine. There are two important advantages to this 
+approach over a monolithic single-machine architecture. Firstly, 
+the ARG being visualised remains in-situ and does not need to be 
+downloaded from the server; summaries of the ARG at various 
+scales are sent to the client on demand.
+Secondly, this provides the flexibility to use a powerful server
+when generating summaries of ARG while using a standard desktop 
+machine for visualisation.
+
+\texttt{Tsbrowse} is run in a step-wise approach (Fig.~\ref{fig:Figure_1}), 
+which makes it convenient to handle large datasets. 
+The initial step (\texttt{tsbrowse
 preprocess}) prepares the data for visualisation. It can be performed
@@
 experience.
-
-\begin{figure} \centering
-\includegraphics[width=0.95\linewidth]{figures/MainFig1.png}
-\caption{\textbf{Overview of tsbrowse architecture. } In tskit, ARGs are
-defined by a collection of tables which broadly map to ARG properties. Exemplar
-table names are denoted in dark blue and columns that describe the property in
-light blue. In the pre-process step, tables are augmented with additional
-information computed for each property (yellow). The output from this
-pre-processing step is stored as a \texttt{.tsbrowse} file. Next the serve step
-renders the visualisation in a web browser by leveraging tools in the Holoviz
-ecosystem. Optional annotations are provided as an input file, allowing users
-to overlay contextual information about genes or other sequence features. }
-\label{fig:Figure_1} \end{figure}
-
-\subsection{Data model} \label{subsec:Data_Model} The foundation of
-\texttt{tsbrowse} is the \texttt{tskit} data model. As \texttt{tsbrowse} is
-based on this generalised ARG representation \citep{wong2024general} it can be applied
-to ARGs from several widely-used inference methods including \texttt{Relate}
-\citep{speidel2019method}, \texttt{ARG-Needle} \citep{zhang2023biobank} and \texttt{SINGER}
-\citep{deng2024robust}. In \texttt{tskit}, ARGs are encoded as a
-collection of tables, each of which store information about a specific ARG
-property. For example, the Node table describes sampled or ancestral sequences
-and records their time of birth, and the individual and population they belong
-to. The Edge table describes the relationship between pairs of nodes, and
-records the node IDs and the genomic locations over which the edge spans. The
-Mutation table records mutation events in the history of the sample, including
-the change of allelic state, time of event and node in which the event
-occurred. The Node and Edge tables describe the genealogical relationships
-between the sampled and ancestral sequences, while the Mutation table describes
-sequence variation contained in the samples. This compact
-representation enables the efficient storage and querying of biobank-scale
-ARGs. \texttt{Tsbrowse} augments these tables with additional columns to create
-a compressed \texttt{.tsbrowse} file, containing all necessary semantic
-metadata required for automatic visualisation (see
-\hyperref[subsec:User_Interface]{User Interface}).
 

### 2025-03-11 manuscript/paper.tex
Pass through the architecture section

@@
 \subsection{Architecture} 
-\texttt{Tsbrowse} is a client-server application (Fig.~\ref{fig:Figure_1}).
-The server is a command-line interface program written in Python, which
-is run on the computer where the ARG we wish to visualise is available.
-The client is a web-browser, which may optionally be running 
-on a different machine. There are two important advantages to this 
-approach over a monolithic single-machine architecture. Firstly, 
+\texttt{Tsbrowse} is a modular client-server application 
+written in Python, optimised for handling ARGs with millions
+of samples (Fig.~\ref{fig:Figure_1}). 
+To ensure that such large ARGs can be visualised quickly
+we perform a preprocessing step (\texttt{tsbrowse preprocess})
+which augments the \texttt{tskit} 
+data model with some additional derived information (e.g., the number 
+of inheritors of each mutation).
+To visualise an ARG we then run \texttt{tsbrowse serve}, which 
+by default will open a browser window on the local machine, but 
+also supports running the server on a remote machine across the 
+network.
+This client-server architecture has some important advantages over
+a monolithic single-machine approach. Most importantly,
 the ARG being visualised remains in-situ and does not need to be 
-downloaded from the server; summaries of the ARG at various 
-scales are sent to the client on demand.
-Secondly, this provides the flexibility to use a powerful server
-when generating summaries of ARG while using a standard desktop 
-machine for visualisation.
+downloaded from the server.
 
-\texttt{Tsbrowse} is run in a step-wise approach (Fig.~\ref{fig:Figure_1}), 
-which makes it convenient to handle large datasets. 
-The initial step (\texttt{tsbrowse
-preprocess}) prepares the data for visualisation. It can be performed
-separately on a large machine, and the output stored for later use. The second
-step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
-and renders visualisations. This step follows a client-server architecture. On
-the server side (the machine on which \texttt{tsbrowse serve} is run), the
-Holoviz suite of libraries \citep{Holoviz} is leveraged to create
-visualisations based on the pre-processed data. This choice of architecture is
-prompted by the need to represent properties of large-scale ARGs in a
-meaningful and performant way. For example, a visual representation of millions
-of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
-lead to performance bottlenecks and overplotting. Datashader
-\citep{Datashader}, a component Holoviz package, overcomes this challenge by
-rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
-for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
-tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
-server to render and interact with these visualisations. Heavy data processing
-is limited to the server side, keeping the client-side interface lightweight.
-For large datasets with complex visualisations, this achieves a smoother user
-experience.
+The express goal of \texttt{tsbrowse} is to provide interpretable
+and interactive visual summaries of ARGs containing millions of nodes, edges 
+and mutations. Mutations, for example, have a clear interpretation
+when plotted on genome coordinate vs time axes, but at this scale 
+the data density is far too high to simply plot each mutation as a point.
+We overcome this problem by using Datashader and the wider 
+Holoviz ecosystem~\citep{Holoviz}, which efficiently rasterises large datasets
+at the requested resolution on the server, sends the image to the web-browser
+for display, and dynamically updates as the user interatively navigates the ARG.
+
+% and renders visualisations. This step follows a client-server architecture. On
+% the server side (the machine on which \texttt{tsbrowse serve} is run), the
+% Holoviz suite of libraries \citep{Holoviz} is leveraged to create
+% visualisations based on the pre-processed data. This choice of architecture is
+% prompted by the need to represent properties of large-scale ARGs in a
+% meaningful and performant way. For example, a visual representation of millions
+% of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
+% lead to performance bottlenecks and overplotting. Datashader
+% \citep{Datashader}, a component Holoviz package, overcomes this challenge by
+% rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
+% for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
+% tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
+
+% \texttt{Tsbrowse} is run in a step-wise approach (Fig.~\ref{fig:Figure_1}), 
+% which makes it convenient to handle large datasets. 
+% The initial step (\texttt{tsbrowse
+% preprocess}) prepares the data for visualisation. It can be performed
+% separately on a large machine, and the output stored for later use. The second
+% step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
+% and renders visualisations. This step follows a client-server architecture. On
+% the server side (the machine on which \texttt{tsbrowse serve} is run), the
+% Holoviz suite of libraries \citep{Holoviz} is leveraged to create
+% visualisations based on the pre-processed data. This choice of architecture is
+% prompted by the need to represent properties of large-scale ARGs in a
+% meaningful and performant way. For example, a visual representation of millions
+% of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
+% lead to performance bottlenecks and overplotting. Datashader
+% \citep{Datashader}, a component Holoviz package, overcomes this challenge by
+% rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
+% for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
+% tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
+% server to render and interact with these visualisations. Heavy data processing
+% is limited to the server side, keeping the client-side interface lightweight.
+% For large datasets with complex visualisations, this achieves a smoother user
+% experience.
 
@@
 \label{subsec:User_Interface}
-
 \begin{figure}
@@
 \end{figure}
-
-\texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem
-\citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh},
-\texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data
-visualisation pipelines even for large datasets. A key challenge with
-large-scale data visualisation is to maintain a responsive user interface.
-\texttt{Datashader} aggregates data and renders only the necessary data points
-at a given zoom level, thereby minimising the computational load on the
-browser. A \texttt{Bokeh} plotting back-end provides useful tools and widgets.
-The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity
-with individual data points, allowing users to gain deeper insights from the
-data. The \texttt{serve} step accepts an optional input file containing
-sequence annotations. This allows users to overlay contextual information about
-genes or other sequence features.
-
-The user interface is presented as a \texttt{Panel} dashboard, with pages to
+% \texttt{Tsbrowse}'s interface is built on the \texttt{Holoviz} ecosystem
+% \citep{Holoviz}, which brings together \texttt{Datashader}, \texttt{Bokeh},
+% \texttt{Panel}, \texttt{Holoviews} and other libraries to create seamless data
+% visualisation pipelines even for large datasets. A key challenge with
+% large-scale data visualisation is to maintain a responsive user interface.
+% \texttt{Datashader} aggregates data and renders only the necessary data points
+% at a given zoom level, thereby minimising the computational load on the
+% browser. A \texttt{Bokeh} plotting back-end provides useful tools and widgets.
+% The \textit{pan}, \textit{select} and \textit{zoom} tools enable interactivity
+% with individual data points, allowing users to gain deeper insights from the
+% data. The \texttt{serve} step accepts an optional input file containing
+% sequence annotations. This allows users to overlay contextual information about
+% genes or other sequence features. 
+The user interface is presented as a dashboard, with pages to
 describe mutations, nodes, edges and trees in the input ARG. Figure 2

### 2025-03-27 manuscript/paper.tex
polish intro

@@
 gunnarsson2024scalable,deng2024robust} that practical application has 
-become possible. Varied applications
+become possible. Varied applications have been recently proposed,
 such as inferring selection~\citep{stern2019approximate,hejase2022deep} and 
@@
 ARGs as input~\citep{hejase2022deep,pearson2023local,korfmann2024simultaneous,
-whitehouse2024tree} all depend critically on the \emph{accuracy}
-of the inferred ARGs. 
-Although studies benchmarking the various inference methods on 
+whitehouse2024tree}.
+While these developements are exciting, the performance 
+of these new methods depend critically on the accuracy of the inferred ARGs,
+and although studies benchmarking the various inference methods on 
 simulated data have emerged 
@@
 the practicalities of applying ARG inference to real data is 
-understudied, and there is a critial lack of software infrastructure
-to support evaluating the quality of estimated ARGs.
+understudied. In particular, there is a critial lack of software 
+infrastructure to support evaluation and quality control of 
+inferred ARGs.
 
 Visualisation is fundamentally important to data analysis.
-Many specialised tools exists to aid the visual analysis and 
+Many specialised tools exist to aid the visual analysis and 
 quality control
@@
 species~\citep{cesarani2022multibreed,stark2024call,hunt2024addressing}
-and ARGs of this scale being actively
-inferred~\citep{kelleher2019inferring,zhang2023biobank,
-zhan2023towards,gunnarsson2024scalable}.
+and ARGs of this scale are a particular focus of ongoing 
+research~\citep{kelleher2019inferring,zhang2023biobank,
+zhan2023towards,anderson2023on,gunnarsson2024scalable}.
 

### 2025-03-27 manuscript/paper.tex
polish data model & architecture

@@
 In \texttt{tskit}, ARGs are encoded as a
-collection of tables, storing information about a specific ARG
-property (Fig.~\ref{fig:Figure_1}).
+collection of tables, storing information about the 
+nodes and edges that describe the graph topology and the sites 
+and mutations encoding the sequence variation (Fig.~\ref{fig:Figure_1}).
 % For example, the Node table describes sampled or ancestral sequences
@@
 % occurred. 
-The Node and Edge tables describe the genealogical relationships
-between the sampled and ancestral sequences, 
-while the Mutation table describes
-sequence variation contained in the samples. 
-\texttt{Tsbrowse} augments these tables with additional columns to create
-a compressed \texttt{.tsbrowse} file, containing all necessary semantic
-metadata required for automatic visualisation.
+% The Node and Edge tables describe the genealogical relationships
+% between the sampled and ancestral sequences, 
+% while the Mutation table describes
+% sequence variation contained in the samples. 
 % (see \hyperref[subsec:User_Interface]{User Interface}).
@@
 \texttt{Tsbrowse} is a modular client-server application 
-written in Python, optimised for handling ARGs with millions
+written in Python, optimised for ARGs with millions
 of samples (Fig.~\ref{fig:Figure_1}). 
-To ensure that such large ARGs can be visualised quickly
-we perform a preprocessing step (\texttt{tsbrowse preprocess})
-which augments the \texttt{tskit} 
-data model with some additional derived information (e.g., the number 
-of inheritors of each mutation).
+The application has two basic commands: \texttt{preprocess}
+and \texttt{serve}. The \texttt{preprocess} command 
+takes an input ARG file and augments the \texttt{tskit} 
+tables with additional columns,
+precomputing all the information required for visualisation,
+and storing as a compressed \texttt{.tsbrowse} file.
+(The \texttt{.tsbrowse} file is also a valid input for 
+\texttt{tszip}, a general utility for compressing \texttt{tskit}
+files.)
 To visualise an ARG we then run \texttt{tsbrowse serve}, which 
@@
 
-The express goal of \texttt{tsbrowse} is to provide interpretable
+The goal of \texttt{tsbrowse} is to provide interpretable
 and interactive visual summaries of ARGs containing millions of nodes, edges 
@@
 for display, and dynamically updates as the user interatively navigates the ARG.
+This approach allows us to summarise very large ARGs interactively;
+Fig~\ref{fig:Supplementary_Figure_4}, for example, shows a screenshot of 
+\texttt{tsbrowse} summarising almost 1 million mutations along the SARS-Cov-2
+genome.
+
+% Fig~\ref{fig:Figure_2} shows an example of how edges are visualised 
+% under this
 

### 2025-03-27 manuscript/paper.tex
Polish UI section

@@
 
-\subsection{User Interface}
-\label{subsec:User_Interface}
-\begin{figure}
-    \centering
-    \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
-    \caption{\textbf{A view of ARG edges in \texttt{tsbrowse}.}
-    A screenshot of the Edges tab for a 5 Mb sequence simulated with selection coefficient, \textit{s = 0.25}, and beneficial allele situated at \textit{x = 2.5 Mb}.
-The plot on the left represents each edge as a horizontal line along the sequence (X-axis).
-The time of parent node is shown on the Y-axis.
-The persistence of edges across various dimensions is shown through histograms on the right: top, distribution of genomic spans; middle, distribution of time spans; bottom, edge area, which is the product of genomic and time spans.}
-
-    \label{fig:Figure_2}
+\subsection{User Interface} \label{subsec:User_Interface} \begin{figure}
+\centering \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
+\caption{Screenshot of the Edges tab in \texttt{tsbrowse}.
+Visualisation of the edges in a simulated ARG with a strong selective
+sweep in the middle of the genome (see Supplement for details).
+Edges are shown in the main browser pane on the left, with additional
+histograms summarising the edges on the right.
+The effect of the sweep can be seen by the lack of edges crossing the 
+centre of the simulated chromosome. Other tabs are accessed by the 
+buttons on the top bar.
+\label{fig:Figure_2}}
 \end{figure}
@@
 The user interface is presented as a dashboard, with pages to
-describe mutations, nodes, edges and trees in the input ARG. Figure 2
-demonstrates the user interface using the Edges page as an example. The plot on
-the left is an overview of all 33,929 edges (parent-child relationships between
-node-pairs; see \hyperref[subsec:Data_Model]{Data Model}) in a simulated ARG
-\hyperref[subsec:sweep_simulation]{described in Supplementary Methods}; each
-edge is depicted as a horizontal line connecting the genomic coordinates of the
-parent and child nodes on the x-axis. The y-axis denotes the age of either
-node. A useful feature of the app is the ability to interact with the
-visualisations. The tools on the right of the plot allow the user to examine
-edges within a sequence/time slice, and query individual edges. The histograms
-to the right provide summaries of edge spans over sequence and time.
+describe various aspects of the ARG. 
+Figure~\ref{fig:Figure_2}
+demonstrates the user interface using the Edges page as an example. 
+The plot on the left is an overview of the 33,929 edges 
+(parent-child relationships between nodes along the genome)
+in a simulated ARG with a strong selective sweep (see
+Sec.~\ref{subsec:sweep_simulation} for details).
+Each edge is depicted as a horizontal line connecting the genomic coordinates of the
+parent and child nodes on the x-axis, and the y-axis shows the time of 
+either the parent or the child, as chosen by the user.
+The user can interact with the main ``genome browser'' window
+using a set of controls on the top-right corner (provided by Bokeh),
+allowing the user to pan and zoom as required. The histograms to the
+right then summarise the edges as depicted in the browser window.
+A similar browser interface is provided for 
+nodes
+(e.g. Fig~\ref{fig:Supplementary_Figure_3}) 
+and mutations 
+(e.g. Fig~\ref{fig:Supplementary_Figure_4}).
+\texttt{Tsbrowse} also provides an interactive table viewer with flexible
+search and sorting utilities, which is a valuable debugging utility for 
+developers.
 
@@
 \clearpage \onecolumn \setcounter{figure}{0}  % Reset the figure counter
-\renewcommand{\thefigure}{S\arabic{figure}}  % Change figure numbering to S1,
-S2, etc. \section{Supplementary Material} \subsection{Simulation of truth
-dataset (Fig. \ref{fig:Figure_2})} \label{subsec:sweep_simulation} Ancestral
+\renewcommand{\thefigure}{S\arabic{figure}}  
+% Change figure numbering to S1, S2, etc.
+
+\section{Supplementary Material} 
+\subsection{Simulation of truth dataset (Fig. \ref{fig:Figure_2})} \label{subsec:sweep_simulation} 
+Ancestral
 histories of 300 samples were simulated with \texttt{SweepGenicSelection}

### 2025-03-27 manuscript/paper.tex
Good opening para for Applications

@@
-%%
+%Fig~\ref{fig:Figure_2} %
 %% Copyright 2022 OXFORD UNIVERSITY PRESS
@@
 \subsection{Applications}
-
- Tsbrowse can be applied to versatile scenarios. Figure \ref{fig:Figure_2}
-highlights a use-case in visually characterising ARGs. The input ARG is a
-selective sweep dataset simulated using \texttt{msprime}
-\citep{baumdicker2022efficient}. The distinct pattern of edges observed in the middle of
-the sequence between \textit{y=0} to \textit{y=2.5} is attributable to a burst
-in recent coalescent events, showing that tsbrowse captures this characteristic
-of a selective sweep. Supplementary Figure \ref{fig:Supplementary_Figure_1} is
-another example of characterising ARGs with tsbrowse. The view of edges from
-four different inferences of the same dataset (inferred with \texttt{tsinfer} +
-\texttt{tsdate} (Panel A), \texttt{Relate} (Panel B), \texttt{ARG-Needle}
-(Panel C) and \texttt{SINGER} (Panel D)), reflects qualitative differences
-between inference methods.
+The purpose of \texttt{tsbrowse} is to provide an interactive view of the 
+\texttt{tskit} ARG data model to guide intuition, improve inference quality control
+and facilitate debugging. An example of how we can deepen our understanding
+using the genome-browser-like perspective provided by \texttt{tsbrowse}
+is provided the simulated ARG of Fig~\ref{fig:Figure_2}, where 
+the gap in the edges view corresponds to the characteristic dip in diversity
+of a selective sweep. Comparing this ground truth to the ARGs inferred
+by four different inference methods in Fig~\ref{fig:Supplementary_Figure_1},
+we can see that there are substantial qualitative differences between the
+results. These differences in ancestral haplotypes illustrated by 
+the edges view are unlikely to be captured by the tree-by-tree distance
+metrics usually used to evaluate
+inference methods~\citep[e.g.][]{kelleher2019inferring,zhang2023biobank},
+providing further motivation for new and improved ways to compare
+simulated and inferred ARGs~\citep{fritze2024forest}.
 

### 2025-03-27 manuscript/paper.tex
Good version of Applications section

@@
 Fig~\ref{fig:Supplementary_Figure_4}, for example, shows a screenshot of 
-\texttt{tsbrowse} summarising almost 1 million mutations along the SARS-Cov-2
+\texttt{tsbrowse} summarising almost 1 million mutations along the SARS-CoV-2
 genome.
@@
 
-We demonstrate a further application of tsbrowse in validating inferences
-(Supplementary Figures \ref{fig:Supplementary_Figure_2} and
-\ref{fig:Supplementary_Figure_3}). Supplementary Figure
-\ref{fig:Supplementary_Figure_2} shows how mutation patterns in a 2D plot of
-genomic position and mutation age are helpful in spotting poorly inferred
-regions. Excluding regions with low site density from the variant data improves
-the quality of inference (Panel B). Supplementary Figure
-\ref{fig:Supplementary_Figure_3} shows the genomic spans of ancestral nodes in
-an ARG inferred with tsinfer. Population genetics theory predicts that IBD
-(Identical By Descent) segments break down over time, resulting in shorter
-ancestral segments further back in the history of the samples. At time,
-\textit{y \textgreater{0.8}}, the plot reveals a pattern of node spans contrary
-to this expectation. It confirms a known algorithmic artefact due to which
-disproportionately long old ancestors are generated.
+Interest in ARG-based methods is burgeoning, but the methods are new 
+and practical guidance on applying inferences to real data is lacking.
+Data filtering is essential, and the effects of the 
+choices that must be made along any bioinformatics pipeline on the final
+ARG are hard to predict and quantify. \texttt{Tsbrowse} was primarily
+developed as a way to quickly visualise the effects of such filtering
+choices on ARGs inferred by \texttt{tsinfer}, and it is now an indispensible
+element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_2} 
+shows [a before and after view for 1000G data, where tsbrowse helped 
+diagnose and resolve an opaque data problem].
+These interactive visualisations have also helped diagnose issues with
+the \texttt{tsinfer} inference algorithm. 
+Fig~\ref{fig:Supplementary_Figure_3} shows the genomic spans of 
+ancestral nodes in an ARG inferred with tsinfer, demonstrating a clear 
+excess of long haplotypes in the very ancient past. These insights 
+have helped guide development and 
+may lead to significant improvements in performance.
 
-Supplementary Fig. \ref{fig:Supplementary_Figure_4}, a screenshot of the
-Mutations page for SARS-CoV-2 ARGs \citep{zhan2023towards} demonstrates
-tsbrowse's ability to handle diverse data sources. This plot represents a total
-of \textcolor{red}{X million} mutations from an inference of
-\textcolor{red}{1.27 million} sequences. After executing the
-\texttt{preprocess} step, the page loads in \textcolor{red}{X} seconds on a
-system with \textcolor{red}{Y GB RAM}, showing that it is possible to visualise
-large-scale datasets in a performant way. Panel B demonstrates a useful
-interactive feature; additional information about a mutation event (for
-example, the mutational frequency in predefined populations, or the number of
-sample nodes carrying the mutation) is displayed on mouse-over of individual
-data points.
+% . Population genetics theory predicts that IBD
+% (Identical By Descent) segments break down over time, resulting in shorter
+% ancestral segments further back in the history of the samples. At time,
+% \textit{y \textgreater{0.8}}, the plot reveals a pattern of node spans contrary
+% to this expectation. It confirms a known algorithmic artefact due to which
+% disproportionately long old ancestors are generated.
+
+% We demonstrate a further application of tsbrowse in validating inferences
+% (Supplementary Figures 
+% \ref{fig:Supplementary_Figure_3}). Supplementary Figure
+% \ref{fig:Supplementary_Figure_2} shows how mutation patterns in a 2D plot of
+% genomic position and mutation age are helpful in spotting poorly inferred
+% regions. Excluding regions with low site density from the variant data improves
+% the quality of inference (Panel B). 
+% Supplementary Figure
+% \ref{fig:Supplementary_Figure_3} shows the genomic spans of ancestral nodes in
+% an ARG inferred with tsinfer. Population genetics theory predicts that IBD
+% (Identical By Descent) segments break down over time, resulting in shorter
+% ancestral segments further back in the history of the samples. At time,
+% \textit{y \textgreater{0.8}}, the plot reveals a pattern of node spans contrary
+% to this expectation. It confirms a known algorithmic artefact due to which
+% disproportionately long old ancestors are generated.
+
+% Supplementary Fig. \ref{fig:Supplementary_Figure_4}, a screenshot of the
+% Mutations page for SARS-CoV-2 ARGs \citep{zhan2023towards} demonstrates
+% tsbrowse's ability to handle diverse data sources. This plot represents a total
+% of \textcolor{red}{X million} mutations from an inference of
+% \textcolor{red}{1.27 million} sequences. After executing the
+% \texttt{preprocess} step, the page loads in \textcolor{red}{X} seconds on a
+% system with \textcolor{red}{Y GB RAM}, showing that it is possible to visualise
+% large-scale datasets in a performant way. Panel B demonstrates a useful
+% interactive feature; additional information about a mutation event (for
+% example, the mutational frequency in predefined populations, or the number of
+% sample nodes carrying the mutation) is displayed on mouse-over of individual
+% data points.
 

### 2025-03-27 manuscript/paper.tex
Update to discussion

@@
 Fig~\ref{fig:Supplementary_Figure_3} shows the genomic spans of 
-ancestral nodes in an ARG inferred with tsinfer, demonstrating a clear 
+ancestral nodes in an ARG inferred with \texttt{tsinfer}, demonstrating a clear 
 excess of long haplotypes in the very ancient past. These insights 
@@
 
-\section{Discussion} \label{sec:Discussion} Visualisation techniques are
-essential for data-driven insights. Specialised tools such as \textit{Treenome
-browser} \citep{Kramer2023} tackle the challenge of semantic visualisation of
-large phylogenies. Yet capturing the data model underlying ARGs is crucial for
-the systematic exploration of local genealogies. \texttt{Tsbrowse} addresses
-this open problem by condensing ARG properties into a set of browsable,
-interactive visual summaries, thus aiding in ARG-based data exploration and
-hypothesis generation.
+\section{Discussion} \label{sec:Discussion} 
+Visualisation of tree topology is a central task in phylogenetics, 
+and although numerous tools
+exist~\citep[e.g.][]{huson2007dendroscope,vaughan2017icytree},
+the methods can typically only handle a few hundred nodes.
+Visualisation of large-scale tree topologies with millions of nodes
+requires much more sophisticated approaches
+to capture topological features at different scales, 
+and is active research area~\citep{wong2022dynamic,kramer2023treenome}.
+Adapting such methods, and integrating into \texttt{tsbrowse} to provide
+a local tree viewer that operates at the million-node scale
+is an important direction for future work.
+An interactive viewer for the entire ARG topology, capturing 
+semantic properties of the graph at a range of scales, is an even more 
+ambitious goal, and would be a major asset for the field.
 

### 2025-03-27 manuscript/paper.tex

@@
 
-\section*{Acknowledgments}
-\section*{Author contributions}
-\section*{Supplementary Data}
-Supplementary data are available at \textcolor{red}{Bioinformatics online}.
-\section*{Conflict of interest}
-No competing interest is declared.
+% \section*{Acknowledgments}
+% \section*{Author contributions}
+% \section*{Supplementary Data}
+% Supplementary data are available at \textcolor{red}{Bioinformatics online}.
+% \section*{Conflict of interest}
+% No competing interest is declared.
+
 \section*{Funding}
@@
 Sciences Research Council (UKRI-BBSRC) [grant number BB/T008784/1],
-\textcolor{red}{add Novo funding and funding details for Ben, Duncan, Jerome}.
+\textcolor{red}{add Novo funding and funding details for Duncan}.
+JK acknowledges EPSRC (research grant EP/X024881/1),
+NIH (research grants HG011395 and HG012473)
+and the Robertson Foundation.
+
 \bibliographystyle{natbib}

### 2025-03-27 manuscript/paper.tex
Add funding info for Duncan
Update sc2 details
Closes #14

@@
 Fig~\ref{fig:Supplementary_Figure_4}, for example, shows a screenshot of 
-\texttt{tsbrowse} summarising almost 1 million mutations along the SARS-CoV-2
-genome.
+\texttt{tsbrowse} summarising the 1.9 million mutations in a SARS-CoV-2
+ARG with around 2.7 million nodes and edges.
 
@@
 Sciences Research Council (UKRI-BBSRC) [grant number BB/T008784/1],
-\textcolor{red}{add Novo funding and funding details for Duncan}.
+\textcolor{red}{add Novo funding and funding details for Duncan}. DMR is funded by 
+a studentship from the Wellcome Programme in Genomic Medicine and Statistics.
 JK acknowledges EPSRC (research grant EP/X024881/1),
@@
 
-\subsection{Inference of SARS-Cov-2 ARGs}
+\subsection{Inference of SARS-CoV-2 ARGs}
+The ARG shown in Fig \ref{fig:Supplementary_Figure_4} was inferred with
+\texttt{sc2ts}~\citep{zhang2023biobank} using the Viridian
+dataset~\citep{hunt2024addressing}. It consists of 
+2,482,157 samples
+2,689,054 nodes,
+2,689,982 edges
+and
+1,923,169 mutations.
+% Trees   348
+% Sequence Length 29904.0
+% Time Units  days
+% Sample Nodes    2482157
+% Total Size  1.5 GiB
+% Table   Rows    Size    Has Metadata
+% Edges   2689982 82.1 MiB
+% Individuals 0   24 Bytes
+% Migrations  0   8 Bytes
+% Mutations   1923169 202.7 MiB
+% Nodes   2689054 1.1 GiB
+% Populations 0   8 Bytes
+% Provenances 1131    1.7 MiB
+% Sites   29803   3.1 MiB
+Running \texttt{tsbrowse preprocess} on the input \texttt{tszip} file (113M)
+required 2m19s time elapsed (15m15s CPU time) on an Intel Core(TM) i7-9700 CPU.
+The resulting \texttt{.tsbrowse} file size was 130M.
+% Preprocessing completed. You can now view with `python -m tsbrowse serve v1-beta1_2023-02-21.pp.md.il.ts.tsbrowse`
 
-The ARG shown in \ref{fig:Supplementary_Figure_4} was inferred with
-\texttt{sc2ts}~\citep{zhang2023biobank} using the Viridian
-dataset~\citep{hunt2024addressing}. It consists of 516,443 samples, 612,209
-nodes, 612,655 edges and 988,303 mutations.
+% real    2m19.028s
+% user    15m15.049s
+% sys     0m8.159s
+% ```
+% Sizes:
+% ```
+% $ ls -lh v1-beta1_2023-02-21.pp.md.il.ts.ts*
+% -rw-rw-r-- 1 jk jk 130M Mar 27 15:08 v1-beta1_2023-02-21.pp.md.il.ts.tsbrowse
+% -rw-r--r-- 1 jk jk 113M Mar 27 15:04 v1-beta1_2023-02-21.pp.md.il.ts.tsz
+% ```
 
@@
     \includegraphics[width=0.95\linewidth]{figures/SuppFig4.png}
-    \caption{\textbf{\texttt{tsbrowse} applied to SARS-CoV2 ARGs.}
-    A screenshot of \texttt{tsbrowse}'s depiction of 988,303 mutations in an 
+    \caption{\textbf{\texttt{tsbrowse} applied to SARS-CoV-2 ARGs.}
+    A screenshot of \texttt{tsbrowse}'s depiction of 
+    1,923,169 mutations in an
     ARG inferred by \texttt{sc2ts}; see  text for details. Also shown are the 

### 2025-03-28 manuscript/paper.tex
fixed labels for main figure 2 fixed figure caption fixed typo in sweep simulation
Closes #6

@@
 
-\address[1]{\orgdiv{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery}, \orgname{University of Oxford}, \orgaddress{\street{Old Road Campus}, \postcode{Oxford OX3 7LF}, \country{United Kingdom}}}
+\address[1]{\orgdiv{Big Data Institute, Li Ka Shing Centre for Health
+        Information and Discovery}, \orgname{University of Oxford},
+    \orgaddress{\street{Old Road Campus}, \postcode{Oxford OX3 7LF},
+        \country{United Kingdom}}}
 
-\corresp[$\ast$]{Corresponding author. \href{email:savita.karthikeyan@st-hughs.ox.ac.uk}{savita.karthikeyan@st-hughs.ox.ac.uk}}
+\corresp[$\ast$]{Corresponding author.
+    \href{email:savita.karthikeyan@st-hughs.ox.ac.uk}{savita.karthikeyan@st-hughs.ox.ac.uk}}
 
@@
 Ancestral recombination graphs (ARGs) describe how a set of sample
-sequences relate to each other at each position along the genome in a recombining
+sequences relate to each other at each position along the genome in a
+recombining
 species, and are currently the subject of intense research
@@
 such as inferring selection~\citep{stern2019approximate,hejase2022deep} and
-the spatial 
-location of genetic
-ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2024geographic},
+the spatial location of genetic ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2024geographic},
 more powerful approaches to quantifying genetic
@@
 While these developements are exciting, the performance 
-of these new methods depend critically on the accuracy of the inferred ARGs,
-and although studies benchmarking the various inference methods on 
+of these new methods depends critically on the accuracy of the inferred ARGs.
+Although studies benchmarking the various inference methods on 
 simulated data have emerged 
@@
 the practicalities of applying ARG inference to real data is 
-understudied. In particular, there is a critial lack of software 
+understudied. In particular, there is a critical lack of software 
 infrastructure to support evaluation and quality control of 
@@
 quality control
-of assembly~\citep{wick2015bandage,challis2020blobtoolkit},
+of genome assembly~\citep{wick2015bandage,challis2020blobtoolkit},
 read mapping~\citep{robinson2011integrative},
@@
 \texttt{Tsbrowse} uses the ``succinct tree sequence'' encoding of 
-ARGS~\citep{wong2024general}. This efficient ARG encoding 
+ARGs~\citep{wong2024general}. This efficient ARG encoding 
 is implemented by the 
@@
 nodes and edges that describe the graph topology and the sites
-and mutations encoding the sequence variation (Fig.~\ref{fig:Figure_1}).
+and mutations that encode the sequence variation (Fig.~\ref{fig:Figure_1}).
 % For example, the Node table describes sampled or ancestral sequences
@@
 at the requested resolution on the server, sends the image to the web-browser
-for display, and dynamically updates as the user interatively navigates the ARG.
+for display, and dynamically updates as the user interactively navigates the
+ARG.
 This approach allows us to summarise very large ARGs interactively;
@@
         The effect of the sweep can be seen by the lack of edges crossing the
-centre of the simulated chromosome. Other tabs are accessed by the 
-buttons on the top bar.
+        centre of the simulated chromosome due to an excess of recent coalescent
+        events. Moving away from the focal site, the oldest edges are the
+        first to rejoin the ARG, followed by more recent edges, resulting in a
+        wedge-like pattern of missing edges in the centre.
         \label{fig:Figure_2}}
@@
 The plot on the left is an overview of the 33,929 edges
-(parent-child relationships between nodes along the genome)
 in a simulated ARG with a strong selective sweep (see
 Sec.~\ref{subsec:sweep_simulation} for details).
-Each edge is depicted as a horizontal line connecting the genomic coordinates of the
+Each edge is depicted as a horizontal line connecting the genomic coordinates
+of the
 parent and child nodes on the x-axis, and the y-axis shows the time of
@@
 The purpose of \texttt{tsbrowse} is to provide an interactive view of the
-\texttt{tskit} ARG data model to guide intuition, improve inference quality control
-and facilitate debugging. An example of how we can deepen our understanding
+\texttt{tskit} ARG data model to guide intuition, improve inference quality
+control and facilitate debugging. An example of how we can deepen our understanding
 using the genome-browser-like perspective provided by \texttt{tsbrowse}
-is provided the simulated ARG of Fig~\ref{fig:Figure_2}, where 
+is given in the simulated ARG of Fig~\ref{fig:Figure_2}, where
 the gap in the edges view corresponds to the characteristic dip in diversity
@@
 developed as a way to quickly visualise the effects of such filtering
-choices on ARGs inferred by \texttt{tsinfer}, and it is now an indispensible
+choices on ARGs inferred by \texttt{tsinfer}, and it is now an indispensable
 element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_2} 
-shows [a before and after view for 1000G data, where tsbrowse helped 
-diagnose and resolve an opaque data problem].
+shows a region of the 1000G data with gaps in site density that are spanned by
+ exceptionally long edges, which are likely to bias downstream statistics.
 These interactive visualisations have also helped diagnose issues with
@@
 to capture topological features at different scales,
-and is active research area~\citep{wong2022dynamic,kramer2023treenome}.
+and is an active research area~\citep{wong2022dynamic,kramer2023treenome}.
 Adapting such methods, and integrating into \texttt{tsbrowse} to provide
@@
 Sciences Research Council (UKRI-BBSRC) [grant number BB/T008784/1],
-\textcolor{red}{add Novo funding and funding details for Duncan}. DMR is funded by 
+\textcolor{red}{add Novo funding}. DMR is funded by 
 a studentship from the Wellcome Programme in Genomic Medicine and Statistics.
@@
 
-\clearpage \onecolumn \setcounter{figure}{0}  % Reset the figure counter
+\clearpage \onecolumn \setcounter{figure}{0}
+% Reset the figure counter
 \renewcommand{\thefigure}{S\arabic{figure}}
@@
 \section{Supplementary Material}
-\subsection{Simulation of truth dataset (Fig. \ref{fig:Figure_2})} \label{subsec:sweep_simulation} 
+\subsection{Simulation of truth dataset (Fig. \ref{fig:Figure_2})}
+\label{subsec:sweep_simulation}
 Ancestral
@@
 \textit{dt = 1e-6} was used to step through the sweep. Mutations were added to
-the ARG at a rate of 1e-6. For simulating history before occurrence of the
+the ARG at a rate of 1e-8 per base pair per generation. A recombination rate of 1e-8 per base pair per generation. For simulating history before occurrence of the
 sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was
@@
 \citep{kelleher2019inferring}, \texttt{tsdate version 0.2.1}, 
-\texttt{Relate version 1.2.2} \cite{speidel2019method}, 
+\texttt{Relate version 1.2.2} \citep{speidel2019method}, 
 \texttt{ARG-needle version 1.0.3} \citep{zhang2023biobank},
@@
 \ref{fig:Supplementary_Figure_3})} The 1000 Genomes dataset was downloaded from
-\textcolor{red}{[insert link]}. Ancestral fasta sequence for chromosome 20
-(GRCh38) was downloaded from the Ensembl database. Inference was performed with
-a Snakemake pipeline \textcolor{red}{(cite Ben's repo?)} using \texttt{tsinfer
-version 0.3.3} for the long arm of chromosome 20 after filtering out duplicate
-variant positions, variants with missing or low quality ancestral allele,
-singletons, n-1-tons and n-2-tons. Only bi-allelic SNPs were included for
-inference. For the inference in Suppl. Fig. 2A, all genomic regions were
-included. For the inference in Suppl. Figs 2B and 3, genomic regions with a
-site density lesser than 5 per kb were discarded. For Suppl. Fig. 2,
-\texttt{tsdate version 0.2.1} was used to estimate the age of ancestral nodes
-with \textit{mutation rate = 1.29e-8}, setting all other parameters to default
-values.
+\texttt{\url{https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/working/20220422_3202_phased_SNV_INDEL_SV/}}.
+Ancestral fasta sequences for chromosomes 17 and 20 (GRCh38) were downloaded
+from the Ensembl database. Inference was performed with a Snakemake pipeline
+(\texttt{\url{https://github.com/benjeffery/tsinfer-snakemake/}}) using
+\texttt{tsinfer version 0.3.3} for the long arms of chromosome 17 and 20 after
+filtering out duplicate variant positions, variants with missing or low quality
+ancestral allele, singletons, n-1-tons and n-2-tons. Only bi-allelic SNPs were
+included for inference. For the inference in Suppl. Fig.
+\ref{fig:Supplementary_Figure_2}, all genomic regions were included. For the
+inference in Suppl. Fig. \ref{fig:Supplementary_Figure_3}, genomic regions with
+a site density lesser than 5 per kb were discarded. For Suppl. Fig.
+\ref{fig:Supplementary_Figure_2}, \texttt{tsdate version 0.2.1} was used to
+estimate the age of ancestral nodes with \textit{mutation rate = 1.29e-8},
+setting all other parameters to default values.
 
@@
     \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
-    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate (A),
-ARG-Needle (B), Relate (C) and SINGER (4) inferences of the truth dataset
+    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate,
+ARG-Needle, Relate and SINGER inferences of the truth dataset
 simulated under a selective sweep model (in Figure 2 of the main text). For
@@
     \centering
-    % \includegraphics[width=0.95\linewidth]{figures/SuppFig2.png}
-    \caption{\textbf{A view of mutations in the 1000G WGS dataset.} A
-screenshot of \texttt{tsbrowse}'s Mutations view for a X Mb region on
-chromosome 17 inferred from from 3,202 participants from the 1000 Genomes Whole
-Genome Sequencing dataset (1000G) \citep{1000G2015, 1000GWGS}. The genomic
-position of the mutations is shown on the X axis, and the age of mutation is
-represented on the Y axis. Panel B represents an inference after filtering out
-low site-density regions in the samples.}
+    \includegraphics[width=0.95\linewidth]{figures/SuppFig2.png}
+    \caption{\textbf{Identifying ARG inference problems with \texttt{tsbrowse}} 
+Screenshots of \texttt{tsbrowse}'s Mutations view (A) and Edges view (B) for a
+ 600 kB region of chromosome 17 inferred from 3,202 participants from the 1000 
+ Genomes Whole Genome Sequencing dataset (1000G) \citep{1000G2015, 1000GWGS}. 
+ The poor performance of \texttt{tsinfer} in this variant-poor region is evidenced
+  by the long edges spanning gaps in mutation density.}
     \label{fig:Supplementary_Figure_2}

### 2025-04-09 manuscript/paper.tex
readthrough for typos; incorporated Ana's explanation for Fig 2 caption

--- 2025-03-27 manuscript/paper.tex
@@
 element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_2}
-shows [a before and after view for 1000G data, where tsbrowse helped 
-diagnose and resolve an opaque data problem].
+shows [a before and after view for 1000G data, where tsbrowse helped diagnose 
+and resolve an opaque data problem].
 These interactive visualisations have also helped diagnose issues with
@@
 Sciences Research Council (UKRI-BBSRC) [grant number BB/T008784/1],
-\textcolor{red}{add Novo funding and funding details for Duncan}. DMR is funded by 
+\textcolor{red}{add Novo funding and funding details for Duncan}. DMR is funded
+by
 a studentship from the Wellcome Programme in Genomic Medicine and Statistics.

### 2025-04-09 manuscript/paper.tex
Add SuppFig2 and fix some typos

--- 2025-03-28 manuscript/paper.tex
@@
 \textit{dt = 1e-6} was used to step through the sweep. Mutations were added to
-the ARG at a rate of 1e-8 per base pair per generation. A recombination rate of 1e-8 per base pair per generation. For simulating history before occurrence of the
+the ARG at a rate of 1e-6. For simulating history before occurrence of the
 sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was
@@
 ancestral allele, singletons, n-1-tons and n-2-tons. Only bi-allelic SNPs were
-included for inference. For the inference in Suppl. Fig.
-\ref{fig:Supplementary_Figure_2}, all genomic regions were included. For the
-inference in Suppl. Fig. \ref{fig:Supplementary_Figure_3}, genomic regions with
-a site density lesser than 5 per kb were discarded. For Suppl. Fig.
-\ref{fig:Supplementary_Figure_2}, \texttt{tsdate version 0.2.1} was used to
-estimate the age of ancestral nodes with \textit{mutation rate = 1.29e-8},
-setting all other parameters to default values.
+included for inference. For the inference in Suppl. Fig. 2, all genomic regions
+were included. For the inference in Suppl. Fig. 3, genomic regions with a site
+density lesser than 5 per kb were discarded. For Suppl. Fig. 2, \texttt{tsdate
+version 0.2.1} was used to estimate the age of ancestral nodes with
+\textit{mutation rate = 1.29e-8}, setting all other parameters to default
+values.
 
@@
     \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
-    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate,
-ARG-Needle, Relate and SINGER inferences of the truth dataset
+    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate (A),
+ARG-Needle (B), Relate (C) and SINGER (D) inferences of the truth dataset
 simulated under a selective sweep model (in Figure 2 of the main text). For

### 2025-04-09 manuscript/paper.tex
Fix hardcoded refs

@@
 ancestral allele, singletons, n-1-tons and n-2-tons. Only bi-allelic SNPs were
-included for inference. For the inference in Suppl. Fig. 2, all genomic regions
-were included. For the inference in Suppl. Fig. 3, genomic regions with a site
-density lesser than 5 per kb were discarded. For Suppl. Fig. 2, \texttt{tsdate
-version 0.2.1} was used to estimate the age of ancestral nodes with
-\textit{mutation rate = 1.29e-8}, setting all other parameters to default
-values.
+included for inference. For the inference in Suppl. Fig.
+\ref{fig:Supplementary_Figure_2}, all genomic regions were included. For the
+inference in Suppl. Fig. \ref{fig:Supplementary_Figure_3}, genomic regions with
+a site density lesser than 5 per kb were discarded. For Suppl. Fig.
+\ref{fig:Supplementary_Figure_2}, \texttt{tsdate version 0.2.1} was used to
+estimate the age of ancestral nodes with \textit{mutation rate = 1.29e-8},
+setting all other parameters to default values.
 

### 2025-04-18 manuscript/cover_letter.tex

\documentclass{letter}

\signature{Savita Karthikeyan}

\address{Big Data Institute\\University of Oxford\\UK}
\begin{document}

\begin{letter}{Bioinformatics}

    \opening{Dear Editors,}

    I write on behalf of my co-authors to submit our original research work
    entitled
    \emph{'tsbrowse: an interactive browser for Ancestral Recombination
        Graphs'} to
    Bioinformatics,
    and request that it be considered for publication as an Application Note.

    The manuscript describes an open-source Python web-app to visualise
    fundamental
    entities
    of Ancestral Recombination Graphs (ARGs). ARGs are a topic of intense
    research
    interest and
    have revolutionised the field of genomics following recent technological
    advancements
    which have made their inference from large-scale empirical datasets
    possible: see Grundler et al. (Science, 2025), Speidel et al. (Nature,
    2025)
    for recent ARG-based findings, and Nielsen et al. (Nature Reviews Genetics,
    2024) for a recent review of ARG inference.

    Visual inspection is a crucial part of data analysis pipelines. For omics,
    this
    vital
    infrastructure is provided by stat-of-art genome browsers. The lack of
    tools
    for the
    visual inspection of ARGs is a major gap in a rapidly advancing field. With
    tsbrowse,
    we aim to bridge this gap by presenting interactive views of the nodes,
    edges
    and mutations
    in an ARG. The manuscript describes the applications and architecture of
    tsbrowse, and
    demonstrates user-interactivity and scalability to millions of samples.

    We believe this work directly aligns with the scope of Bioinformatics, and
    that
    its
    readers will find tsbrowse of particular interest. As we show in the
    manuscript, visualising
    the information structure in ARGs greatly helps improve inference
    pipelines,
    making tsbrowse
    particularly useful for modern, large-scale genomic datatsets. We hope
    that, if
    published
    tsbrowse will drive further advancements in the areas of ARG visualisation
    and
    validation,
    and make these steps a routine part of ARG inference.

    \closing{Sincerely,}

\end{letter}
\end{document}

### 2025-04-22 manuscript/cover_letter.tex
Update manuscript/cover_letter.tex
Co-authored-by: Jerome Kelleher <jk@well.ox.ac.uk>

@@
     research
-    interest and
-    have revolutionised the field of genomics following recent technological
+    interest in the field of genomics, following recent technological
     advancements

### 2025-04-22 manuscript/cover_letter.tex
Update manuscript/cover_letter.tex
Co-authored-by: Jerome Kelleher <jk@well.ox.ac.uk>

@@
     vital
-    infrastructure is provided by stat-of-art genome browsers. The lack of
+    infrastructure is provided by state-of-art genome browsers. The lack of
     tools
@@
     tsbrowse,
-    we aim to bridge this gap by presenting interactive views of the nodes,
+    we bridge this gap by presenting interactive views of the nodes,
     edges

### 2025-04-22 manuscript/paper.tex

@@
 \DOI{DOI HERE}
-\copyrightyear{2024}
+\copyrightyear{2025}
 % \pubyear{2019}
@@
 
-\title[tsbrowse]{Tsbrowse: a genome browser for Ancestral Recombination Graphs}
+\title[tsbrowse]{Tsbrowse: an interactive browser for Ancestral Recombination Graphs}
 
@@
 
-\keywords{Ancestral Recombination Graphs, Holoviz, biobank-scale ARG
-    visualisation, genome browser} \maketitle
+\keywords{Ancestral Recombination Graph, HoloViz, data visualisation, interactive browser} \maketitle
 
@@
 \section*{Funding}
-This work was supported by funding from the Biotechnology and Biological
-Sciences Research Council (UKRI-BBSRC) [grant number BB/T008784/1],
-\textcolor{red}{add Novo funding}. DMR is funded by 
+SK acknowledges full support and funding from Novo Nordisk Research Centre Oxford,
+and funding from the Biotechnology and Biological Sciences Research Council 
+(UKRI-BBSRC) [grant number BB/T008784/1]. DMR is funded by 
 a studentship from the Wellcome Programme in Genomic Medicine and Statistics.
@@
     \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
-    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate (A),
-ARG-Needle (B), Relate (C) and SINGER (D) inferences of the truth dataset
+    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate,
+ARG-Needle, Relate and SINGER inferences of the truth dataset
 simulated under a selective sweep model (in Figure 2 of the main text). For

### 2025-04-22 manuscript/paper.tex

@@
 \texttt{\url{https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/working/20220422_3202_phased_SNV_INDEL_SV/}}.
-Ancestral fasta sequences for chromosomes 17 and 20 (GRCh38) were downloaded
+The ancestral fasta sequence for chromosome 17 (GRCh38) was downloaded
 from the Ensembl database. Inference was performed with a Snakemake pipeline
 (\texttt{\url{https://github.com/benjeffery/tsinfer-snakemake/}}) using
-\texttt{tsinfer version 0.3.3} for the long arms of chromosome 17 and 20 after
+\texttt{tsinfer version 0.3.3} for the long arm of chromosome 17 after
 filtering out duplicate variant positions, variants with missing or low quality
 ancestral allele, singletons, n-1-tons and n-2-tons. Only bi-allelic SNPs were
-included for inference. For the inference in Suppl. Fig.
-\ref{fig:Supplementary_Figure_2}, all genomic regions were included. For the
-inference in Suppl. Fig. \ref{fig:Supplementary_Figure_3}, genomic regions with
-a site density lesser than 5 per kb were discarded. For Suppl. Fig.
-\ref{fig:Supplementary_Figure_2}, \texttt{tsdate version 0.2.1} was used to
-estimate the age of ancestral nodes with \textit{mutation rate = 1.29e-8},
-setting all other parameters to default values.
+included for inference. For Suppl. Fig. \ref{fig:Supplementary_Figure_2}, 
+\texttt{tsdate version 0.2.1} was used to estimate the age of ancestral nodes
+ with \textit{mutation rate = 1.29e-8}, setting all other parameters to default 
+ values.
 
@@
         screenshot of \texttt{tsbrowse}'s Nodes view for an inference of the long arm
-        of chromosome 20 from the 1000 Genomes whole-genome sequencing dataset. At the
+        of chromosome 17 from the 1000 Genomes whole-genome sequencing dataset. At the
         top is a plot of node spans over time. The length of sequence that the nodes

### 2025-04-28 manuscript/paper.tex
added ORCID details; fixed suppl. fig 1 floating; final readthrough;

@@
 
-\author[1$\ast$]{Savita Karthikeyan}
-\author[1]{Ben Jeffery}
-\author[1]{Duncan Mbuli-Robertson}
-\author[1]{Jerome Kelleher \ORCID{0000-0000-0000-0000}}
+\author[1,$\ast$]{Savita Karthikeyan \ORCID{0000-0002-4798-5746}}
+\author[1]{Ben Jeffery \ORCID{/0000-0002-1982-6801}}
+\author[1]{Duncan Mbuli-Robertson \ORCID{0000-0002-1660-2415}}
+\author[1]{Jerome Kelleher \ORCID{0000-0002-7894-5253}}
 
@@
 such as inferring selection~\citep{stern2019approximate,hejase2022deep} and
-the spatial location of genetic ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2024geographic},
+the spatial location of genetic ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2025geographic},
 more powerful approaches to quantifying genetic
@@
 \citep{brandt2022evaluation,deng2024robust,peng2024evaluating},
-the practicalities of applying ARG inference to real data is 
+the practicalities of applying ARG inference to real data are 
 understudied. In particular, there is a critical lack of software 
@@
 
-\section{Supplementary Material}
+\section{Supplementary Methods}
 \subsection{Simulation of truth dataset (Fig. \ref{fig:Figure_2})}
@@
 \section{Supplementary Figures}
-\begin{figure}
+\begin{figure}[H]
     \centering

### 2025-06-10 manuscript/revision/response_to_editor.tex
minor revisions and response to reviewers

\documentclass{letter}
\usepackage{xcolor}
\usepackage{hyperref}
\signature{Savita Karthikeyan}

\address{Big Data Institute\\University of Oxford\\UK}
\begin{document}

\begin{letter}{}

\opening{Dear Alison,}

We are delighted that the reviewers have found our manuscript suitable for publication 
in Bionformatics. We thank you, the editors and all the reviewers for the careful 
consideration of our manuscript, and value the kind words and constructive feedback 
provided on our work. We have now revised the manuscript based on the reviewers' 
suggestions. Below is our point-by-point response to the 
reviewers' comments in blue. We have incorporated as many suggestions as possible, 
and where unable to do so we have indicated the reason.\\

\hrule

Reviewer: 1

Comments to the Author \\

Summary: In this manuscript, the authors present tsbrowse, an open-source Python-based 
application for the interactive visualization of the components of Ancestral Recombination 
Graphs (ARGs). Because ARGs (in principle) contain all knowable information concerning 
coalescence and recombination in the history of a set of samples, there is great excitement 
in developing methods for inferring ARGs from empirical genomic sequence data, as well as 
for learning about evolutionary processes from patterns in inferred ARGs. A crucial component 
of both these endeavors is visualizing ARGs, which is difficult or unwieldy. Tsbrowse therefore 
represents an exciting contribution to the field which is likely to be of high utility and 
to be highly cited.

Opinion: This paper was straightforward, clear, and well written. I think it does a good job 
setting up the intellectual context for ARG-based research and advocating for the utility of 
tsbrowse. I had two minor, nitpicky comments, but overall, it was great.  Thanks for a fun and 
easy review!

1. I think the order in which the supplemental figures are cited doesn’t match their numbering 
(i.e., S4 is cited before S1)\\
\textcolor{blue}{Thanks to Reviewer 1 for pointing out this error. We have now rearranged the 
supplementary figures so that the citations appear in ascending order.}

2. In the discussion, the authors mention that methods for visualization of tree topology can 
typically only handle a few hundred nodes, but it might be worth doing a “(but see…)” for some 
of the approaches folks have taken for visualizing much larger trees (e.g., nextstrain, onezoom).\\
\textcolor{blue}{We agree with the reviewer-  it is important to acknowledge previous/on-going visualisation efforts 
for larger trees. We had already included a citation to the OneZoom project (and the Treenome browser) in the 
sentence that followed the one in question "… Visualisation of large-scale tree topologies with millions of nodes 
requires much more sophisticated approaches to capture topological features at different scales, 
and is an active research area". We are thankful to the reviewer for bringing Nextstrain to our 
attention, and have now included Hadfield et al. as a citation.}

Reviewer: 2\\

Comments to the Author\\

I have only a few minor comments:\\
1. "and storing as a compressed .tsbrowse file. (The .tsbrowse file is also a valid input for
tszip, a general utility for compressing tskit files.)" Sorry, but I don't understand this sentence. 
More precisely, both the compressed and the uncompressed files have the .tbrowse extension? If yes, 
that's a weird choice and I would like to see some justification. You should also make clear if tszip 
is a part of tskit or of twbrowse.\\
\textcolor{blue}{To clarify, the tszip is part of the tskit ecosystem. It is not part of the tsbrowse 
package but can be used to compress the tables output from the tsbrowse preprocess step. Thanks to 
the reviewer for this pointing out this potentially unclear statement. We have now deleted the text 
in brackets, which we believe isn't crucial for understanding the data model.}

2. "web-browser" - web browser\\
\textcolor{blue}{Fixed, thanks for pointing out the typographical error.}

3. "Fig S4, for example, shows a screenshot of tsbrowse". This is a matter of personal taste, but I do
not like to refer to a Figure in the supplementary materia just as if it were in the article; if the 
Figure is relevant, it should be in the article. If it's not relevant, that you should refer to it without
hinting that it is important.\\
\textcolor{blue}{We agree with the reviewer's opinion, and believe that Figure S4 is crucial 
in order to demonstrate the application of tsbrowse to large-scale datasets (which is a key design 
feature of the app). However, we are unable to move this figure to the main matter considering journal 
guidelines for page limit.}

4. Looking at the github repo, I can find the installation instruction using pip only. In my experience 
pip is quite picky. I strongly suggest you to provide tsbrowse via conda-forge/bio-conda or a containerized 
solution (e.g. docker).\\
\textcolor{blue}{Thanks to the reviewer for the suggestion to provide alternate installation solutions. 
We have now created a Docker container in order to ensure a smooth experience across computing environments. 
Links have been updated in the manuscript and app documentation}
\textcolor{red}{[TO DO; after release of new package]}

Reviewer: 3

Comments to the Author
This paper offers a browser-based visualisation tool for ancestral recombination graphs (ARGs), which is 
a data structure representing gene genealogies for a sample of DNA sequences. The complication in the p
resence of recombination is that these structures are not simple trees, but - because of the possibility 
of recombination - graphs with reticulated nodes, which makes exploratory data analysis difficult.

I agree with the authors that there is currently a lack of visualisation tools for ARGs, and I expect the 
current paper to provide a useful addition to the catalog of tools for ARG simulation and inference. Tsbrowse 
integrates with the widely used tskit ecosystem of software tools, developed by some of the current authors, 
which will facilitate a rapid uptake. The authors make a compelling argument for the need for visualisation 
tools - Fig S1 in particular shows that four different and popular tools for ARG inference are giving 
qualitatively completely different insights into their reconstructed ARGs (incidentally, it's a shame that 
that figure couldn't make it into the main paper, though I recognise that this journal has strict page limits).\\
\textcolor{blue}{We appreciate the reviewer's feedback on the utility of tsbrowse, and acknowledge the 
importance of S1 in highlighting the need for visualisation. However, considering the necessity of the 
existing main display items to establish the fundamental aspects of tsbrowse, we have decided to retain S1 
in the supplementary. We agree that S1 would have been a compelling main display item had the page/figure 
limits of the journal allowed for it.}

My main comment is that tsbrowse does not actually seem to be an ARG visualisation tool. It is a useful way 
to visualise mutations, tree-sequence edges, and summary statistics, but nowhere will you actually see a tree 
(much less an ARG). On rereading the paper I see that the manuscript has been carefully worded not to include 
such a claim, but a casual reader might expect tsbrowse to be able to do so, as other tools such as IcyTree can. 
I do get fed up with reviewers tossing additional requests on top of an already polished piece of work, but 
in this case I think it would be a helpful addition! - would it really be too much work for tsbrowse to be 
able to draw at least the local trees? (For researchers with smaller datasets, being able to interrogate the 
fine details of an ARG would be equally valuable as the complementary selling-points around scalability.) At 
the very least I would like to see the manuscript at least be a little more up-front about this omission.\\

\textcolor{blue}{We thank the reviewer for making this important observation, and agree that drawing 
(parts of) local trees would be a great feature. For smaller datasets, tools such as arg-visualiser 
(\url{https://pypi.org/project/tskit-arg-visualizer/}) and existing tskit infrastructure, for example 
the \texttt{draw\_text()} and \texttt{draw\_svg()} already provide this functionality. A key 
motivation for this work is to guide large-scale inference, and implementing tree visualisation that is interpretable 
at this scale would require significant thought towards design as well as considerable effort. We therefore 
think that implementing this feature is not feasible for this work and is better scoped as future work or as 
a separate endeavour. In the text, we do state that tsbrowse is a visualiser for the information structure 
in ARGs, which we hope implies that it is a visualiser for the properties of the ARG, rather than the ARG 
topology itself. We have also already explicitly stated in the Discussion that drawing local trees is an 
important future direction.}\\

Minor comments:\\

Manuscript p1: 'it is only with recent breakthroughs in inference methods...that practical application 
has become possible'. I appreciate the need to emphasise the recent increase in interest in inference for ARGs, 
but I think this is selling older methods a little short. There are ARG inference methods going back at least 
as far as Griffiths and Marjoram (1996) 'Ancestral Inference from Samples of DNA Sequences' JCB 3(4):479-502. 
True they are not scalable in the manner of recent methods, but they *are* practical applications, and they have 
the advantage that they actually sample from the claimed model without introducing any model heuristics.
\textcolor{blue}{Thanks to the reviewer for their feedback on this statement on early methods. We agree, 
and have revised the text to acknowledge the contributions of foundational work in ARG inference.}

Manuscript p1: Typo 'developements'.\\
\textcolor{blue}{Thanks, this is now fixed.}

Manuscript p2: Where you mention 'Bokeh', can you clarify what this is (e.g. with a citation)?\\
\textcolor{blue}{Yes, Bokeh is a Python library for interactive visualisation and a citation to it 
has been added.}
\textcolor{red}{TO FIX! Citation appears as a question mark!!.}

Supplement p1, Sec 1.4: "The ARG shown in Fig 4" should say "The ARG shown in Suppl. Fig 4" ?\\
\textcolor{blue}{Thanks, this is now fixed.}\\
\textcolor{red}{(TO DO: make sure to replace links in suppl pdf!)}

Fig S2 caption - 'Consortium (2015)' is citing incorrectly.\\
\textcolor{blue}{Fixed (removed this citation as it was not required)}

From the online documentation, Node page: \url{https://tskit.dev/tsbrowse/docs/stable/nodes.html}
"The histograms at the bottom show the distributions of node spans over different dimensions. 
The leftmost histogram summarises the span of nodes on the sequence; the middle plot summarises 
the span of nodes over time and the rightmost plot summarises the edge “area” defined as the product 
of sequence span and time span for each node."

I didn't understand this. The 'nodes' section of the visualiser only seems to provide one histogram, 
not three.\\

\textcolor{blue}{Thanks to the reviewer for pointing out this documentation error. The Nodes section 
does indeed have one histogram and the documentation has been updated to reflect this.}

Installation notes:

I had a few problems getting tsbrowse to work. There are two different sets of installation instructions:

1. https://pypi.org/project/tsbrowse/
2. https://tskit.dev/tsbrowse/docs/stable/contributing.html

I use a Mac+Homebrew+Chrome and Option 2 seems more fitting. ("python -m pip install tsbrowse" as in 
option 1 will fail because Homebrew's Python environment is externally managed.)

\textcolor{blue}{We apologise for any confusion on installation instructions. The recommended way to install 
tsbrowse is via pip within a virtual environment. We have fixed the documentation to show only one set of 
instructions (i.e, for the user) at the top.}

Bug(?):

I frequently had problems rendering the window in the 'Mutations' tab, which tended to get stuck at the 
circling wheel. When clicking the 'Mutations' button, my terminal returned the following error:

2025-05-27 12:29:41,646 [5441] ERROR    tornado.application: Exception in callback functools.partial(>, .wrapped() 
done, defined at \\
/Users/reviewer/myenv/lib/python3.13/site-packages/panel/io/server.py:153> 
exception=AbbreviatedException(, ValueError('Supplied cmap winter not found among matplotlib, bokeh, or colorcet 
colormaps.'), )>)

\textcolor{blue}{Thanks very much to the reviewer for bringing this bug to our attention. It appears to be a 
packaging error. An issue has been raised on the tsbrowse code repository 
\url{https://github.com/tskit-dev/tsbrowse/issues/248}. We are working on a solution, and will release a new 
Python package as soon as this is fixed.}\\

\hrule

We have also addressed the editorial comments and ensured the following:
\begin{itemize}
    \item \textcolor{red}{TO DO.} Code has been published on Zenodo and archival DOI updated in the code availability in addition 
        to code repository
    \item \textcolor{red}{TO DO.} Manuscript adheres to mandatory template format: \textcolor{red}{INSERT TEMPLATE LINK}
    \item \textcolor{red}{TO DO.} Supplementary material attached as separate pdf to main matter
    \item \textcolor{red}{TO DO.} Following .tex files included: .tex, .cls, .bib, .bst and .ps 
    \item \textcolor{red}{TO DO.} Mark up of changes in the revised text included as \textcolor{red}{INSERT NAME OF DOC}
    \item \textcolor{red}{TO DO.} Conflicts of interests declared for all authors
\end{itemize}

We look forward to hearing from you about our revised manuscript. Please let us know if you require any 
further clarification.

On behalf of all tsbrowse authors,
\closing{Sincerely,}

\end{letter}
\end{document}

### 2025-06-10 manuscript/paper.tex
typo in repo link
minor revisions and response to reviewers

@@
     \texttt{\url{https://pypi.org/project/tsbrowse/}}, \\ Development version:
-    \texttt{\url{https://github.com/tskit.dev/tsbrowse}}, \\ Documentation:
+    \texttt{\url{https://github.com/tskit-dev/tsbrowse}}, \\ Documentation:
     \texttt{\url{https://tskit.dev/tsbrowse/docs/}}\\
@@
 interest~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
-    wong2024general}. ARGs are a fundamental object in population genetics,
-and although they have been of theoretical interest for
-decades~\citep{Hudson1983,Griffiths1997} it is only with recent breakthroughs
-in inference methods~\citep{rasmussen2014genome,
-    speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
-    gunnarsson2024scalable,deng2024robust} that practical application has
-become possible. Varied applications have been recently proposed,
+wong2024general}.
+%Version 1
+%While early methods laid the groundwork for practical applications, 
+%it is with recent breakthroughs in inference methods ~\citep{rasmussen2014genome,
+%speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
+%gunnarsson2024scalable,deng2024robust} that scalability was achieved,
+%enabling the widespread application of ARGs to large-scale genomic data.
+%Version 2
+%ARGs are fundamental objects in population genetics,
+%with decades of foundational work informing their study and practical
+%applications~\citep{Hudson1983,Griffiths1996,Griffiths1997}.
+%~\citep{Hudson1983,Griffiths1996,Griffiths1997}.Recent breakthroughs 
+%in inference methods~\citep{rasmussen2014genome, speidel2019method, 
+%kelleher2019inferring, wohns2022unified, zhang2023biobank, 
+%gunnarsson2024scalable, deng2024robust} have enabled their 
+%widespread application to large-scale genomic data.
+%Version 3
+While decades of foundational work have informed the study of ARGs 
+and established them as fundamental objects in population genetics
+~\citep{Hudson1983,Griffiths1996,Griffiths1997}, recent breakthroughs 
+in inference methods~\citep{rasmussen2014genome, speidel2019method, 
+kelleher2019inferring, wohns2022unified, zhang2023biobank, 
+gunnarsson2024scalable, deng2024robust} have expanded their application
+to large-scale, genome-wide data.
+
+Varied applications have been proposed,
 such as inferring selection~\citep{stern2019approximate,hejase2022deep} and
@@
 whitehouse2024tree}.
-While these developements are exciting, the performance 
+While these developments are exciting, the performance 
 of these new methods depends critically on the accuracy of the inferred ARGs.
@@
 and storing as a compressed \texttt{.tsbrowse} file.
-(The \texttt{.tsbrowse} file is also a valid input for
-\texttt{tszip}, a general utility for compressing \texttt{tskit}
-files.)
+%(The \texttt{.tsbrowse} file is also a valid input for
+%\texttt{tszip}, a general utility for compressing \texttt{tskit}
+%files.)
 To visualise an ARG we then run \texttt{tsbrowse serve}, which
@@
 Holoviz ecosystem~\citep{Holoviz}, which efficiently rasterises large datasets
-at the requested resolution on the server, sends the image to the web-browser
+at the requested resolution on the server, sends the image to the web browser
 for display, and dynamically updates as the user interactively navigates the
@@
 This approach allows us to summarise very large ARGs interactively;
-Fig~\ref{fig:Supplementary_Figure_4}, for example, shows a screenshot of
+Fig~\ref{fig:Supplementary_Figure_1}, for example, shows a screenshot of
 \texttt{tsbrowse} summarising the 1.9 million mutations in a SARS-CoV-2
@@
 % rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
-% for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
+% for details, and Supplementary Figure \ref{fig:Supplementary_Figure_1} for a
 % tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
@@
 The user can interact with the main ``genome browser'' window
-using a set of controls on the top-right corner (provided by Bokeh),
-allowing the user to pan and zoom as required. The histograms to the
+using a set of controls on the top-right corner (provided by 
+Bokeh~\citep{bokeh}),allowing them to pan and zoom as required. 
+The histograms to the
 right then summarise the edges as depicted in the browser window.
 A similar browser interface is provided for
-nodes
-(e.g. Fig~\ref{fig:Supplementary_Figure_3})
-and mutations
-(e.g. Fig~\ref{fig:Supplementary_Figure_4}).
+mutations
+(e.g. Fig~\ref{fig:Supplementary_Figure_1})
+and nodes
+(e.g. Fig~\ref{fig:Supplementary_Figure_2}).
 \texttt{Tsbrowse} also provides an interactive table viewer with flexible
-search and sorting utilities, which is a valuable debugging utility for
+searching and sorting utilities, which is a valuable debugging utility for
 developers.
@@
 of a selective sweep. Comparing this ground truth to the ARGs inferred
-by four different inference methods in Fig~\ref{fig:Supplementary_Figure_1},
+by four different inference methods in Fig~\ref{fig:Supplementary_Figure_3},
 we can see that there are substantial qualitative differences between the
@@
 choices on ARGs inferred by \texttt{tsinfer}, and it is now an indispensable
-element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_2} 
+element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_4} 
 shows a region of the 1000G data with gaps in site density that are spanned by
@@
 the \texttt{tsinfer} inference algorithm.
-Fig~\ref{fig:Supplementary_Figure_3} shows the genomic spans of
+Fig~\ref{fig:Supplementary_Figure_2} shows the genomic spans of
 ancestral nodes in an ARG inferred with \texttt{tsinfer}, demonstrating a clear
@@
 % (Supplementary Figures 
-% \ref{fig:Supplementary_Figure_3}). Supplementary Figure
-% \ref{fig:Supplementary_Figure_2} shows how mutation patterns in a 2D plot of
+% \ref{fig:Supplementary_Figure_2}). Supplementary Figure
+% \ref{fig:Supplementary_Figure_4} shows how mutation patterns in a 2D plot of
 % genomic position and mutation age are helpful in spotting poorly inferred
@@
 
-% Supplementary Fig. \ref{fig:Supplementary_Figure_4}, a screenshot of the
+% Supplementary Fig. \ref{fig:Supplementary_Figure_1}, a screenshot of the
 % Mutations page for SARS-CoV-2 ARGs \citep{zhan2023towards} demonstrates
@@
 to capture topological features at different scales,
-and is an active research area~\citep{wong2022dynamic,kramer2023treenome}.
+and is an active research area~\citep{hadfield2018nextstrain, wong2022dynamic,kramer2023treenome}.
 Adapting such methods, and integrating into \texttt{tsbrowse} to provide
@@
 \subsection{Inference of selective sweep dataset (Suppl. Fig.
-    \ref{fig:Supplementary_Figure_1})} The following software were used to infer
+    \ref{fig:Supplementary_Figure_3})} The following software were used to infer
 ARGs from the truth dataset described above: \texttt{tsinfer version 0.3.3}
@@
 
-\subsection{Inference of 1000 Genomes dataset (Suppl. Fig.
-\ref{fig:Supplementary_Figure_2}, Suppl. Fig.
-\ref{fig:Supplementary_Figure_3})} The 1000 Genomes dataset was downloaded from
-\texttt{\url{https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/working/20220422_3202_phased_SNV_INDEL_SV/}}.
-The ancestral fasta sequence for chromosome 17 (GRCh38) was downloaded
-from the Ensembl database. Inference was performed with a Snakemake pipeline
-(\texttt{\url{https://github.com/benjeffery/tsinfer-snakemake/}}) using
-\texttt{tsinfer version 0.3.3} for the long arm of chromosome 17 after
-filtering out duplicate variant positions, variants with missing or low quality
-ancestral allele, singletons, n-1-tons and n-2-tons. Only bi-allelic SNPs were
-included for inference. For Suppl. Fig. \ref{fig:Supplementary_Figure_2}, 
-\texttt{tsdate version 0.2.1} was used to estimate the age of ancestral nodes
- with \textit{mutation rate = 1.29e-8}, setting all other parameters to default 
- values.
-
 \subsection{Inference of SARS-CoV-2 ARGs}
-The ARG shown in Fig \ref{fig:Supplementary_Figure_4} was inferred with
+The ARG shown in Fig \ref{fig:Supplementary_Figure_1} was inferred with
 \texttt{sc2ts}~\citep{zhang2023biobank} using the Viridian
@@
 
-\subsection{} Code to recreate datasets and infer ARGs are available in an
+\subsection{Inference of 1000 Genomes dataset (Suppl. Fig.
+\ref{fig:Supplementary_Figure_3}, Suppl. Fig.
+\ref{fig:Supplementary_Figure_4})} The 1000 Genomes dataset was downloaded from
+\texttt{\url{https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/working/20220422_3202_phased_SNV_INDEL_SV/}}.
+The ancestral fasta sequence for chromosome 17 (GRCh38) was downloaded
+from the Ensembl database. Inference was performed with a Snakemake pipeline
+(\texttt{\url{https://github.com/benjeffery/tsinfer-snakemake/}}) using
+\texttt{tsinfer version 0.3.3} for the long arm of chromosome 17 after
+filtering out duplicate variant positions, variants with missing or low quality
+ancestral allele, singletons, n-1-tons and n-2-tons. Only bi-allelic SNPs were
+included for inference. For Suppl. Fig. \ref{fig:Supplementary_Figure_4}, 
+\texttt{tsdate version 0.2.1} was used to estimate the age of ancestral nodes
+ with \textit{mutation rate = 1.29e-8}, setting all other parameters to default 
+ values.
+
+\subsection{} Code to recreate datasets and infer ARGs is available in an
 accompanying repository:
@@
 \section{Supplementary Figures}
-\begin{figure}[H]
+\begin{figure}[h]
     \centering
     \includegraphics[width=0.95\linewidth]{figures/SuppFig1.png}
+    \caption{\textbf{\texttt{tsbrowse} applied to SARS-CoV-2 ARGs.}
+        A screenshot of \texttt{tsbrowse}'s depiction of
+        1,923,169 mutations in an
+        ARG inferred by \texttt{sc2ts}; see  text for details. Also shown are the
+        gene annotations along the X-axis.}
+    \label{fig:Supplementary_Figure_1}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.95\linewidth]{figures/SuppFig2.png}
+    \caption{\textbf{A view of the Nodes page for a 1000 Genomes inference.} A
+        screenshot of \texttt{tsbrowse}'s Nodes view for an inference of the long arm
+        of chromosome 17 from the 1000 Genomes whole-genome sequencing dataset. At the
+        top is a plot of node spans over time. The length of sequence that the nodes
+        span is shown on the X axis, and the time of nodes is shown on the Y axis. The
+        histogram at the bottom shows the distribution of node spans. }
+    \label{fig:Supplementary_Figure_2}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.95\linewidth]{figures/SuppFig3.png}
     \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
@@
 time of either the parent or child node in the edge.}
-    \label{fig:Supplementary_Figure_1}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.95\linewidth]{figures/SuppFig2.png}
-    \caption{\textbf{Identifying ARG inference problems with \texttt{tsbrowse}} 
-Screenshots of \texttt{tsbrowse}'s Mutations view (A) and Edges view (B) for a
- 600 kB region of chromosome 17 inferred from 3,202 participants from the 1000 
- Genomes Whole Genome Sequencing dataset (1000G) \citep{1000G2015, 1000GWGS}. 
- The poor performance of \texttt{tsinfer} in this variant-poor region is evidenced
-  by the long edges spanning gaps in mutation density.}
-    \label{fig:Supplementary_Figure_2}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.95\linewidth]{figures/SuppFig3.png}
-    \caption{\textbf{A view of the Nodes page for a 1000 Genomes inference.} A
-        screenshot of \texttt{tsbrowse}'s Nodes view for an inference of the long arm
-        of chromosome 17 from the 1000 Genomes whole-genome sequencing dataset. At the
-        top is a plot of node spans over time. The length of sequence that the nodes
-        span is shown on the X axis, and the time of nodes is shown on the Y axis. The
-        histogram at the bottom shows the distribution of node spans. }
     \label{fig:Supplementary_Figure_3}
@@
     \includegraphics[width=0.95\linewidth]{figures/SuppFig4.png}
-    \caption{\textbf{\texttt{tsbrowse} applied to SARS-CoV-2 ARGs.}
-        A screenshot of \texttt{tsbrowse}'s depiction of
-        1,923,169 mutations in an
-        ARG inferred by \texttt{sc2ts}; see  text for details. Also shown are the
-        gene annotations along the X-axis.}
+    \caption{\textbf{Identifying ARG inference problems with \texttt{tsbrowse}} 
+Screenshots of \texttt{tsbrowse}'s Mutations view (A) and Edges view (B) for a
+ 600 kB region of chromosome 17 inferred from 3,202 participants from the 1000 
+ Genomes Whole Genome Sequencing dataset (1000G) \citep{1000GWGS}. 
+ The poor performance of \texttt{tsinfer} in this variant-poor region is evidenced
+  by the long edges spanning gaps in mutation density.}
     \label{fig:Supplementary_Figure_4}

### 2025-06-11 manuscript/revision/response_to_editor.tex
Addressing PR review; addressing editor's comments

@@
 \address{Big Data Institute\\University of Oxford\\UK}
+
 \begin{document}
@@
 
-We are delighted that the reviewers have found our manuscript suitable for publication 
-in Bionformatics. We thank you, the editors and all the reviewers for the careful 
-consideration of our manuscript, and value the kind words and constructive feedback 
-provided on our work. We have now revised the manuscript based on the reviewers' 
-suggestions. Below is our point-by-point response to the 
-reviewers' comments in blue. We have incorporated as many suggestions as possible, 
-and where unable to do so we have indicated the reason.\\
-
-\hrule
-
-Reviewer: 1
-
-Comments to the Author \\
-
-Summary: In this manuscript, the authors present tsbrowse, an open-source Python-based 
-application for the interactive visualization of the components of Ancestral Recombination 
-Graphs (ARGs). Because ARGs (in principle) contain all knowable information concerning 
-coalescence and recombination in the history of a set of samples, there is great excitement 
-in developing methods for inferring ARGs from empirical genomic sequence data, as well as 
-for learning about evolutionary processes from patterns in inferred ARGs. A crucial component 
-of both these endeavors is visualizing ARGs, which is difficult or unwieldy. Tsbrowse therefore 
-represents an exciting contribution to the field which is likely to be of high utility and 
-to be highly cited.
-
-Opinion: This paper was straightforward, clear, and well written. I think it does a good job 
-setting up the intellectual context for ARG-based research and advocating for the utility of 
-tsbrowse. I had two minor, nitpicky comments, but overall, it was great.  Thanks for a fun and 
-easy review!
-
-1. I think the order in which the supplemental figures are cited doesn’t match their numbering 
-(i.e., S4 is cited before S1)\\
-\textcolor{blue}{Thanks to Reviewer 1 for pointing out this error. We have now rearranged the 
-supplementary figures so that the citations appear in ascending order.}
-
-2. In the discussion, the authors mention that methods for visualization of tree topology can 
-typically only handle a few hundred nodes, but it might be worth doing a “(but see…)” for some 
-of the approaches folks have taken for visualizing much larger trees (e.g., nextstrain, onezoom).\\
-\textcolor{blue}{We agree with the reviewer-  it is important to acknowledge previous/on-going visualisation efforts 
-for larger trees. We had already included a citation to the OneZoom project (and the Treenome browser) in the 
-sentence that followed the one in question "… Visualisation of large-scale tree topologies with millions of nodes 
-requires much more sophisticated approaches to capture topological features at different scales, 
-and is an active research area". We are thankful to the reviewer for bringing Nextstrain to our 
-attention, and have now included Hadfield et al. as a citation.}
-
-Reviewer: 2\\
-
-Comments to the Author\\
-
-I have only a few minor comments:\\
-1. "and storing as a compressed .tsbrowse file. (The .tsbrowse file is also a valid input for
-tszip, a general utility for compressing tskit files.)" Sorry, but I don't understand this sentence. 
-More precisely, both the compressed and the uncompressed files have the .tbrowse extension? If yes, 
-that's a weird choice and I would like to see some justification. You should also make clear if tszip 
-is a part of tskit or of twbrowse.\\
-\textcolor{blue}{To clarify, the tszip is part of the tskit ecosystem. It is not part of the tsbrowse 
-package but can be used to compress the tables output from the tsbrowse preprocess step. Thanks to 
-the reviewer for this pointing out this potentially unclear statement. We have now deleted the text 
-in brackets, which we believe isn't crucial for understanding the data model.}
-
-2. "web-browser" - web browser\\
-\textcolor{blue}{Fixed, thanks for pointing out the typographical error.}
-
-3. "Fig S4, for example, shows a screenshot of tsbrowse". This is a matter of personal taste, but I do
-not like to refer to a Figure in the supplementary materia just as if it were in the article; if the 
-Figure is relevant, it should be in the article. If it's not relevant, that you should refer to it without
-hinting that it is important.\\
-\textcolor{blue}{We agree with the reviewer's opinion, and believe that Figure S4 is crucial 
-in order to demonstrate the application of tsbrowse to large-scale datasets (which is a key design 
-feature of the app). However, we are unable to move this figure to the main matter considering journal 
-guidelines for page limit.}
-
-4. Looking at the github repo, I can find the installation instruction using pip only. In my experience 
-pip is quite picky. I strongly suggest you to provide tsbrowse via conda-forge/bio-conda or a containerized 
-solution (e.g. docker).\\
-\textcolor{blue}{Thanks to the reviewer for the suggestion to provide alternate installation solutions. 
-We have now created a Docker container in order to ensure a smooth experience across computing environments. 
-Links have been updated in the manuscript and app documentation}
-\textcolor{red}{[TO DO; after release of new package]}
-
-Reviewer: 3
-
-Comments to the Author
-This paper offers a browser-based visualisation tool for ancestral recombination graphs (ARGs), which is 
-a data structure representing gene genealogies for a sample of DNA sequences. The complication in the p
-resence of recombination is that these structures are not simple trees, but - because of the possibility 
-of recombination - graphs with reticulated nodes, which makes exploratory data analysis difficult.
-
-I agree with the authors that there is currently a lack of visualisation tools for ARGs, and I expect the 
-current paper to provide a useful addition to the catalog of tools for ARG simulation and inference. Tsbrowse 
-integrates with the widely used tskit ecosystem of software tools, developed by some of the current authors, 
-which will facilitate a rapid uptake. The authors make a compelling argument for the need for visualisation 
-tools - Fig S1 in particular shows that four different and popular tools for ARG inference are giving 
-qualitatively completely different insights into their reconstructed ARGs (incidentally, it's a shame that 
-that figure couldn't make it into the main paper, though I recognise that this journal has strict page limits).\\
-\textcolor{blue}{We appreciate the reviewer's feedback on the utility of tsbrowse, and acknowledge the 
-importance of S1 in highlighting the need for visualisation. However, considering the necessity of the 
-existing main display items to establish the fundamental aspects of tsbrowse, we have decided to retain S1 
-in the supplementary. We agree that S1 would have been a compelling main display item had the page/figure 
-limits of the journal allowed for it.}
-
-My main comment is that tsbrowse does not actually seem to be an ARG visualisation tool. It is a useful way 
-to visualise mutations, tree-sequence edges, and summary statistics, but nowhere will you actually see a tree 
-(much less an ARG). On rereading the paper I see that the manuscript has been carefully worded not to include 
-such a claim, but a casual reader might expect tsbrowse to be able to do so, as other tools such as IcyTree can. 
-I do get fed up with reviewers tossing additional requests on top of an already polished piece of work, but 
-in this case I think it would be a helpful addition! - would it really be too much work for tsbrowse to be 
-able to draw at least the local trees? (For researchers with smaller datasets, being able to interrogate the 
-fine details of an ARG would be equally valuable as the complementary selling-points around scalability.) At 
-the very least I would like to see the manuscript at least be a little more up-front about this omission.\\
-
-\textcolor{blue}{We thank the reviewer for making this important observation, and agree that drawing 
-(parts of) local trees would be a great feature. For smaller datasets, tools such as arg-visualiser 
-(\url{https://pypi.org/project/tskit-arg-visualizer/}) and existing tskit infrastructure, for example 
-the \texttt{draw\_text()} and \texttt{draw\_svg()} already provide this functionality. A key 
-motivation for this work is to guide large-scale inference, and implementing tree visualisation that is interpretable 
-at this scale would require significant thought towards design as well as considerable effort. We therefore 
-think that implementing this feature is not feasible for this work and is better scoped as future work or as 
-a separate endeavour. In the text, we do state that tsbrowse is a visualiser for the information structure 
-in ARGs, which we hope implies that it is a visualiser for the properties of the ARG, rather than the ARG 
-topology itself. We have also already explicitly stated in the Discussion that drawing local trees is an 
-important future direction.}\\
-
-Minor comments:\\
-
-Manuscript p1: 'it is only with recent breakthroughs in inference methods...that practical application 
-has become possible'. I appreciate the need to emphasise the recent increase in interest in inference for ARGs, 
-but I think this is selling older methods a little short. There are ARG inference methods going back at least 
-as far as Griffiths and Marjoram (1996) 'Ancestral Inference from Samples of DNA Sequences' JCB 3(4):479-502. 
-True they are not scalable in the manner of recent methods, but they *are* practical applications, and they have 
-the advantage that they actually sample from the claimed model without introducing any model heuristics.
-\textcolor{blue}{Thanks to the reviewer for their feedback on this statement on early methods. We agree, 
-and have revised the text to acknowledge the contributions of foundational work in ARG inference.}
-
-Manuscript p1: Typo 'developements'.\\
-\textcolor{blue}{Thanks, this is now fixed.}
-
-Manuscript p2: Where you mention 'Bokeh', can you clarify what this is (e.g. with a citation)?\\
-\textcolor{blue}{Yes, Bokeh is a Python library for interactive visualisation and a citation to it 
-has been added.}
-\textcolor{red}{TO FIX! Citation appears as a question mark!!.}
-
-Supplement p1, Sec 1.4: "The ARG shown in Fig 4" should say "The ARG shown in Suppl. Fig 4" ?\\
-\textcolor{blue}{Thanks, this is now fixed.}\\
-\textcolor{red}{(TO DO: make sure to replace links in suppl pdf!)}
-
-Fig S2 caption - 'Consortium (2015)' is citing incorrectly.\\
-\textcolor{blue}{Fixed (removed this citation as it was not required)}
-
-From the online documentation, Node page: \url{https://tskit.dev/tsbrowse/docs/stable/nodes.html}
-"The histograms at the bottom show the distributions of node spans over different dimensions. 
-The leftmost histogram summarises the span of nodes on the sequence; the middle plot summarises 
-the span of nodes over time and the rightmost plot summarises the edge “area” defined as the product 
-of sequence span and time span for each node."
-
-I didn't understand this. The 'nodes' section of the visualiser only seems to provide one histogram, 
-not three.\\
-
-\textcolor{blue}{Thanks to the reviewer for pointing out this documentation error. The Nodes section 
-does indeed have one histogram and the documentation has been updated to reflect this.}
-
-Installation notes:
-
-I had a few problems getting tsbrowse to work. There are two different sets of installation instructions:
-
-1. https://pypi.org/project/tsbrowse/
-2. https://tskit.dev/tsbrowse/docs/stable/contributing.html
-
-I use a Mac+Homebrew+Chrome and Option 2 seems more fitting. ("python -m pip install tsbrowse" as in 
-option 1 will fail because Homebrew's Python environment is externally managed.)
-
-\textcolor{blue}{We apologise for any confusion on installation instructions. The recommended way to install 
-tsbrowse is via pip within a virtual environment. We have fixed the documentation to show only one set of 
-instructions (i.e, for the user) at the top.}
-
-Bug(?):
-
-I frequently had problems rendering the window in the 'Mutations' tab, which tended to get stuck at the 
-circling wheel. When clicking the 'Mutations' button, my terminal returned the following error:
-
-2025-05-27 12:29:41,646 [5441] ERROR    tornado.application: Exception in callback functools.partial(>, .wrapped() 
-done, defined at \\
-/Users/reviewer/myenv/lib/python3.13/site-packages/panel/io/server.py:153> 
-exception=AbbreviatedException(, ValueError('Supplied cmap winter not found among matplotlib, bokeh, or colorcet 
-colormaps.'), )>)
-
-\textcolor{blue}{Thanks very much to the reviewer for bringing this bug to our attention. It appears to be a 
-packaging error. An issue has been raised on the tsbrowse code repository 
-\url{https://github.com/tskit-dev/tsbrowse/issues/248}. We are working on a solution, and will release a new 
-Python package as soon as this is fixed.}\\
-
-\hrule
+We are delighted to receive such positive feedback, and thank the reviewers for their 
+careful consideration of the manuscript and software, and the constructive and helpful 
+input provided. Attached are the revised manuscript (\textit{paper.pdf}), 
+a version of the manuscript showing the differences to the submitted version ("diff.pdf"), 
+and a response to the reviewers' comments (\textit{response\_to\_reviwers.pdf})
 
@@
 \begin{itemize}
-    \item \textcolor{red}{TO DO.} Code has been published on Zenodo and archival DOI updated in the code availability in addition 
-        to code repository
-    \item \textcolor{red}{TO DO.} Manuscript adheres to mandatory template format: \textcolor{red}{INSERT TEMPLATE LINK}
-    \item \textcolor{red}{TO DO.} Supplementary material attached as separate pdf to main matter
-    \item \textcolor{red}{TO DO.} Following .tex files included: .tex, .cls, .bib, .bst and .ps 
-    \item \textcolor{red}{TO DO.} Mark up of changes in the revised text included as \textcolor{red}{INSERT NAME OF DOC}
-    \item \textcolor{red}{TO DO.} Conflicts of interests declared for all authors
+    \item Code published on Zenodo and archival DOI updated 
+    under Availability, in addition to code repository
+    \item Manuscript adheres to the Overleaf template, \textit{OUP General Template}
+    \item Supplementary material attached as separate pdf, \textit{supplement.pdf}
+    \item All relevant LaTeX files included
+    \item Added Conflicts of Interest statement to the manuscript (none declared)
 \end{itemize}

### 2025-06-19 manuscript/revision/response_to_reviewers.tex

\documentclass{article}
\usepackage{xcolor}
\usepackage{hyperref}
\title{Response to reviewers' comments}

\begin{document}
\maketitle

Below is our point-by-point response to the reviewers' comments in blue.\\

\textbf{Reviewer: 1}\\

Comments to the Author \\

Summary: In this manuscript, the authors present tsbrowse, an open-source Python-based 
application for the interactive visualization of the components of Ancestral Recombination 
Graphs (ARGs). Because ARGs (in principle) contain all knowable information concerning 
coalescence and recombination in the history of a set of samples, there is great excitement 
in developing methods for inferring ARGs from empirical genomic sequence data, as well as 
for learning about evolutionary processes from patterns in inferred ARGs. A crucial component 
of both these endeavors is visualizing ARGs, which is difficult or unwieldy. Tsbrowse therefore 
represents an exciting contribution to the field which is likely to be of high utility and 
to be highly cited.

Opinion: This paper was straightforward, clear, and well written. I think it does a good job 
setting up the intellectual context for ARG-based research and advocating for the utility of 
tsbrowse. I had two minor, nitpicky comments, but overall, it was great. Thanks for a fun and 
easy review!\\
\textcolor{blue}{Thank you for your kind words!}

1. I think the order in which the supplemental figures are cited doesn’t match their numbering 
(i.e., S4 is cited before S1)\\
\textcolor{blue}{We have now rearranged the supplementary figures so that figure references 
appear in ascending order.}

2. In the discussion, the authors mention that methods for visualization of tree topology can 
typically only handle a few hundred nodes, but it might be worth doing a ``(but see…)'' for some 
of the approaches folks have taken for visualizing much larger trees (e.g., nextstrain, onezoom).\\
\textcolor{blue}{We had already included a citation to the OneZoom project (and the Treenome browser) in the 
sentence that followed the one in question "… Visualisation of large-scale tree topologies with millions of nodes 
requires much more sophisticated approaches to capture topological features at different scales, 
and is an active research area". Thanks for bringing Nextstrain to our 
attention, and we have now Hadfield et al. as a citation for visualisation of medium-sized toplogies.}\\

\textbf{Reviewer: 2}\\

Comments to the Author\\

I have only a few minor comments:\\
1. ``and storing as a compressed .tsbrowse file. (The .tsbrowse file is also a valid input for
tszip, a general utility for compressing tskit files.)'' Sorry, but I don't understand this sentence. 
More precisely, both the compressed and the uncompressed files have the .tbrowse extension? If yes, 
that's a weird choice and I would like to see some justification. You should also make clear if tszip 
is a part of tskit or of twbrowse.\\
\textcolor{blue}{To clarify, the tszip is part of the tskit ecosystem. It is not part of the tsbrowse 
package but can be used to compress the tables output from the tsbrowse preprocess step. Thank you 
for pointing out this potentially unclear statement. We have now deleted the text 
in brackets, which we believe isn't crucial for understanding the data model.}\\

2. ``web-browser'' - web browser\\
\textcolor{blue}{Fixed.}\\

3. ``Fig S4, for example, shows a screenshot of tsbrowse''. This is a matter of personal taste, but I do
not like to refer to a Figure in the supplementary materia just as if it were in the article; if the 
Figure is relevant, it should be in the article. If it's not relevant, that you should refer to it without
hinting that it is important.\\
\textcolor{blue}{We agree with the reviewer's opinion, and believe that this figure is important 
to demonstrate the application of tsbrowse to large-scale datasets (which is a key design 
feature of the app). However, we are unable to move this figure to the main matter considering journal 
guidelines for page limit.}\\

4. Looking at the github repo, I can find the installation instruction using pip only. In my experience 
pip is quite picky. I strongly suggest you to provide tsbrowse via conda-forge/bio-conda or a containerized 
solution (e.g. docker).\\
\textcolor{blue}{We have now created a Docker container in order to ensure a smooth experience across 
computing environments. Usage instructions have been added to the app documentation.}\\

\textbf{Reviewer: 3}\\

Comments to the Author\\
This paper offers a browser-based visualisation tool for ancestral recombination graphs (ARGs), which is 
a data structure representing gene genealogies for a sample of DNA sequences. The complication in the p
resence of recombination is that these structures are not simple trees, but - because of the possibility 
of recombination - graphs with reticulated nodes, which makes exploratory data analysis difficult.

I agree with the authors that there is currently a lack of visualisation tools for ARGs, and I expect the 
current paper to provide a useful addition to the catalog of tools for ARG simulation and inference. Tsbrowse 
integrates with the widely used tskit ecosystem of software tools, developed by some of the current authors, 
which will facilitate a rapid uptake. The authors make a compelling argument for the need for visualisation 
tools - Fig S1 in particular shows that four different and popular tools for ARG inference are giving 
qualitatively completely different insights into their reconstructed ARGs (incidentally, it's a shame that 
that figure couldn't make it into the main paper, though I recognise that this journal has strict page limits).\\
\textcolor{blue}{Thank you for your valuable feedback on the manuscript, and on the utility of tsbrowse!}

\textcolor{blue}{We completely agree that this figure would have been a compelling main display item had the journal 
page/figure limits allowed for it. Considering the necessity of the existing main display items in establishing 
the fundamental aspects of tsbrowse, we have decided to retain the figure in the supplementary.}\\

My main comment is that tsbrowse does not actually seem to be an ARG visualisation tool. It is a useful way 
to visualise mutations, tree-sequence edges, and summary statistics, but nowhere will you actually see a tree 
(much less an ARG). On rereading the paper I see that the manuscript has been carefully worded not to include 
such a claim, but a casual reader might expect tsbrowse to be able to do so, as other tools such as IcyTree can. 
I do get fed up with reviewers tossing additional requests on top of an already polished piece of work, but 
in this case I think it would be a helpful addition! - would it really be too much work for tsbrowse to be 
able to draw at least the local trees? (For researchers with smaller datasets, being able to interrogate the 
fine details of an ARG would be equally valuable as the complementary selling-points around scalability.) At 
the very least I would like to see the manuscript at least be a little more up-front about this omission.\\
\textcolor{blue}{This is an important observation, and we agree that drawing (parts of) local trees would be 
a great feature. For smaller datasets, tools such as \texttt{tskit-arg-visualizer} 
(\url{https://pypi.org/project/tskit-arg-visualizer/}) and existing tskit infrastructure, for example 
functions \texttt{draw\_text()} and \texttt{draw\_svg()} already provide this functionality. A key 
motivation for this work is to guide large-scale inference, and implementing tree visualisation that is 
interpretable at this scale would require significant thought towards design, as well as considerable effort. 
We therefore think that implementing this feature is not feasible for this work and is better scoped as future 
work or as a separate endeavour. In the text, we do state that tsbrowse is a visualiser for the information 
structure in ARGs, which we hope implies that it is a visualiser for the properties of the ARG, rather than 
the ARG topology itself. We have also already explicitly stated in the Discussion that drawing local trees is 
an important future direction.}\\

Minor comments:\\

Manuscript p1: 'it is only with recent breakthroughs in inference methods...that practical application 
has become possible'. I appreciate the need to emphasise the recent increase in interest in inference for ARGs, 
but I think this is selling older methods a little short. There are ARG inference methods going back at least 
as far as Griffiths and Marjoram (1996) 'Ancestral Inference from Samples of DNA Sequences' JCB 3(4):479-502. 
True they are not scalable in the manner of recent methods, but they *are* practical applications, and they have 
the advantage that they actually sample from the claimed model without introducing any model heuristics.\\
\textcolor{blue}{Thank you for pointing this out, and you are right that we are somewhat overstating the 
case for recent methods. We have changed the word "practical" to "widespread", which we feel is an uncontroversial 
statement. The history of ARG inference is a complicated and confusing subject, dealt with exhaustively by 
Wong et al 2024, and would be impossible to accurately summarise in the available space. We note, though, 
that two of the recent methods cited here (ARGweaver and SINGER) sample ARGs from the posterior defined by a 
coalescent model in a statistically rigorous manner}\\

Manuscript p1: Typo 'developements'.\\
\textcolor{blue}{This is now fixed.}\\

Manuscript p2: Where you mention 'Bokeh', can you clarify what this is (e.g. with a citation)?\\
\textcolor{blue}{Yes, Bokeh is a Python library for interactive visualisation and a citation to it 
has been added.}\\

Supplement p1, Sec 1.4: ``The ARG shown in Fig 4'' should say "The ARG shown in Suppl. Fig 4" ?\\
\textcolor{blue}{This is now fixed.}\\

Fig S2 caption - 'Consortium (2015)' is citing incorrectly.\\
\textcolor{blue}{Fixed (removed this citation as it was not required)}\\

From the online documentation, Node page: \url{https://tskit.dev/tsbrowse/docs/stable/nodes.html}
``The histograms at the bottom show the distributions of node spans over different dimensions. 
The leftmost histogram summarises the span of nodes on the sequence; the middle plot summarises 
the span of nodes over time and the rightmost plot summarises the edge “area” defined as the product 
of sequence span and time span for each node.''

I didn't understand this. The 'nodes' section of the visualiser only seems to provide one histogram, 
not three.\\
\textcolor{blue}{Thank you for pointing out this documentation error. The Nodes section 
does indeed have one histogram and the documentation has been updated to reflect this.}\\

Installation notes:

I had a few problems getting tsbrowse to work. There are two different sets of installation instructions:

1. https://pypi.org/project/tsbrowse/
2. https://tskit.dev/tsbrowse/docs/stable/contributing.html

I use a Mac+Homebrew+Chrome and Option 2 seems more fitting. ("python -m pip install tsbrowse" as in 
option 1 will fail because Homebrew's Python environment is externally managed.)\\
\textcolor{blue}{We apologise for the confusion surrounding installation notes. The recommended way to install 
tsbrowse is via pip within a virtual environment. We have fixed the documentation so that instructions 
to install tsbrowse as a new user are clearly stated at the beginning of the documentation,
and separate from developer instructions.}\\

Bug(?):\\

I frequently had problems rendering the window in the 'Mutations' tab, which tended to get stuck at the 
circling wheel. When clicking the 'Mutations' button, my terminal returned the following error:

2025-05-27 12:29:41,646 [5441] ERROR    tornado.application: Exception in callback functools.partial(>, .wrapped() 
done, defined at \\
/Users/reviewer/myenv/lib/python3.13/site-packages/panel/io/server.py:153> 
exception=AbbreviatedException(, ValueError('Supplied cmap winter not found among matplotlib, bokeh, or colorcet 
colormaps.'), )>)\\
\textcolor{blue}{We are grateful to the reviewer for bringing this bug to our attention. 
We have solved the issue and released an updated version of the package.}\\

\end{document}

### 2025-06-19 manuscript/paper.tex

@@
     \texttt{\url{https://github.com/tskit-dev/tsbrowse}}, \\ Documentation:
-    \texttt{\url{https://tskit.dev/tsbrowse/docs/}}\\
+    \texttt{\url{https://tskit.dev/tsbrowse/docs/}}, \\ DOI: 
+    \texttt{\url{https://doi.org/10.5281/zenodo.15683039}}\\
     %\textbf{Contact:} \href{name@email.com}{name@email.com}\\
@@
 interest~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
-wong2024general}.
-%Version 1
-%While early methods laid the groundwork for practical applications, 
-%it is with recent breakthroughs in inference methods ~\citep{rasmussen2014genome,
-%speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
-%gunnarsson2024scalable,deng2024robust} that scalability was achieved,
-%enabling the widespread application of ARGs to large-scale genomic data.
-%Version 2
-%ARGs are fundamental objects in population genetics,
-%with decades of foundational work informing their study and practical
-%applications~\citep{Hudson1983,Griffiths1996,Griffiths1997}.
-%~\citep{Hudson1983,Griffiths1996,Griffiths1997}.Recent breakthroughs 
-%in inference methods~\citep{rasmussen2014genome, speidel2019method, 
-%kelleher2019inferring, wohns2022unified, zhang2023biobank, 
-%gunnarsson2024scalable, deng2024robust} have enabled their 
-%widespread application to large-scale genomic data.
-%Version 3
-While decades of foundational work have informed the study of ARGs 
-and established them as fundamental objects in population genetics
-~\citep{Hudson1983,Griffiths1996,Griffiths1997}, recent breakthroughs 
-in inference methods~\citep{rasmussen2014genome, speidel2019method, 
-kelleher2019inferring, wohns2022unified, zhang2023biobank, 
-gunnarsson2024scalable, deng2024robust} have expanded their application
-to large-scale, genome-wide data.
-
-Varied applications have been proposed,
+wong2024general}. ARGs are a fundamental object in population genetics,
+and although they have been of theoretical interest for
+decades~\citep{Hudson1983,Griffiths1996,Griffiths1997} it is only with recent breakthroughs
+in inference methods~\citep{rasmussen2014genome,speidel2019method,kelleher2019inferring,
+wohns2022unified,zhang2023biobank,gunnarsson2024scalable,deng2024robust} that 
+widespread application has become possible. Varied applications have been proposed,
 such as inferring selection~\citep{stern2019approximate,hejase2022deep} and
@@
 This approach allows us to summarise very large ARGs interactively;
-Fig~\ref{fig:Supplementary_Figure_1}, for example, shows a screenshot of
+%Fig~\ref{fig:Supplementary_Figure_1}, for example, shows a screenshot of
+Fig.~S1 for example, shows a screenshot of
 \texttt{tsbrowse} summarising the 1.9 million mutations in a SARS-CoV-2
@@
 describe various aspects of the ARG.
-Figure~\ref{fig:Figure_2}
+Fig.~\ref{fig:Figure_2}
 demonstrates the user interface using the Edges page as an example.
@@
 in a simulated ARG with a strong selective sweep (see
-Sec.~\ref{subsec:sweep_simulation} for details).
+Supplement for details).
 Each edge is depicted as a horizontal line connecting the genomic coordinates
@@
 The user can interact with the main ``genome browser'' window
-using a set of controls on the top-right corner (provided by 
-Bokeh~\citep{bokeh}),allowing them to pan and zoom as required. 
+using a set of controls on the top-right corner provided by 
+Bokeh~\citep{bokeh},allowing them to pan and zoom as required. 
 The histograms to the
@@
 mutations
-(e.g. Fig~\ref{fig:Supplementary_Figure_1})
+%(e.g. Fig~\ref{fig:Supplementary_Figure_1})
+(e.g., Fig.~S1)
 and nodes
-(e.g. Fig~\ref{fig:Supplementary_Figure_2}).
+%(e.g. Fig~\ref{fig:Supplementary_Figure_2}).
+(e.g., Fig.~S2).
 \texttt{Tsbrowse} also provides an interactive table viewer with flexible
@@
 
-%The ability to explore ARG properties across the sequence and interact with the data makes the analysis of local genealogical patterns more intuitive.
+%The ability to explore ARG properties across the sequence and interact with the 
+%data makes the analysis of local genealogical patterns more intuitive.
 
@@
 using the genome-browser-like perspective provided by \texttt{tsbrowse}
-is given in the simulated ARG of Fig~\ref{fig:Figure_2}, where
+is given in the simulated ARG of Fig.~\ref{fig:Figure_2}, where
 the gap in the edges view corresponds to the characteristic dip in diversity
 of a selective sweep. Comparing this ground truth to the ARGs inferred
-by four different inference methods in Fig~\ref{fig:Supplementary_Figure_3},
+%by four different inference methods in Fig~\ref{fig:Supplementary_Figure_3},
+by four different inference methods in Fig.~S3,
 we can see that there are substantial qualitative differences between the
@@
 choices on ARGs inferred by \texttt{tsinfer}, and it is now an indispensable
-element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_4} 
+%element of the inference pipeline. Fig~\ref{fig:Supplementary_Figure_4} 
+element of the inference pipeline. Fig.~S4
 shows a region of the 1000G data with gaps in site density that are spanned by
@@
 the \texttt{tsinfer} inference algorithm.
-Fig~\ref{fig:Supplementary_Figure_2} shows the genomic spans of
+%Fig~\ref{fig:Supplementary_Figure_2} shows the genomic spans of
+Fig.~S2 shows the genomic spans of
 ancestral nodes in an ARG inferred with \texttt{tsinfer}, demonstrating a clear
@@
 exist~\citep[e.g.][]{huson2007dendroscope,vaughan2017icytree},
-the methods can typically only handle a few hundred nodes.
+the methods can typically only handle a few hundred nodes 
+(but see e.g., \citet{hadfield2018nextstrain})
 Visualisation of large-scale tree topologies with millions of nodes
@@
 to capture topological features at different scales,
-and is an active research area~\citep{hadfield2018nextstrain, wong2022dynamic,kramer2023treenome}.
+and is an active research area~\citep{wong2022dynamic,kramer2023treenome}.
 Adapting such methods, and integrating into \texttt{tsbrowse} to provide
@@
 % Supplementary data are available at \textcolor{red}{Bioinformatics online}.
-% \section*{Conflict of interest}
-% No competing interest is declared.
+\section*{Conflict of interest}
+No competing interest is declared.
 
@@
 \bibliography{reference}
-
-\clearpage \onecolumn \setcounter{figure}{0}
-% Reset the figure counter
-\renewcommand{\thefigure}{S\arabic{figure}}
-% Change figure numbering to S1, S2, etc.
-
-\section{Supplementary Methods}
-\subsection{Simulation of truth dataset (Fig. \ref{fig:Figure_2})}
-\label{subsec:sweep_simulation}
-Ancestral
-histories of 300 samples were simulated with \texttt{SweepGenicSelection}
-function in \texttt{msprime (version 1.3.3)}. A combination of models was used:
-in the recent past, a selective sweep was simulated with a beneficial allele
-situated in the middle of a 5 Mb sequence. The frequency of the allele in the
-population was set at 0.0001 at the beginning of the sweep. The allele fixed in
-the population at a frequency of 0.9999. The strength of selection was set
-using the selection coefficient, \textit{s = 0.25}. A time increment,
-\textit{dt = 1e-6} was used to step through the sweep. Mutations were added to
-the ARG at a rate of 1e-8 per base pair per generation. A recombination rate of 1e-8 per base pair per generation. For simulating history before occurrence of the
-sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was
-used until coalescence was achieved.
-
-\subsection{Inference of selective sweep dataset (Suppl. Fig.
-    \ref{fig:Supplementary_Figure_3})} The following software were used to infer
-ARGs from the truth dataset described above: \texttt{tsinfer version 0.3.3}
-\citep{kelleher2019inferring}, \texttt{tsdate version 0.2.1}, 
-\texttt{Relate version 1.2.2} \citep{speidel2019method}, 
-\texttt{ARG-needle version 1.0.3} \citep{zhang2023biobank},
-\texttt{SINGER version 0.1.8-beta} \citep{deng2024robust}.
-For all
-methods, the following parameters of inference were used: \textit{recombination
-    rate = 1e-8}, \textit{mutation rate = 1e-8}, \textit{effective population size
-    = 10,000}. Default values were used for all other inference parameters.
-
-\subsection{Inference of SARS-CoV-2 ARGs}
-The ARG shown in Fig \ref{fig:Supplementary_Figure_1} was inferred with
-\texttt{sc2ts}~\citep{zhang2023biobank} using the Viridian
-dataset~\citep{hunt2024addressing}. It consists of
-2,482,157 samples
-2,689,054 nodes,
-2,689,982 edges
-and
-1,923,169 mutations.
-% Trees   348
-% Sequence Length 29904.0
-% Time Units  days
-% Sample Nodes    2482157
-% Total Size  1.5 GiB
-% Table   Rows    Size    Has Metadata
-% Edges   2689982 82.1 MiB
-% Individuals 0   24 Bytes
-% Migrations  0   8 Bytes
-% Mutations   1923169 202.7 MiB
-% Nodes   2689054 1.1 GiB
-% Populations 0   8 Bytes
-% Provenances 1131    1.7 MiB
-% Sites   29803   3.1 MiB
-Running \texttt{tsbrowse preprocess} on the input \texttt{tszip} file (113M)
-required 2m19s time elapsed (15m15s CPU time) on an Intel Core(TM) i7-9700 CPU.
-The resulting \texttt{.tsbrowse} file size was 130M.
-% Preprocessing completed. You can now view with `python -m tsbrowse serve v1-beta1_2023-02-21.pp.md.il.ts.tsbrowse`
-
-% real    2m19.028s
-% user    15m15.049s
-% sys     0m8.159s
-% ```
-% Sizes:
-% ```
-% $ ls -lh v1-beta1_2023-02-21.pp.md.il.ts.ts*
-% -rw-rw-r-- 1 jk jk 130M Mar 27 15:08 v1-beta1_2023-02-21.pp.md.il.ts.tsbrowse
-% -rw-r--r-- 1 jk jk 113M Mar 27 15:04 v1-beta1_2023-02-21.pp.md.il.ts.tsz
-% ```
-
-\subsection{Inference of 1000 Genomes dataset (Suppl. Fig.
-\ref{fig:Supplementary_Figure_3}, Suppl. Fig.
-\ref{fig:Supplementary_Figure_4})} The 1000 Genomes dataset was downloaded from
-\texttt{\url{https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/working/20220422_3202_phased_SNV_INDEL_SV/}}.
-The ancestral fasta sequence for chromosome 17 (GRCh38) was downloaded
-from the Ensembl database. Inference was performed with a Snakemake pipeline
-(\texttt{\url{https://github.com/benjeffery/tsinfer-snakemake/}}) using
-\texttt{tsinfer version 0.3.3} for the long arm of chromosome 17 after
-filtering out duplicate variant positions, variants with missing or low quality
-ancestral allele, singletons, n-1-tons and n-2-tons. Only bi-allelic SNPs were
-included for inference. For Suppl. Fig. \ref{fig:Supplementary_Figure_4}, 
-\texttt{tsdate version 0.2.1} was used to estimate the age of ancestral nodes
- with \textit{mutation rate = 1.29e-8}, setting all other parameters to default 
- values.
-
-\subsection{} Code to recreate datasets and infer ARGs is available in an
-accompanying repository:
-\texttt{\url{https://github.com/savitakartik/tsbrowse-paper}}.
-
-\clearpage
-\section{Supplementary Figures}
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=0.95\linewidth]{figures/SuppFig1.png}
-    \caption{\textbf{\texttt{tsbrowse} applied to SARS-CoV-2 ARGs.}
-        A screenshot of \texttt{tsbrowse}'s depiction of
-        1,923,169 mutations in an
-        ARG inferred by \texttt{sc2ts}; see  text for details. Also shown are the
-        gene annotations along the X-axis.}
-    \label{fig:Supplementary_Figure_1}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.95\linewidth]{figures/SuppFig2.png}
-    \caption{\textbf{A view of the Nodes page for a 1000 Genomes inference.} A
-        screenshot of \texttt{tsbrowse}'s Nodes view for an inference of the long arm
-        of chromosome 17 from the 1000 Genomes whole-genome sequencing dataset. At the
-        top is a plot of node spans over time. The length of sequence that the nodes
-        span is shown on the X axis, and the time of nodes is shown on the Y axis. The
-        histogram at the bottom shows the distribution of node spans. }
-    \label{fig:Supplementary_Figure_2}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.95\linewidth]{figures/SuppFig3.png}
-    \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
-    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate,
-ARG-Needle, Relate and SINGER inferences of the truth dataset
-simulated under a selective sweep model (in Figure 2 of the main text). For
-SINGER, an examplar ARG from the set of output posterior ARG samples is shown.
-The X coordinate represents genomic position, each horizontal segment on the
-plot shows the genomic coordinates that the edge spans, and Y coordinate shows
-time of either the parent or child node in the edge.}
-    \label{fig:Supplementary_Figure_3}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.95\linewidth]{figures/SuppFig4.png}
-    \caption{\textbf{Identifying ARG inference problems with \texttt{tsbrowse}} 
-Screenshots of \texttt{tsbrowse}'s Mutations view (A) and Edges view (B) for a
- 600 kB region of chromosome 17 inferred from 3,202 participants from the 1000 
- Genomes Whole Genome Sequencing dataset (1000G) \citep{1000GWGS}. 
- The poor performance of \texttt{tsinfer} in this variant-poor region is evidenced
-  by the long edges spanning gaps in mutation density.}
-    \label{fig:Supplementary_Figure_4}
-\end{figure}
 \end{document}

### 2025-06-19 manuscript/supplement.tex

%\documentclass{article}
\documentclass[unnumsec,webpdf,contemporary,large,namedate]{oup-authoring-template}
\onecolumn
\usepackage{hyperref}
\usepackage{lmodern}
\usepackage{graphicx}
\usepackage{xr}

\begin{document}
\renewcommand{\figurename}{Fig.~S}
\section{Supplementary Methods}
%\subsection{Simulation of truth dataset (Fig. \ref{fig:Figure_2})}
\subsection{Simulation of truth dataset (Fig.~2)}
\label{subsec:sweep_simulation}
Ancestral
histories of 300 samples were simulated with \texttt{SweepGenicSelection}
function in \texttt{msprime (version 1.3.3)}. A combination of models was used:
in the recent past, a selective sweep was simulated with a beneficial allele
situated in the middle of a 5 Mb sequence. The frequency of the allele in the
population was set at 0.0001 at the beginning of the sweep. The allele fixed in
the population at a frequency of 0.9999. The strength of selection was set
using the selection coefficient, \textit{s = 0.25}. A time increment,
\textit{dt = 1e-6} was used to step through the sweep. Mutations were added to
the ARG at a rate of 1e-8 per base pair per generation. A recombination rate of 
1e-8 per base pair per generation. For simulating history before occurrence of the
sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was
used until coalescence was achieved.

\subsection{Inference of selective sweep dataset (Fig.~S\ref{fig:Supplementary_Figure_3})} 
The following software were used to infer ARGs from the truth dataset described above: 
\texttt{tsinfer version 0.3.3}
\citep{kelleher2019inferring}, \texttt{tsdate version 0.2.1}, 
\texttt{Relate version 1.2.2} \citep{speidel2019method}, 
\texttt{ARG-needle version 1.0.3} \citep{zhang2023biobank},
\texttt{SINGER version 0.1.8-beta} \citep{deng2024robust}.
For all methods, the following parameters of inference were used: \textit{recombination
    rate = 1e-8}, \textit{mutation rate = 1e-8}, \textit{effective population size
    = 10,000}. Default values were used for all other inference parameters.

\subsection{Inference of SARS-CoV-2 ARGs}
The ARG shown in Fig.~S\ref{fig:Supplementary_Figure_1} was inferred with
\texttt{sc2ts}~\citep{zhang2023biobank} using the Viridian
dataset~\citep{hunt2024addressing}. It consists of
2,482,157 samples
2,689,054 nodes,
2,689,982 edges
and
1,923,169 mutations.
% Trees   348
% Sequence Length 29904.0
% Time Units  days
% Sample Nodes    2482157
% Total Size  1.5 GiB
% Table   Rows    Size    Has Metadata
% Edges   2689982 82.1 MiB
% Individuals 0   24 Bytes
% Migrations  0   8 Bytes
% Mutations   1923169 202.7 MiB
% Nodes   2689054 1.1 GiB
% Populations 0   8 Bytes
% Provenances 1131    1.7 MiB
% Sites   29803   3.1 MiB
Running \texttt{tsbrowse preprocess} on the input \texttt{tszip} file (113M)
required 2m19s time elapsed (15m15s CPU time) on an Intel Core(TM) i7-9700 CPU.
The resulting \texttt{.tsbrowse} file size was 130M.
% Preprocessing completed. You can now view with `python -m tsbrowse serve v1-beta1_2023-02-21.pp.md.il.ts.tsbrowse`

% real    2m19.028s
% user    15m15.049s
% sys     0m8.159s
% ```
% Sizes:
% ```
% $ ls -lh v1-beta1_2023-02-21.pp.md.il.ts.ts*
% -rw-rw-r-- 1 jk jk 130M Mar 27 15:08 v1-beta1_2023-02-21.pp.md.il.ts.tsbrowse
% -rw-r--r-- 1 jk jk 113M Mar 27 15:04 v1-beta1_2023-02-21.pp.md.il.ts.tsz
% ```

\subsection{Inference of 1000 Genomes dataset (Fig.~S\ref{fig:Supplementary_Figure_2}, 
Fig.~S\ref{fig:Supplementary_Figure_4})} The 1000 Genomes dataset was downloaded from
\texttt{\url{https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/working/20220422_3202_phased_SNV_INDEL_SV/}}.
The ancestral fasta sequence for chromosome 17 (GRCh38) was downloaded
from the Ensembl database. Inference was performed with a Snakemake pipeline
(\texttt{\url{https://github.com/benjeffery/tsinfer-snakemake/}}) using
\texttt{tsinfer version 0.3.3} for the long arm of chromosome 17 after
filtering out duplicate variant positions, variants with missing or low quality
ancestral allele, singletons, n-1-tons and n-2-tons. Only bi-allelic SNPs were
included for inference. For Supplementary Figure \ref{fig:Supplementary_Figure_4}, 
\texttt{tsdate version 0.2.1} was used to estimate the age of ancestral nodes
 with \textit{mutation rate = 1.29e-8}, setting all other parameters to default 
 values.

\subsection{Code availability} Code to recreate datasets and infer ARGs is available in an
accompanying repository:
\texttt{\url{https://github.com/savitakartik/tsbrowse-paper}}.

\clearpage
\section{Supplementary Figures}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/SuppFig1.png}
    \caption{\textbf{\texttt{tsbrowse} applied to SARS-CoV-2 ARGs.}
        A screenshot of \texttt{tsbrowse}'s depiction of
        1,923,169 mutations in an
        ARG inferred by \texttt{sc2ts}; see text for details. Also shown are the
        gene annotations along the X-axis.}
    \label{fig:Supplementary_Figure_1}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/SuppFig2.png}
    \caption{\textbf{A view of the Nodes page for a 1000 Genomes inference.} A
        screenshot of \texttt{tsbrowse}'s Nodes view for an inference of the long arm
        of chromosome 17 from the 1000 Genomes whole-genome sequencing dataset. At the
        top is a plot of node spans over time. The length of sequence that the nodes
        span is shown on the X axis, and the time of nodes is shown on the Y axis. The
        histogram at the bottom shows the distribution of node spans. }
    \label{fig:Supplementary_Figure_2}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/SuppFig3.png}
    \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate,
ARG-Needle, Relate and SINGER inferences of the truth dataset
simulated under a selective sweep model (in Figure 2 of the main text). For
SINGER, an examplar ARG from the set of output posterior ARG samples is shown.
The X coordinate represents genomic position, each horizontal segment on the
plot shows the genomic coordinates that the edge spans, and Y coordinate shows
time of either the parent or child node in the edge.}
    \label{fig:Supplementary_Figure_3}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/SuppFig4.png}
    \caption{\textbf{Identifying ARG inference problems with \texttt{tsbrowse}} 
Screenshots of \texttt{tsbrowse}'s Mutations view (A) and Edges view (B) for a
 600 kB region of chromosome 17 inferred from 3,202 participants from the 1000 
 Genomes Whole Genome Sequencing dataset (1000G) \citep{1000GWGS}. 
 The poor performance of \texttt{tsinfer} in this variant-poor region is evidenced
  by the long edges spanning gaps in mutation density.}
    \label{fig:Supplementary_Figure_4}
\end{figure}
\clearpage
\bibliographystyle{natbib}
\bibliography{reference}
\end{document}

### 2025-06-19 manuscript/revision/response_to_reviewers.tex
(hopefully) final readthrough before submitting revision

@@
 and is an active research area". Thanks for bringing Nextstrain to our 
-attention, and we have now Hadfield et al. as a citation for visualisation of medium-sized toplogies.}\\
+attention, and we have now included \href{{Wong et al 2024}}{Hadfield et al.} as a citation for visualisation of 
+medium-sized toplogies.}\\
 
@@
 is a part of tskit or of twbrowse.\\
-\textcolor{blue}{To clarify, the tszip is part of the tskit ecosystem. It is not part of the tsbrowse 
+\textcolor{blue}{To clarify, tszip is part of the tskit ecosystem. It is not part of the tsbrowse 
 package but can be used to compress the tables output from the tsbrowse preprocess step. Thank you 
@@
 hinting that it is important.\\
-\textcolor{blue}{We agree with the reviewer's opinion, and believe that this figure is important 
+\textcolor{blue}{We agree with your opinion, and believe that this figure is important 
 to demonstrate the application of tsbrowse to large-scale datasets (which is a key design 
-feature of the app). However, we are unable to move this figure to the main matter considering journal 
-guidelines for page limit.}\\
+feature of the app). However, given the journal guidelines for page limit, we are unable 
+to move this figure to the main matter.}\\
 
@@
 solution (e.g. docker).\\
-\textcolor{blue}{We have now created a Docker container in order to ensure a smooth experience across 
+\textcolor{blue}{We have now created a Docker container in order to ensure a smooth user experience across 
 computing environments. Usage instructions have been added to the app documentation.}\\
@@
 This paper offers a browser-based visualisation tool for ancestral recombination graphs (ARGs), which is 
-a data structure representing gene genealogies for a sample of DNA sequences. The complication in the p
-resence of recombination is that these structures are not simple trees, but - because of the possibility 
+a data structure representing gene genealogies for a sample of DNA sequences. The complication in the 
+presence of recombination is that these structures are not simple trees, but - because of the possibility 
 of recombination - graphs with reticulated nodes, which makes exploratory data analysis difficult.
@@
 \textcolor{blue}{We completely agree that this figure would have been a compelling main display item had the journal 
-page/figure limits allowed for it. Considering the necessity of the existing main display items in establishing 
-the fundamental aspects of tsbrowse, we have decided to retain the figure in the supplementary.}\\
+page/figure limits allowed for it. Considering the importance of the existing main display items in illustrating 
+the core aspects of tsbrowse, we have decided to retain this figure in the supplementary.}\\
 
@@
 \textcolor{blue}{This is an important observation, and we agree that drawing (parts of) local trees would be 
-a great feature. For smaller datasets, tools such as \texttt{tskit-arg-visualizer} 
-(\url{https://pypi.org/project/tskit-arg-visualizer/}) and existing tskit infrastructure, for example 
-functions \texttt{draw\_text()} and \texttt{draw\_svg()} already provide this functionality. A key 
+a great feature. For smaller datasets, tools such as 
+\href{https://pypi.org/project/tskit-arg-visualizer/}{tskit-arg-visualizer} and existing tskit infrastructure, 
+for example functions \texttt{draw\_text()} and \texttt{draw\_svg()} already provide this functionality. A key 
 motivation for this work is to guide large-scale inference, and implementing tree visualisation that is 
@@
 We therefore think that implementing this feature is not feasible for this work and is better scoped as future 
-work or as a separate endeavour. In the text, we do state that tsbrowse is a visualiser for the information 
-structure in ARGs, which we hope implies that it is a visualiser for the properties of the ARG, rather than 
-the ARG topology itself. We have also already explicitly stated in the Discussion that drawing local trees is 
-an important future direction.}\\
+work or as a separate endeavour. In line with this, we have already explicitly stated in the Discussion that drawing 
+local trees is an important future direction. In the text, we do state that tsbrowse "provides visualisations of 
+the information structure of ARGs", which we hope implies that it is a visualiser for the \textit{properties} 
+of the ARG, rather than its topology itself.}\\
 
@@
 statement. The history of ARG inference is a complicated and confusing subject, dealt with exhaustively by 
-Wong et al 2024, and would be impossible to accurately summarise in the available space. We note, though, 
+\href{https://academic.oup.com/genetics/article/228/1/iyae100/7714980}{Wong et al. 2024}, and would be impossible to accurately summarise in the available space. We note, though, 
 that two of the recent methods cited here (ARGweaver and SINGER) sample ARGs from the posterior defined by a 
-coalescent model in a statistically rigorous manner}\\
+coalescent model in a statistically rigorous manner.}\\
 
@@
 Fig S2 caption - 'Consortium (2015)' is citing incorrectly.\\
-\textcolor{blue}{Fixed (removed this citation as it was not required)}\\
+\textcolor{blue}{Fixed (removed this citation as it was not required).}\\
 
@@
 to install tsbrowse as a new user are clearly stated at the beginning of the documentation,
-and separate from developer instructions.}\\
+and are separate from developer instructions.}\\
 
@@
 colormaps.'), )>)\\
-\textcolor{blue}{We are grateful to the reviewer for bringing this bug to our attention. 
-We have solved the issue and released an updated version of the package.}\\
+\textcolor{blue}{Thank you for bringing this bug to our attention. We have solved the issue and released an 
+updated version of the package.}\\
 

### 2025-06-19 manuscript/paper.tex
(hopefully) final readthrough before submitting revision

@@
     Ancestral Recombination Graphs (ARGs) represent the interwoven paths of genetic
-    ancestry for a set of recombining sequences. The ability to capture the
+    ancestry of a set of recombining sequences. The ability to capture the
     evolutionary history of samples makes ARGs valuable in a wide range of
@@
 precomputing all the information required for visualisation,
-and storing as a compressed \texttt{.tsbrowse} file.
+and storing it as a compressed \texttt{.tsbrowse} file.
 %(The \texttt{.tsbrowse} file is also a valid input for
@@
     \centering \includegraphics[width=0.95\linewidth]{figures/MainFig2.png}
-    \caption{Screenshot of the Edges tab in \texttt{tsbrowse}.
+    \caption{Screenshot of the Edges view in \texttt{tsbrowse}.
         Visualisation of the edges in a simulated ARG with a strong selective
@@
 % genes or other sequence features. 
-The user interface is presented as a dashboard, with pages to
+The user interface is presented as a dashboard, with views to
 describe various aspects of the ARG.
 Fig.~\ref{fig:Figure_2}
-demonstrates the user interface using the Edges page as an example.
+demonstrates the user interface using the Edges view as an example.
 The plot on the left is an overview of the 33,929 edges
@@
 Supplement for details).
-Each edge is depicted as a horizontal line connecting the genomic coordinates
+Each edge in the ARG is depicted as a horizontal line connecting the genomic coordinates
 of the
@@
 using a set of controls on the top-right corner provided by 
-Bokeh~\citep{bokeh},allowing them to pan and zoom as required. 
+Bokeh~\citep{bokeh}, allowing them to pan and zoom as required. 
 The histograms to the
@@
 is given in the simulated ARG of Fig.~\ref{fig:Figure_2}, where
-the gap in the edges view corresponds to the characteristic dip in diversity
+the gap in the Edges view corresponds to the characteristic dip in diversity
 of a selective sweep. Comparing this ground truth to the ARGs inferred
@@
 results. These differences in ancestral haplotypes illustrated by
-the edges view are unlikely to be captured by the tree-by-tree distance
+the Edges view are unlikely to be captured by the tree-by-tree distance
 metrics usually used to evaluate
@@
 element of the inference pipeline. Fig.~S4
-shows a region of the 1000G data with gaps in site density that are spanned by
+shows a region of the 1000 Genomes data with gaps in site density that are spanned by
  exceptionally long edges, which are likely to bias downstream statistics.
@@
 the methods can typically only handle a few hundred nodes 
-(but see e.g., \citet{hadfield2018nextstrain})
+(but see e.g., \citet{hadfield2018nextstrain}).
 Visualisation of large-scale tree topologies with millions of nodes
@@
 and is an active research area~\citep{wong2022dynamic,kramer2023treenome}.
-Adapting such methods, and integrating into \texttt{tsbrowse} to provide
+Adapting such methods, and integrating them into \texttt{tsbrowse} to provide
 a local tree viewer that operates at the million-node scale

### 2025-06-19 manuscript/revision/response_to_editor.tex
(hopefully) final readthrough before submitting revision

@@
 input provided. Attached are the revised manuscript (\textit{paper.pdf}), 
-a version of the manuscript showing the differences to the submitted version ("diff.pdf"), 
-and a response to the reviewers' comments (\textit{response\_to\_reviwers.pdf})
+a version of the manuscript showing the differences to the submitted version 
+(\textit{diff.pdf}), and a response to the reviewers' comments 
+(\textit{response\_to\_reviwers.pdf})
 
@@
     under Availability, in addition to code repository
-    \item Manuscript adheres to the Overleaf template, \textit{OUP General Template}
-    \item Supplementary material attached as separate pdf, \textit{supplement.pdf}
+    \item Manuscript adheres to the Overleaf template: \textit{OUP General Template}
+    \item Supplementary material attached as a separate pdf (\textit{supplement.pdf})
     \item All relevant LaTeX files included

### 2025-06-19 manuscript/supplement.tex
(hopefully) final readthrough before submitting revision

@@
 \renewcommand{\figurename}{Fig.~S}
+\begin{center}
+    {\large Tsbrowse: an interactive browser for Ancestral Recombination Graphs}\\[1ex]
+    {\large Supplementary Information}
+\end{center}
 \section{Supplementary Methods}
@@
 Ancestral
-histories of 300 samples were simulated with \texttt{SweepGenicSelection}
+histories of 300 samples were simulated with the \texttt{SweepGenicSelection}
 function in \texttt{msprime (version 1.3.3)}. A combination of models was used:
@@
 the ARG at a rate of 1e-8 per base pair per generation. A recombination rate of 
-1e-8 per base pair per generation. For simulating history before occurrence of the
-sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was
+1e-8 per base pair per generation was used. For simulating history before the occurrence 
+of the sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was
 used until coalescence was achieved.
-
-\subsection{Inference of selective sweep dataset (Fig.~S\ref{fig:Supplementary_Figure_3})} 
-The following software were used to infer ARGs from the truth dataset described above: 
-\texttt{tsinfer version 0.3.3}
-\citep{kelleher2019inferring}, \texttt{tsdate version 0.2.1}, 
-\texttt{Relate version 1.2.2} \citep{speidel2019method}, 
-\texttt{ARG-needle version 1.0.3} \citep{zhang2023biobank},
-\texttt{SINGER version 0.1.8-beta} \citep{deng2024robust}.
-For all methods, the following parameters of inference were used: \textit{recombination
-    rate = 1e-8}, \textit{mutation rate = 1e-8}, \textit{effective population size
-    = 10,000}. Default values were used for all other inference parameters.
 
@@
 dataset~\citep{hunt2024addressing}. It consists of
-2,482,157 samples
+2,482,157 samples, 
 2,689,054 nodes,
@@
 Running \texttt{tsbrowse preprocess} on the input \texttt{tszip} file (113M)
-required 2m19s time elapsed (15m15s CPU time) on an Intel Core(TM) i7-9700 CPU.
+required 2m19s of elapsed time (15m15s CPU time) on an Intel Core(TM) i7-9700 CPU.
 The resulting \texttt{.tsbrowse} file size was 130M.
@@
 
-\subsection{Code availability} Code to recreate datasets and infer ARGs is available in an
-accompanying repository:
+\subsection{Inference of selective sweep dataset (Fig.~S\ref{fig:Supplementary_Figure_3})} 
+The following software was used to infer ARGs from the truth dataset described in the
+section 'Simulation of truth dataset' above: 
+\texttt{tsinfer version 0.3.3}
+\citep{kelleher2019inferring}, \texttt{tsdate version 0.2.1}, 
+\texttt{Relate version 1.2.2} \citep{speidel2019method}, 
+\texttt{ARG-needle version 1.0.3} \citep{zhang2023biobank},
+\texttt{SINGER version 0.1.8-beta} \citep{deng2024robust}.
+For all inferences, the following parameters were used: \textit{recombination
+    rate = 1e-8}, \textit{mutation rate = 1e-8}, \textit{effective population size
+    = 10,000}. Default values were used for other parameters.
+
+\subsection{Code availability} Code to recreate datasets used in this paper are available at:
 \texttt{\url{https://github.com/savitakartik/tsbrowse-paper}}.
@@
     \includegraphics[width=0.95\linewidth]{figures/SuppFig2.png}
-    \caption{\textbf{A view of the Nodes page for a 1000 Genomes inference.} A
+    \caption{\textbf{Nodes view for a 1000 Genomes inference.} A
         screenshot of \texttt{tsbrowse}'s Nodes view for an inference of the long arm
@@
     \includegraphics[width=0.95\linewidth]{figures/SuppFig3.png}
-    \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
+    \caption{\textbf{\texttt{Tsbrowse} applied to inference methods.}
     A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate,
 ARG-Needle, Relate and SINGER inferences of the truth dataset
-simulated under a selective sweep model (in Figure 2 of the main text). For
-SINGER, an examplar ARG from the set of output posterior ARG samples is shown.
+simulated under a selective sweep model (shown in Figure 2 of the main text). For
+SINGER, one of the posterior ARG samples is shown.
 The X coordinate represents genomic position, each horizontal segment on the
@@
     \includegraphics[width=0.95\linewidth]{figures/SuppFig4.png}
-    \caption{\textbf{Identifying ARG inference problems with \texttt{tsbrowse}} 
+    \caption{\textbf{Identifying ARG inference problems with \texttt{tsbrowse}.} 
 Screenshots of \texttt{tsbrowse}'s Mutations view (A) and Edges view (B) for a
- 600 kB region of chromosome 17 inferred from 3,202 participants from the 1000 
- Genomes Whole Genome Sequencing dataset (1000G) \citep{1000GWGS}. 
+ 600 kb region of chromosome 17 inferred from 3,202 participants from the 1000 
+ Genomes Whole Genome Sequencing dataset \citep{1000GWGS}. 
  The poor performance of \texttt{tsinfer} in this variant-poor region is evidenced

### 2025-06-19 manuscript/revision/response_to_editor.tex
change to filename

@@
 a version of the manuscript showing the differences to the submitted version 
-(\textit{diff.pdf}), and a response to the reviewers' comments 
+(\textit{main\_diff.pdf, supplement\_diff.pdf}), and a response to the reviewers' comments 
 (\textit{response\_to\_reviwers.pdf})

### 2025-06-20 manuscript/revision/response_to_reviewers.tex
jk review response

@@
 and is an active research area". Thanks for bringing Nextstrain to our 
-attention, and we have now included \href{{Wong et al 2024}}{Hadfield et al.} as a citation for visualisation of 
-medium-sized toplogies.}\\
+attention, and we have now included \href{{https://academic.oup.com/bioinformatics/article/34/23/4121/5001388}}{Hadfield et al.} 
+as a citation for visualisation of medium-sized toplogies.}\\
 

### 2025-06-20 manuscript/supplement.tex
jk review response

@@
 The following software was used to infer ARGs from the truth dataset described in the
-section 'Simulation of truth dataset' above: 
+section ``Simulation of truth dataset'' above: 
 \texttt{tsinfer version 0.3.3}

### 2025-07-04 manuscript/paper.tex

@@
 \journaltitle{Bioinformatics}
-\journaltitle{Preprint}
+%\journaltitle{Preprint}
 \DOI{DOI HERE}
@@
 
+%Summary, Availability and Implementation, Contact, and Supplementary information
 \abstract{
-    %\textbf{Motivation:}
+    \textbf{Summary:}
     Ancestral Recombination Graphs (ARGs) represent the interwoven paths of genetic
@@
     of visualisation tools, which are crucial for validating inferences and
-    generating hypotheses.
-    %\textbf{Results:} 
-    We present \texttt{tsbrowse}, an open-source Python web-app for the interactive
-    visualisation of the fundamental building-blocks of ARGs, i.e.,  nodes, edges
-    and mutations. We demonstrate the application of \texttt{tsbrowse} to various
-    data sources and scenarios, and highlight its key features of browsability
-    along the genome, user interactivity, and scalability to very large sample
-    sizes.\\ \textbf{Availability:}\\ Python package:
-    \texttt{\url{https://pypi.org/project/tsbrowse/}}, \\ Development version:
-    \texttt{\url{https://github.com/tskit-dev/tsbrowse}}, \\ Documentation:
-    \texttt{\url{https://tskit.dev/tsbrowse/docs/}}, \\ DOI: 
+    generating hypotheses. We present \texttt{tsbrowse}, an open-source,
+    web-based Python application for the interactive visualisation of the fundamental 
+    building blocks of ARGs, i.e., nodes, edges and mutations. We demonstrate the 
+    application of \texttt{tsbrowse} to various data sources and scenarios, and highlight 
+    its key features of browsability along the genome, user interactivity, and scalability 
+    to very large sample sizes.\\\textbf{Availability and Implementation:}
+    \texttt{Tsbrowse} is installed as a Python package from PyPI
+    (\texttt{\url{https://pypi.org/project/tsbrowse/}}), while a development version 
+    is maintained at \texttt{\url{https://github.com/tskit-dev/tsbrowse}}. Documentation 
+    is available at \texttt{\url{https://tskit.dev/tsbrowse/docs/}}.
+    Source code is archived on Zenodo with DOI, 
     \texttt{\url{https://doi.org/10.5281/zenodo.15683039}}\\
-    %\textbf{Contact:} \href{name@email.com}{name@email.com}\\
-    %\textbf{Supplementary information:} Available at
-    %\textit{\textcolor{red}{\href{https://www.google.com}{Bioinformatics}
-    %online}.}}
+    \textbf{Contact:} \href{savita.karthikeyan@st-hughs.ox.ac.uk}{savita.karthikeyan@st-hughs.ox.ac.uk}\\
+    \textbf{Supplementary information:} Available at
+    \textit{Bioinformatics} online.
 }
@@
 \subsection{Architecture}
-\texttt{Tsbrowse} is a modular client-server application
+\texttt{Tsbrowse} is a modular application
 written in Python, optimised for ARGs with millions
@@
 the data density is far too high to simply plot each mutation as a point.
-We overcome this problem by using Datashader and the wider
+We overcome this problem by using Datashader~\citep{Datashader} and the wider
 Holoviz ecosystem~\citep{Holoviz}, which efficiently rasterises large datasets
@@
 using a set of controls on the top-right corner provided by 
-Bokeh~\citep{bokeh}, allowing them to pan and zoom as required. 
+%Bokeh~\citep{bokeh}, allowing them to pan and zoom as required. 
+Bokeh (http://www.bokeh.pydata.org), allowing them to pan and zoom as required.
 The histograms to the
@@
 
-Interest in ARG-based methods is burgeoning, but the methods are new
+Interest in ARG-based methods is burgeoning, but the methods are new,
 and practical guidance on applying inferences to real data is lacking.
@@
 \section*{Funding}
-SK acknowledges full support and funding from Novo Nordisk Research Centre Oxford,
-and funding from the Biotechnology and Biological Sciences Research Council 
+SK acknowledges full support and funding from Novo Nordisk Research Centre Oxford
+and the Biotechnology and Biological Sciences Research Council 
 (UKRI-BBSRC) [grant number BB/T008784/1]. DMR is funded by 
